---
title: "Size frequency (CPUE) distributions for SAFE report"
author: Johanna Wren
date: April 4, 2018
output:
  html_notebook:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
      tod_depth: 3
---

<script>
$(document).ready(function() {
  $items = $('div#TOC li');
  $items.each(function(idx) {
    num_ul = $(this).parentsUntil('#TOC').length;
    $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
  });

});
</script>

# Description

*UPDATE: in April 2023 we separated the data download from the data analysis. This script is data analysis and plot making only. The data used here is downloaded from NOAA PIFSC fisheries servers using the SAFE_LogDeal_DataDownload.Rmd script.*

This script is a rework of RI_SizeFreqHist.Rmd from the bigeye recruitment index paper. This script has been updated to download data from oracle directly

This script can be used to update the size frequency each quarter so that we can monitor eventual new recruitment pulses and keep the council etc. updated. 

# Prepare workspace
```{r message=FALSE}
# Clear workspace
rm(list=ls())

# Set working directory
mainDir <- '~/Documents/SAFEindicators/RAnalysis_SAFE/'

# Load libraries
library(tidyverse)
library(lubridate)
library(rje)
library(zoo)
```

# Import data

Define what year you are working on here and you don't have to change it anywhere else in the script
```{r}
# The last full year included in the analysis
yr = 2023
```

Here we import data saved locally and filter out only bigeye tuna (coded as species 6 in the dealer dataset) for deep-set trips onlu and that we don't include trips in the current year. 
```{r}
# Import the data from saved files
setwd(file.path(mainDir, 'Data'))
dealerRaw <- readRDS(paste0('SAFE_dealer_', yr, '.rds'))
logbookRaw <- readRDS(paste0('SAFE_logbook_', yr, '.rds'))

# Filter data to bigeye only for deep set trips before 2023
dealerRaw <- dealerRaw %>%
  filter(REPORT_DATE < as.Date(paste0(yr+1, '-01-01')),
         SPECIES_FK == 6, 
         SET_TYPE == 'D')

# Make sure we only have information from the Hawaii fleet for trips spanning 2000- 2023
logbookRaw <- logbookRaw %>%
  filter(LANDYR >= 2000,  LANDYR <= yr, FLEET == 'HI', )
```

# Analyze data

## Convert weights from gilld & gutted to whole fish weights

Dealer weights are reported in lbs for gilled and gutted fresh, chilled fish. Here we convert their weight to whole fish weight using a non-linear conversion from *Langley et al. 2006* table 8. 

NOTE: Dealer data comes with a conversion factor (I kept it in the raw dataset) but it's fixed and for multiple entries it's quite complicated. I opted to use the Langley conversion here rather than the standard conversion factor reported because I felt it was most consistent and a better fit. 

NOTE 2: in 2024 Removing any records of negative NUM_SOLD. Total of 28 records, all in 2023.
```{r}
dealer <- dealerRaw %>%
  filter(NUM_SOLD >= 0) %>% 
  mutate(Weight_GilGut=MEAN_SOLD_WEIGHT/2.204624, 
         Weight_Whole=1.274959*round(Weight_GilGut)^0.960613, 
         Year=year(REPORT_DATE), 
         Quarter=quarter(REPORT_DATE)) %>%
  uncount(NUM_SOLD) %>%   # unbin the binned records
  mutate(NUM_SOLD=1) %>%  # make a num sold column since it was removed during the uncount operation
  select(TRIP_NUM, REPORT_DATE, NUM_SOLD, Weight_Whole, Year, Quarter) %>%
  mutate(Weight_Whole=ifelse(Weight_Whole > 100, 100, Weight_Whole))

```

## Calculate Age
This is only used to mark the dashed age lines in the figure

Here we calculate the age of fish from the weight. First we use a Richards growth parameterization from an intergrated analysis in Aires-da-Silva et al. 2015 getting an length from age information:

$$L_a=L_\infty(1+\frac{1}{p}e^{-K(a-t_0)})^{-p}$$
where $L_\infty$ is the asymptotic length, $K$ is the growth coefficient, $t_0$ is the inflection point on the curve, and $p$ is the shape parameter related to the ratio $L_a/L_\infty$

Then we convert the length given from the growth equation above to a weight using data from Phoebe that she derived based on the Hawaii deep-set fishery. 
```{r}
# using Aires-da-Silva 2015 Richard growth model using integrated data
a <- 1:7   # age in years
Linf <- 200.8               # asymptotic length
K <- 0.44                   # growth coefficient
t0 <- 1.26                  # inflexion point on the curve
p <- -4.27                  # shape parameter related to the ratio La/Linf
La <- Linf*(1+(1/p)*exp(-K*(a-t0)))^(-p)

# Convert from Length to weight using Phoebe's numbers for the Hawaii fishery
Weight <- 0.0187*La^2.9605  # these numbers are from phoebe
WeightAge <- round(Weight)/1000
```

## Calculate effort
Calculating both climatological quarterly efforts and annual quarter efforts.
```{r}
# Make a trip level dataset
logbook <- logbookRaw %>% 
  mutate(Quarter=quarter(LANDMON), Year=LANDYR) %>%
  filter(!RSCH_EXPMTL_CODE %in% c('R', 'X'), FLEET == 'HI', SET_TYPE == 'D')
  
# Calculate quarterly climatology effort
efrtTot <-  logbook %>%
  filter(Year <= 2009) %>%
  group_by(Quarter) %>%
  summarise(EffortClim=sum(HOOKSSET/1000, na.rm=T))
 
# Calculate quarterly annual efforts 
efrtYr <- logbook %>%
  group_by(Year, Quarter) %>%
  summarise(Effort=sum(HOOKSSET/1000, na.rm=T))
```

## Histograms
Calculating quarterly climatological histograms for use in the figures. Calculate these outside the loop since they are the same for each quarter. 
```{r}
# Calculate quarterly climatology histograms
cCPUE <- list()
for (q in 1:4) {
  cCPUE[[q]] <- dealer %>% 
    filter(Quarter==q, Year <= 2009) %>% 
    select(Weight_Whole) %>% 
    lapply(hist, plot=F)
  cCPUE[[q]]$Weight_Whole$CPUE <- cCPUE[[q]]$Weight_Whole$counts/efrtTot$EffortClim[q]
}
```

# Plot
This script plots two different figures. The first one is the size-CPUE distribution by quarter for the past two years along with the climatology. the second figure is the CPUE timeseries for all and small bigeye along with weight per unit effort. 

## Size-CPUE distribution
### Define plotting variables
```{r}
# Define the center bin locations for plotting
cXpos <- cCPUE[[1]]$Weight_Whole$mids
# Write out quarter names for plotting
qName <- c('First', 'Second', 'Third', 'Fourth')
# Set color palette
pal <- rev(cubeHelix(8, start=0, r=-0.1, gamma=1, hue=0.8))
```

Define the year range you want to plot
```{r}
# Set years you want to plot
yrRange <- (yr-2):yr
# Isolate the number of panels you want to plot so we get the quarters right
efrtPlot <- efrtYr %>% 
  filter(Year %in% yrRange)
```

### Plot the data
Here we plot and save the data in the output folder. 
```{r}
# Load Fisheries Data
setwd(file.path(mainDir, 'Output'))
# Start plotting
# Save file
png(paste0('SAFE_SizeFreqHist_', yr, '.png'), height=8, width=5.5, unit='in', res=300)
# Set plot dimensions
nPanels <- nrow(efrtPlot)
m <- matrix(c(rep(nPanels+1,nPanels+2),nPanels+2,1:nPanels,nPanels+3),nrow=nPanels+2, ncol=2)
layout(m, height=c(2,rep(3,nPanels),2), widths = c(0.5,10))
# loop for year and quarter
for (y in yrRange) {
  qIdx <- efrtPlot %>% filter(Year == y)
  for (q in qIdx$Quarter) {
    qEffort <- efrtYr %>% filter(Quarter==q, Year==y)
    qWeight <- dealer %>% filter(Quarter==q, Year==y) %>% select(Weight_Whole) %>% lapply(hist, plot=F)
    qCPUE <- qWeight$Weight_Whole$counts/qEffort$Effort
    qCPUE[is.infinite(qCPUE)] <- rep(0, 20)
    
    par(mar=c(0.25,3,0,1))
    plot(cCPUE[[q]]$Weight_Whole$mids, cCPUE[[q]]$Weight_Whole$CPUE, type='l', lwd=2, col='gray50', xlab=NA, ylab=NA, axes=F, xaxs='i', yaxs='i',
         xlim=c(0,100), ylim=c(0,1.1))
    par(new=T)
    plot(qWeight$Weight_Whole$mids, qCPUE, type='l', lwd=3, col=pal[q+2], xlab=NA, ylab=NA, axes=F, xaxs='i', yaxs='i',
         xlim=c(0,100), ylim=c(0,1.1))
    axis(2, las=1, at=c(0,0.5,1,1.1), labels=F)
    abline(v=WeightAge, lty=2, col='gray50')
    legend('topright', legend=c(paste(qName[q], 'Quarter', y, sep=' '), paste('Climatology', min(efrtYr$Year), '-', 2009, sep=' ')), col=c(pal[5], 'gray50'), lwd=2, bg='white', box.col='white')
    box()
    if (y == min(yrRange) & q == 1) {
      axis(2, las=1, at=c(0,0.5,1,1.1))
      axis(side=3, at = WeightAge, labels = c(1:7))
      mtext(side=3, line=2, text='Age (year)', cex=0.8)
    }
  }
}
# Add axis labels
axis(1)
mtext(side=1, line=2.25, 'Whole Fish Weight (kg)', cex=0.8)
axis(2, las=1, at=c(0,0.5,1,1.1))
par(mar=c(1,0,1,0))
plot(x=0.5,y=8,type='n', ylim=c(0,16), xlim=c(0,1), axes=F, xlab=NA, ylab=NA)
text(x=1,y=8, 'CPUE (no. fish/1000 hooks)', srt=90, cex=1.2, xpd=NA)
dev.off()
```
## Recruitment index
Make the running mean recruitment index plot

### Calculate plot variables
Summarize the logbook and dealer data on trip level and calculate number of small bigeye per trip
```{r}
logbookTrip <- logbook %>% 
  mutate(TRIP_NUM=as.numeric(paste(LANDYR, sprintf("%04d", TRIPNUM),sep=''))) %>%
  group_by(TRIP_NUM) %>%
  summarize(HOOKSSET=sum(HOOKSSET), LANDYR=unique(LANDYR), LANDMON=max(unique(LANDMON)))

# This splits up large and small bigeye and groups by trip number
dealer$Num_Small <- ifelse(dealer$Weight_Whole <= 15, dealer$NUM_SOLD, 0)

dealer2 <- dealer %>% 
  group_by(TRIP_NUM) %>% 
  summarise_at(.vars=c('NUM_SOLD', 'Num_Small', 'Weight_Whole'), .funs=sum)
```

### Merge and save dealer and logbook datasets
Now we merge the dealer and logbook (logData) datasets since we only really need the effort information from the logbook.

I did an `inner_join` here because I wanted to avoid having weight for sold fish (from dealer data) but no effort information for them (logbook data) or vice versa. 
```{r}
# merge dealer and logbook since we only really need the hooksset info
dealerFinal <- full_join(dealer2, logbookTrip, by='TRIP_NUM')

# Check output
dealerFinal
```
### Summarize by year
Summarizing the joined dealer-logbook data by year and calculating running mean. Here we use a three year running mean that is centered on the middle year. 
```{r}
# Raw data
dealerYear <- dealerFinal %>%
  filter(LANDYR >= 2000, LANDYR <= yr) %>%
  group_by(LANDYR) %>%
  summarize_at(c('Num_Small', 'NUM_SOLD', 'Weight_Whole', 'HOOKSSET'), sum, na.rm=T) %>%
  mutate(CPUEsmall=Num_Small/(HOOKSSET/1000), CPUEall= NUM_SOLD/(HOOKSSET/1000), WPUE=Weight_Whole/(HOOKSSET/1000))
```

### Plot the data
Make plot that is color coded and axes are colored as well
```{r}
# Change working directory
setwd(file.path(mainDir, 'Output'))
# Calculate plot range
idx <- range(dealerYear$CPUEall, na.rm=T)
mn <- signif(idx[1], digits=2)-0.1
mx <- signif(idx[2], digits=2)
stp <- round((mx-mn)/6,digits=1)

# Make plot
png(paste0('RecIndexRaw_', yr, '.png'), height=6, width=11, res=300, unit='in')
# Make raw plot
par(mar=c(5,9,1,4.5))
with(dealerYear, plot(LANDYR, CPUEsmall, type='l', lwd=3, lty=1, col='gray40', ylab=NA, xlab='Year', axes=F, xaxs='i'))
axis(1, at=seq(2000,yr,1))
axis(2, las=1)
mtext(side=2,line=3, expression("Bigeye Tuna "<=15 ~"kg CPUE (# fish/1000 hks)"))
box()
par(new=T)
with(dealerYear, plot(LANDYR, CPUEall, type='l', lwd=3, lty=2, col='#A3218E', ylab=NA, xlab=NA, ylim=c(mn,mx), axes=F, xaxs='i'))
axis(2, line=5, at=seq(mn,mx,stp), las=1, col='#A3218E', col.axis='#A3218E')
mtext(side=2, line=7.5, col='#A3218E', 'Bigeye Tuna CPUE (# fish /1000 hks)')
par(new=T)
with(dealerYear, plot(LANDYR, WPUE, type='l', lwd=3, lty=3, col='#00ACAC', ylab=NA, xlab=NA, axes=F, xaxs='i'))
axis(4, las=1, col='#00ACAC', col.axis='#00ACAC')
mtext(side=4, line=3, col='#00ACAC', 'Bigeye  biomass CPUE (kg/1000 hks)')
legend(2003.5, 172, legend=c(expression("CPUE Bigeye Tuna "<=15 ~"kg"), 'CPUE All Bigeye Tuna', 'Biomass CPUE All Bigeye Tuna'), lty=1:3, col=c('gray30', '#A3218E' , '#00ACAC'), bty='n', lwd=3)

dev.off()

```
