---
title: "Atmospheric_Carbon_Dioxide"
author: "phoebe.woodworth-jefcoats@noaa.gov"
format:
  docx:
    reference-doc: SAFE-Reference-Doc.docx
---

## Atmospheric Concentration of Carbon Dioxide at Mauna Loa

```{r}
#| include: false
### Load libraries
# Commenting out the plotly, reticulate, python stuff because it is 
# inexplicably not working this year
library(tidyverse, quietly = TRUE)
library(lubridate)
library(here)
# remotes::install_github("nmfs-fish-tools/nmfspalette")
library(nmfspalette)
# library(plotly)
# library(reticulate)

# reticulate::install_miniconda()
# reticulate::py_install("packaging")
# reticulate::use_miniconda('r-reticulate')
# reticulate::use_condaenv(condaenv = "/Users/phoebe.woodworth/Library/r-miniconda", required = TRUE)
# reticulate::conda_install('r-reticulate', 'python-kaleido')
# reticulate::conda_install('r-reticulate', 'plotly', channel = 'plotly')
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2025

# Set path to variable: Atmospheric_CO2
# This is where the data are and where the plots will go
Dir <- here("Atmospheric_CO2")

# Check if it exists; if not, create it
if (!dir.exists(Dir)) {
  dir.create(Dir)
}

# The csvs go in their own directory, so we are making that here if it doesn't already exist
Dir_csv <- here(paste0("PelagicClimate_",RptYr))
# Check if it exists; if not, create it
if (!dir.exists(Dir_csv)) {
  dir.create(Dir_csv)
}

```

```{r}
#| include: false
### Load data
# https://gml.noaa.gov/ccgg/trends/data.html
# Note that the number of lines to skip sometimes changes from year to year, which is frustrating.
# Monthly CO2 concentration
atm_co2 <- read_csv(file = paste(Dir, '/co2_mm_mlo_', RptYr, '.csv', sep = ""), skip = 40)

# Annual Growth Rates
ann_gr <- read_csv(file = paste(Dir, '/co2_gr_mlo_', RptYr, '.csv', sep = ""), skip = 45)

# Annual concentrations, so that we're not providing a potentially different value
ann_co2 <- read_csv(file = paste(Dir, '/co2_annmean_mlo_', RptYr, '.csv', sep = ""), skip = 43)
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# Pull out the ANNUAL values we need for the report
ann_mean_idx <- which(ann_co2$year == RptYr) 
ann_mean <- ann_co2$mean[ann_mean_idx] 
max_ann_mean <- max(ann_co2$mean) 

if (ann_mean != max_ann_mean) {
  print('The highest annual value was prior to the report year.')
}

# Pull out the MONTHLY values we need for the report & plots
yr_of_int <- which(atm_co2$year == RptYr)
prev_yrs <- which(atm_co2$year < RptYr)
all_yrs <- which(atm_co2$year <= RptYr)
monthly_max_RptYr <- max(atm_co2$average[yr_of_int])
monthly_max_PrevYrs <- max(atm_co2$average[prev_yrs])

if (monthly_max_PrevYrs > monthly_max_RptYr) {
  print('The highest monthly value was prior to the report year.')
}

# Confirm current text hasn't changed for some reason
pass_350 <- which(ann_co2$mean >= 350)
pass_400 <- which(ann_co2$mean >= 400)

if (ann_co2$year[pass_350[1]] != 1988) {
  print('Reevaluate the year that annual mean CO2 passed 350 ppm.')
}

if (ann_co2$year[pass_400[1]] != 2015) {
  print('Reevaluate the year that annual mean CO2 passed 400 ppm.')
}
```

```{r}
#| include: false
# ### Create growth rate plot for reference
# 2024: spudkgray --> whale_gray
seasonadj <- nmfspalette::nmfs_cols("whale_gray") # seasonally adj
pdf(paste(Dir, '/AtmosphericCO2growth_ts_', RptYr, '.pdf', sep = ""), width = 5, height = 3.5)
plot(ann_gr$year, ann_gr$`ann inc`, type = "l", lwd = 2, col = seasonadj,
     xlim = c(min(ann_gr$year), RptYr), ylim = c(0, 3.5),
     xlab = "Year", ylab = "Parts per Million (ppm)",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
axis((1), at = seq(1960, 2020, 5), tck = 0.025)
axis((2), at = seq(0, 3.5, 0.5), tck = 0.025, las = 1)
axis((3), at = seq(1960, 2020, 5), tck = 0.025, labels = FALSE)
axis((4), at = seq(0, 3.5, 0.5), tck = 0.025, labels = FALSE)
dev.off()
```

```{r}
#| include: false
# Write csv for portal
# Renaming 'mean' to 'ppm'
# Note that output csvs go in their own folder
AtmosphericCO2 <- ann_co2 %>% select(year, mean)
AtmosphericCO2 <- rename(AtmosphericCO2, ppm = mean)
write_csv(AtmosphericCO2, file = paste(here(), '/PelagicClimate_', RptYr, '/AtmosphericCO2_', RptYr, '.csv', sep = ""))
```

```{r}
#| include: false
# Write csv for dashboard
# Note that dashboard output has its own folder
# Thanks to Emily Conklin for this chunk of code!
CO2_dashboard <- atm_co2 %>%
  mutate(Date_Time = date_decimal(`decimal date`) %>% 
           format('%d-%b-%Y %H:%M') %>%
           toupper) %>%
  mutate(ID = "CO2") %>%
  select(Date_Time, Year = year, Month = month, Value = average, Anom = deseasonalized, ID) %>%
  mutate(Value_lm = NA, Anom_lm = NA, Units = "ppm")

write_csv(CO2_dashboard, file = here("Indicator_Dashboard", "Data",
                                     paste('CO2_Dashboard_Data_', RptYr, '.csv', sep = "")))
```

```{r}
#| include: false
# 2024 update: color for annual values changed from #D0D0D0 to #A5AAAF

# Borrowing code from the dashboard for this chunk
# so that figures look the same across products
indicator_data <- CO2_dashboard |>
  filter(Year <= RptYr)

# Create color palette for easy reference 
oceans <- nmfs_palette("oceans")(3) # 1 = report_year, 3 = previous years
crustacean <- nmfs_palette("crustacean")(4) # 1 = linear trend
coral <- nmfs_palette("coral")(3) # 3 = annual average for Rpt Yr
ann_grey <- "#A5AAAF" # annual means; in NMFS branding guide as dove gray tint 1
waves <- nmfs_palette("waves")(3) # annual means; in NMFS branding guide but not in package
seagrass <- nmfs_palette("seagrass")(3)
pal <- c(oceans[3], coral[2], waves[2], coral[3], crustacean[2])

# Formatting
plot_title_font <- list(size = 14)
plot_height <- 350 #in pixels

# Calculate annual means 
ann_vals <- indicator_data |>
  group_by(Year) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Identify the current year, to overlay on plot
given_yr <- indicator_data |>
  filter(Year == RptYr)

given_yr_ann <- bind_cols(rep(ann_vals$Value[dim(ann_vals)[1]]),
                          given_yr$Date_Time)
given_yr_ann <- rename(given_yr_ann, Value = ...1)
given_yr_ann <- rename(given_yr_ann, Date_Time = ...2)

# Commented out the plotly code because it's not working this year.  So,
# going with base R, because that's what I'm most familiar with
# Fair warning, this coding is tedious.

# x axis ticks and grid
x_grid <- c(min(dmy_hm(indicator_data$Date_Time)), 
                dmy(01011960),
                dmy(01011970),
                dmy(01011980),
                dmy(01011990),
                dmy(01012000),
                dmy(01012010),
                dmy(01012020),
                max(dmy_hm(indicator_data$Date_Time)))

pdf(paste(Dir, '/AtmosphericCO2_ts_', RptYr, '.pdf', sep = ""))
par(mar=c(8.6, 5.1, 2.1, 3.1))
plot(dmy_hm(indicator_data$Date_Time), indicator_data$Value, type = "l", lwd = 2, 
     xlim = c(min(dmy_hm(indicator_data$Date_Time)), max(dmy_hm(indicator_data$Date_Time))+(86400*16)), ylim = c(300, 435), 
     xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
     xaxs = "i", yaxs = "i", # remove the padding R adds
     bty = "n", # remove box
     col = pal[1],
     xlab = "Year", ylab = "Parts per Million (ppm)",
     # put a grid behind the plot
     panel.first = c(abline(v = x_grid, col = "#F1F3F3"),
                     abline(h = c(seq(300, 420, 20), 435), col = "#F1F3F3")),
     font.lab = 2)
par(new = TRUE)
plot(ymd_hm(paste(ann_vals$Year, '0601 00:00', sep = "")), ann_vals$Value, type = "l", lwd = 2,
     xlim = c(min(dmy_hm(indicator_data$Date_Time)), max(dmy_hm(indicator_data$Date_Time))+(86400*16)), ylim = c(300, 435), 
     xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
     xaxs = "i", yaxs = "i", # remove the padding R adds
     bty = "n", # remove box
     col = pal[3],
     xlab = " ", ylab = " ")
par(new = TRUE)
plot(dmy_hm(given_yr$Date_Time), given_yr$Value, type = "l", lwd = 2, 
     xlim = c(min(dmy_hm(indicator_data$Date_Time)), max(dmy_hm(indicator_data$Date_Time))+(86400*16)), ylim = c(300, 435), 
     xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
     xaxs = "i", yaxs = "i", # remove the padding R adds
     bty = "n", # remove box
     col = pal[2],
     xlab = " ", ylab = " ")
par(new = TRUE)
plot(dmy_hm(given_yr_ann$Date_Time), given_yr_ann$Value, type = "l", lwd = 2,
     xlim = c(min(dmy_hm(indicator_data$Date_Time)), max(dmy_hm(indicator_data$Date_Time))+(86400*16)), ylim = c(300, 435), 
     xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
     xaxs = "i", yaxs = "i", # remove the padding R adds
     bty = "n", # remove box
     col = pal[4],
     xlab = " ", ylab = " ")
axis(1, at = x_grid, label = c(" ", seq(1960, 2025, 10), " "), col = "transparent", col.ticks = "transparent") 
axis(2, at = c(seq(300, 420, 20), 435), las = 1, col = "transparent", col.ticks = "transparent")
legend('bottom', legend=c(paste0('Monthly average, 1958-',RptYr-1), paste0('Annual average, 1958-',RptYr-1), paste0('Monthly average, ',RptYr), paste0('Annual average, ',RptYr)), col=c(pal[1],pal[3], pal[2], pal[4]), lty=1, lwd=1.5, bg='gray80', horiz=F, ncol=2, inset = -0.31, xpd=T, box.lwd = 0, box.col = 'gray80')
dev.off()

# p1 <- plot_ly(indicator_data, x = dmy_hm(indicator_data$Date_Time), y = ~Value,
#               type = "scatter", mode = "lines", line = list(color = pal[1]),
#               height = plot_height, name = "Monthly average") |>
#   add_trace(indicator_data, x = dmy_hm(indicator_data$Date_Time), y = ~Value_lm,
#             type = "scatter", mode = "lines", line = list(color = pal[5]),
#             name = "Long-term Trend") |>
#   add_trace(ann_vals, x = ymd_hm(paste(ann_vals$Year, '0601 00:00', sep = "")), y = ann_vals$Value,
#             type = "scatter", mode = "lines", line = list(color = pal[3]),
#             name = "Annual average") |>
#   add_trace(given_yr, x = dmy_hm(given_yr$Date_Time), y = given_yr$Value,
#             type = "scatter", mode = "lines", line = list(color = pal[2]),
#             name = paste("Monthly average,", RptYr)) |>
#   add_trace(given_yr_ann, x = dmy_hm(given_yr_ann$Date_Time), y = given_yr_ann$Value,
#             type = "scatter", mode = "lines", line = list(color = pal[4]),
#             name = paste("Annual average,", RptYr))
# 
# #apply same layout parameters for all plots
# 
# #custom x axis (min, every decade, report year)
# all_years <- unique(indicator_data$Year)
# first_date <- as.character(parse_date_time(indicator_data$Date_Time[1], orders = "d-b-Y H:M"))
# date_axis <- c(first_date, 
#               all_years[which(all_years %% 10 == 0)], 
#               RptYr)
# 
# p1 <- p1 |> layout(title = list(text = "Indicator Time Series", x = 0.01, font = plot_title_font), #add title
#                    xaxis = list(type = "date", tickformat = "%Y", tickmode = "array", tickvals = date_axis, tickangle = 90),
#                    #xaxis = list(tick0 = min(indicator_data$Date_Time), dtick = "M24"),
#                    yaxis = list(title = indicator_data$Units[1], hoverformat = '.3f', range = list(300, 430), tickvals = list(300, 320, 340, 360, 380, 400, 420, 430)), #add units and round values in hover display; not sure of a better way to set axis limits and ticks...
#                    paper_bgcolor = 'transparent', plot_bgcolor = 'transparent', #transparent background
#                    hovermode = "x unified", #show data for all traces simultaneously
#                    hoverlabel = list(namelength = -1)) #don't cutoff hoverinfo due to length
#   
#return plot - oddly works in R but results in an error when rendering
# save_image(p1, paste(Dir, '/AtmosphericCO2_ts_', RptYr, '.pdf', sep = ""))

# NOTE - added below in the library import chunk at the top
# Handy reference, in case this doesn't work in the future
# from: https://stackoverflow.com/questions/64028289/export-plot-from-plotly-into-pdf
# install.packages('reticulate')
# reticulate::install_miniconda()
# reticulate::conda_install('r-reticulate', 'python-kaleido')
# reticulate::conda_install('r-reticulate', 'plotly', channel = 'plotly')
# reticulate::use_miniconda('r-reticulate')
# To export:
# 
# library(plotly)
# kaleido(FINAL_plot, "FINAL_plot.pdf")

# ## ggplot version with legend
# #devtools::install_github("teunbrand/elementalist")
# legendInfo <- as.factor(c(paste0('Mothly average, 1958-', RptYr-1), paste0('Annual average, 1958-', RptYr-1), paste0('Mothly average, ', RptYr), paste0('Annual average, ', RptYr)))
# levels(legendInfo) <- levels(legendInfo)[c(3,4,1,2)]
# dmy_hm(paste0('01-01-',ann_vals$Year,' 00:00'))
# 
# ggplot() + 
#   geom_path(data=indicator_data, aes(dmy_hm(Date_Time), Value, color=legendInfo[2])) + 
#   geom_path(data=ann_vals, aes(dmy_hm(paste0('01-01-',Year,' 00:00')), Value, color=legendInfo[1])) + 
#   geom_path(data=given_yr, aes(dmy_hm(Date_Time), Value, color=legendInfo[4])) + 
#   geom_path(data=given_yr_ann, aes(dmy_hm(Date_Time), Value, color=legendInfo[3])) + 
#   scale_x_datetime(breaks = as.POSIXct(seq.Date(dmy('01-01-1960'), dmy('01-01-2025'),by='10 year')), date_labels = '%Y', expand = F) + 
#   scale_y_continuous(expand = F, breaks = seq(300,440,20), limits = c(300,440)) + 
#   ylab('Parts per Million (ppm)') + 
#   xlab('Year') + 
#   theme_minimal() + 
#    theme(panel.grid.minor = element_blank(), axis.title = element_text(size=14, face="bold"), legend.position = 'bottom', legend.background = element_rect_round(fill = 'gray80', color='gray80')) +
#     scale_color_manual(name=NULL, values=pal[c(1,3,2,4)]) + 
#   guides(color = guide_legend(ncol = 2))
  


```

Rationale: Atmospheric carbon dioxide is a measure of what human activity has already done to affect the climate system through greenhouse gas emissions. It provides quantitative information in a simplified, standardized format that decision makers can easily understand. This indicator demonstrates that the concentration (and, in turn, warming influence) of greenhouse gases in the atmosphere has increased substantially over the last several decades.  
\
Status: Atmospheric CO~2~ is increasing exponentially. This means that atmospheric CO~2~ is increasing more quickly over time. In `r RptYr`, the annual mean concentration of CO~2~ was `r ann_mean` ppm.  This is the highest annual value recorded. This year also saw the highest monthly value, which was `r monthly_max_RptYr` ppm. In 1959, the first year full of the time series, the atmospheric concentration of CO~2~ was 316 ppm. The annual mean passed 350 ppm in 1988, and 400 ppm in 2015.  
\
Description: Monthly mean atmospheric carbon dioxide (CO~2~) at Mauna Loa Observatory, Hawaiʻi in parts per million (ppm) from March 1958 to present. The observed increase in monthly average carbon dioxide concentration is primarily due to CO~2~ emissions from fossil fuel burning. Carbon dioxide remains in the atmosphere for a very long time, and emissions from any location mix throughout the atmosphere in approximately one year. The annual variations at Mauna Loa, Hawaiʻi are due to the seasonal imbalance between the photosynthesis and respiration of terrestrial plants. During the summer growing season, photosynthesis exceeds respiration, and CO~2~ is removed from the atmosphere. In the winter (outside the growing season), respiration exceeds photosynthesis, and CO~2~ is returned to the atmosphere. The seasonal cycle is strongest in the northern hemisphere because of its larger land mass.  
\
Timeframe: Annual, monthly.  
\
Region/Location: Mauna Loa, Hawaiʻi, but representative of global atmospheric carbon dioxide concentration.  Note that due to the eruption of the Mauna Loa Volcano, measurements from Mauna Loa Observatory were suspended as of 29 November 2022. Observations from December 2022 to 4 July 2023 are from a site at the Maunakea Observatories, approximately 21 miles north of the Mauna Loa Observatory.  Mauna Loa observations resumed in July 2023.  
\
Measurement Platform: *In-situ* station.  
\
Data available at: <https://gml.noaa.gov/ccgg/trends/data.html>.   
\
Sourced from: Keeling et al. (1976), Thoning et al. (1989), and NOAA (2025a).  
\
Graphics produced in part using Stawitz and Li (2025).

## Additional Information

The following .csv files were downloaded from <https://gml.noaa.gov/ccgg/trends/data.html>:

 - Mauna Loa CO2 monthly mean data (co2_mm_mlo.csv)
 - Mauna Loa CO2 annual mean data  (co2_annmean_mlo.csv)
 - Mauna Loa CO2 annual mean growth rates  (co2_gr_mlo.csv)
 
These files were manually appended with the report year (i.e., `r RptYr`).  Information from these data was used to lend detail and context to the indicator status.  

The files could be accessed directly without downloading them first.  I decided against this only so that there's a full record of the data used to create the indicators.  But I'm curious what other think.  If you have an opinion, please start a discussion or open an issue.