---
title: "El_Nino_Southern_Oscillation"
author: "phoebe.woodworth-jefcoats@noaa.gov"
format:
  docx:
    reference-doc: SAFE-Reference-Doc.docx
---

## EL NIÑO – SOUTHERN OSCILLATION

```{r}
#| include: false
### Load libraries
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
# remotes::install_github("nmfs-fish-tools/nmfspalette")
library(nmfspalette)
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
# Note that there's a one year lag for HOTS data
RptYr <- 2022

# Set path to variable: Atmospheric_CO2
# This is where the data are and where the plots will go
Dir <- here("El_Nino_Southern_Oscillation")
```

```{r}
#| include: false
### Load data
ONI_full <- read_csv(file = paste(Dir, '/ONI_', RptYr, '.csv', sep = ""))
```

```{r}
#| include: false
### Round data to the tenth of a degree
ONI_rounded <- round(ONI_full$ANOM, digits = 1)
```

```{r}
#| include: false
### Identifying El Niños and La Niñas, based on the criterion that they persist for at least 5 consecutive seasons
ElNino_idx = which(ONI_rounded >= 0.5);
LaNina_idx = which(ONI_rounded <= -0.5);
Neutral_idx = which(ONI_rounded > -0.5 & ONI_rounded < 0.5);
ElNino <- array(0, dim = c(length(ONI_rounded), 3))
LaNina <- array(0, dim = c(length(ONI_rounded), 3))
Neutral <- array(0, dim = c(length(ONI_rounded), 3))

ElNino[ElNino_idx] = ONI_rounded[ElNino_idx]
LaNina[LaNina_idx] = ONI_rounded[LaNina_idx]
Neutral[Neutral_idx] = ONI_rounded[Neutral_idx]

if (ElNino[1,1] > 0) {
  ElNino[1,2] = 1
  ElNino[1,3] = 1
}

if (LaNina[1,1] < 0) {
  LaNina[1,2] = 1
  LaNina[1,3] = 1
}
    
if (Neutral[1,1] != 0) {
  Neutral[1,2] = 1
  Neutral[1,3] = 1
}

for (r in seq(2, length(ONI_rounded), 1)) {
  if (ElNino[r,1] > 0) {
    ElNino[r,2] = 1
    ElNino[r,3] = ElNino[r,2] + ElNino[r - 1,3]
  }
  
  if (LaNina[r,1] < 0) {
    LaNina[r,2] = 1
    LaNina[r,3] = LaNina[r,2] + LaNina[r - 1,3]
  }
  
  if (Neutral[r,1] != 0) {
    Neutral[r,2] = 1
    Neutral[r,3] = Neutral[r,2] + Neutral[r - 1,3]
  }
}


for (l in seq(4, 1, -1)) {
  for (r in seq(2, length(ONI_rounded), 1)) {
   if (ElNino[r,3] == 0 && ElNino[r - 1,3] <= l) {
     ElNino[r - 1,3] = 0
   }
   if (LaNina[r,3] == 0 && LaNina[r - 1,3] <= l) {
      LaNina[r - 1,3] = 0
   }
   if (Neutral[r,3] == 0 && Neutral[r - 1,3] <= l) {
      Neutral[r - 1,3] = 0
   }
  }
}

pos_idx = which(ElNino[,3] == 0)
ElNino[pos_idx,1] = NA
neg_idx = which(LaNina[,3] == 0)
LaNina[neg_idx,1] = NA
neu_idx = which(Neutral[,3] == 0)
Neutral[neu_idx,1] = NA
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# Pull out the values we need for the report 
yr_of_int <- which(ONI_full$YR == RptYr)
prev_yrs <- which(ONI_full$YR < RptYr)
all_yrs <- which(ONI_full$YR <= RptYr)
monthly_min_RptYr <- min(ONI_full$ANOM[yr_of_int])
monthly_min_PrevYrs <- min(ONI_full$ANOM[prev_yrs])
monthly_max_RptYr <- max(ONI_full$ANOM[yr_of_int])
monthly_max_PrevYrs <- max(ONI_full$ANOM[prev_yrs])

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The highest monthly value was in the report year.')
}

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The lowest monthly value was in the report year.')
}
```

```{r}
#| include: false
### Create plot for report
# Note: plotting the neutral phases is just an error check, not part of the final figure

# x-axes
# Identify time period from Jan 2000 forward
x_2000 <- which(ONI_full$YR >= 2000)
x_all <- which(ONI_full$YR <= 2022)

# Create axes limit values, to make things simpler
ONI_xlim_full <- c(1, length(ONI_rounded))
ONI_xlim_2000 <- c(min(x_2000), max(x_2000))
ONI_ylim <- c(-3, 3)

# Plot
pdf(paste(Dir, '/ONI_ts_full_', RptYr, '.pdf', sep = ""), width = 7, height = 4.5)
plot(x_all, ONI_rounded[x_all], type = "l", lwd = 2, col = "#D0D0D0", 
     xlim = ONI_xlim_full, ylim = ONI_ylim, xlab = 'Year', ylab = 'Oceanic Niño Index (ONI)',
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(x_all, ElNino[x_all], type = "l", lwd = 2, col = "red", 
     xlim = ONI_xlim_full, ylim = ONI_ylim, xlab = ' ', ylab = ' ',
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(x_all, LaNina[x_all], type = "l", lwd = 2, col = "blue", 
     xlim = ONI_xlim_full, ylim = ONI_ylim, xlab = ' ', ylab = ' ',
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
# par(new = TRUE)
# plot(x_all, Neutral[x_all], type = "l", lwd = 2, col = "yellow", 
#      xlim = ONI_xlim_full, ylim = ONI_ylim, xlab = ' ', ylab = ' ',
#      xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(c(min(x_all), max(x_all)), c(0, 0), type = "l", lwd = 1, col = "black", 
     xlim = ONI_xlim_full, ylim = ONI_ylim, xlab = ' ', ylab = ' ',
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
axis((1), at = c(seq(1, max(x_all) - 35, 120), max(x_all) - 11), tck = 0.025, labels = c(seq(1950, 2020, 10), 2022))
axis((2), at = c(seq(-3, 3, 1)), tck = 0.025, las = 1)
axis((3), at = c(seq(1, max(x_all) - 35, 120), max(x_all) - 11), tck = 0.025, labels = FALSE)
axis((4), at = c(seq(-3, 3, 1)), tck = 0.025, labels = FALSE)
dev.off()

pdf(paste(Dir, '/ONI_ts_2000s_', RptYr, '.pdf', sep = ""), width = 7, height = 4.5)
plot(x_2000, ONI_rounded[x_2000], type = "l", lwd = 2, col = "#D0D0D0", 
     xlim = ONI_xlim_2000, ylim = ONI_ylim, xlab = 'Year', ylab = 'Oceanic Niño Index (ONI)',
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(x_2000, ElNino[x_2000], type = "l", lwd = 2, col = "red", 
     xlim = ONI_xlim_2000, ylim = ONI_ylim, xlab = ' ', ylab = ' ',
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(x_2000, LaNina[x_2000], type = "l", lwd = 2, col = "blue", 
     xlim = ONI_xlim_2000, ylim = ONI_ylim, xlab = ' ', ylab = ' ',
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
# par(new = TRUE)
# plot(x_2000, Neutral[x_2000], type = "l", lwd = 2, col = "yellow", 
#     xlim = ONI_xlim_2000, ylim = ONI_ylim, xlab = ' ', ylab = ' ',
#     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(c(min(x_2000), max(x_2000)), c(0, 0), type = "l", lwd = 1, col = "black", 
     xlim = ONI_xlim_2000, ylim = ONI_ylim, xlab = ' ', ylab = ' ',
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
axis((1), at = c(seq(min(x_2000), max(x_2000) - 35, 60), max(x_2000) - 11), tck = 0.025, labels = c(seq(2000, 2020, 5), 2022))
axis((2), at = c(seq(-3, 3, 1)), tck = 0.025, las = 1)
axis((3), at = c(seq(min(x_2000), max(x_2000) - 35, 60), max(x_2000) - 11), tck = 0.025, labels = FALSE)
axis((4), at = c(seq(-3, 3, 1)), tck = 0.025, labels = FALSE)
dev.off()
# _axt = "n" removes tick labels, _axs = "i" removes whitespace beyond axes maxima
# bty = "n" removes the box around the legend
```

Rationale: The El Niño – Southern Oscillation (ENSO) cycle is known to have impacts on Pacific fisheries including tuna fisheries. The ONI focuses on ocean temperature, which has the most direct effect on these fisheries. 

Status: The Oceanic Niño Index (ONI) indicated La Niña conditions throughout `r RptYr`. In `r RptYr`, the ONI ranged from `r monthly_min_RptYr` to `r monthly_max_RptYr`.  This is within the range of values observed previously in the time series.

Description: The three-month running mean (referred to as a season) of satellite remotely-sensed sea surface temperature (SST) anomalies in the Niño 3.4 region (5°S – 5°N, 120° – 170°W). The Oceanic Niño Index (ONI) is a measure of the El Niño – Southern Oscillation (ENSO) phase. Warm and cool phases, termed El Niño and La Niña respectively, are based in part on an ONI threshold of ± 0.5 °C being met for a minimum of five consecutive overlapping seasons. Additional atmospheric indices are needed to confirm an El Niño or La Niña event, as the ENSO is a coupled ocean-atmosphere phenomenon. The atmospheric half of ENSO is measured using the Southern Oscillation Index.

Timeframe: Three-month running average.

Region/Location: Niño 3.4 region, 5°S – 5°N, 120° – 170°W.

Measurement Platform: *In-situ* station, satellite, model.

Data available at: <https://www.cpc.ncep.noaa.gov/data/indices/oni.ascii.txt>. 

Sourced from NOAA CPC (2023).

Graphics produced in part using Stawitz (2022).

## Additional Information

Data were manually copied and pasted into a .csv.  These data have precision to the hundredth of a degree.  They are subsequently rounded in this script to match the data on the more commonly used [CPC site](https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/ONI_v5.php), which are rounded to the tenth of a degree.  

No data are saved for the Council portal because ONI isn't examined as an annual mean.
