
## EL NIÑO – SOUTHERN OSCILLATION

```{r}
#| include: false

# This includes all commonly used packages and variables, like RptYr so you only have to change it once. 
source("_common.R")
```

```{r}
#| include: false
# Set path to variable: El_Nino_Southern_Oscillation
# This is where the data are and where the plots will go
Dir <- here("El_Nino_Southern_Oscillation")
# Check if it exists; if not, create it
if (!dir.exists(Dir)) {
  dir.create(Dir)
}
```

```{r}
#| include: false
### Load data
ONI_full <- read_csv(file = paste(Dir, '/ONI_', RptYr, '.csv', sep = ""))
```

```{r}
#| include: false
### Round data to the tenth of a degree
ONI_rounded <- round(ONI_full$ANOM, digits = 1)
```

```{r}
#| include: false
### Identifying El Niños and La Niñas, based on the criterion that they persist for at least 5 consecutive seasons
# First, identify values that are above or below the appropriate threshold
ElNino_idx = which(ONI_rounded >= 0.5);
LaNina_idx = which(ONI_rounded <= -0.5);
Neutral_idx = which(ONI_rounded > -0.5 & ONI_rounded < 0.5);

# Create empty arrays of the correct size
ElNino <- array(0, dim = c(length(ONI_rounded), 3))
LaNina <- array(0, dim = c(length(ONI_rounded), 3))
Neutral <- array(0, dim = c(length(ONI_rounded), 3))
Neutral[,1] <- NA

# Fill each phase's array with values that pass the appropriate threshold
ElNino[ElNino_idx] = ONI_rounded[ElNino_idx]
LaNina[LaNina_idx] = ONI_rounded[LaNina_idx]
Neutral[Neutral_idx] = ONI_rounded[Neutral_idx]

# This next part is a counter.  Fill the first row first, 
# flagging with a 1 values at meet that phase's threshold
if (ElNino[1,1] > 0) {
  ElNino[1,2] = 1
  ElNino[1,3] = 1
}

if (LaNina[1,1] < 0) {
  LaNina[1,2] = 1
  LaNina[1,3] = 1
}
    
if (!is.na(Neutral[1,1])) {
  Neutral[1,2] = 1
  Neutral[1,3] = 1
}

# Loop through the remaining rows
# Column 2 is a yes/no flag (1 if true)
# Column 3 is a counter
for (r in seq(2, length(ONI_rounded), 1)) {
  if (ElNino[r,1] > 0) {
    ElNino[r,2] = 1
    ElNino[r,3] = ElNino[r,2] + ElNino[r - 1,3]
  }
  
  if (LaNina[r,1] < 0) {
    LaNina[r,2] = 1
    LaNina[r,3] = LaNina[r,2] + LaNina[r - 1,3]
  }
  
  if (!is.na(Neutral[r,1])) {
    Neutral[r,2] = 1
    Neutral[r,3] = Neutral[r,2] + Neutral[r - 1,3]
  }
}

# For months where the counter maxes out at <5 sequential seasons
# set the counter to 0
for (l in seq(4, 1, -1)) {
  for (r in seq(2, length(ONI_rounded), 1)) {
   if (ElNino[r,3] == 0 && ElNino[r - 1,3] <= l) {
     ElNino[r - 1,3] = 0
   }
   if (LaNina[r,3] == 0 && LaNina[r - 1,3] <= l) {
      LaNina[r - 1,3] = 0
   }
   if (Neutral[r,3] == 0 && Neutral[r - 1,3] <= l) {
      Neutral[r - 1,3] = 0
   }
  }
}

# Identify only months that meet the >= 5 season criterion
pos_idx = which(ElNino[,3] == 0)
ElNino[pos_idx,1] = NA
neg_idx = which(LaNina[,3] == 0)
LaNina[neg_idx,1] = NA
neu_idx = which(Neutral[,3] == 0)
Neutral[neu_idx,1] = NA
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# Pull out the values we need for the report 
yr_of_int <- which(ONI_full$YR == RptYr)
prev_yrs <- which(ONI_full$YR < RptYr)
all_yrs <- which(ONI_full$YR <= RptYr)
monthly_min_RptYr <- min(ONI_full$ANOM[yr_of_int])
monthly_min_PrevYrs <- min(ONI_full$ANOM[prev_yrs])
monthly_max_RptYr <- max(ONI_full$ANOM[yr_of_int])
monthly_max_PrevYrs <- max(ONI_full$ANOM[prev_yrs])

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The highest monthly value was in the report year.')
}

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The lowest monthly value was in the report year.')
}
```

```{r}
#| include: false
# Write csv for dashboard
# Note that dashboard output has its own folder
# Thanks to Emily Conklin for this chunk of code!

month_list <- c("DJF", "JFM", "FMA", "MAM", "AMJ", "MJJ", 
                "JJA", "JAS", "ASO", "SON", "OND", "NDJ")

ENSO_dashboard <- ONI_full %>%
  rowwise() %>%
  mutate(Month = which(month_list == SEAS)) %>%
  mutate(Date_Time = paste0("15-", Month, "-", as.character(YR), " 00:00")) %>%
  mutate(Value_lm = NA, Anom = NA, Anom_lm = NA, ID = "ENSO", Units = "Oceanic Nino Index") %>%
  select(Date_Time, Year = YR, Month, Value = ANOM, Anom, ID, Value_lm, Anom_lm, Units) |>
  filter(Year <= RptYr)

write_csv(ENSO_dashboard, file = here("Indicator_Dashboard", "Data", paste('ENSO_Dashboard_Data_', RptYr, '.csv', sep = "")))
```

```{r}
#| include: false
# Borrowing code from the dashboard for this chunk
# so that figures look the same across products
indicator_data <- ENSO_dashboard |>
  filter(Year <= RptYr) %>% 
  mutate(Date_Time=dmy_hm(Date_Time))

# Create color palette for easy reference 
oceans <- nmfs_palette("oceans")(3) # 1 = report_year, 2 = La Niña, 3 = previous years
crustacean <- nmfs_palette("crustacean")(4) # 1 = linear trend
coral <- nmfs_palette("coral")(3) # 2 = El Niño,  3 = annual average for Rpt Yr
ann_grey <- "#A5AAAF" # annual means; in NMFS branding guide as dove gray tint 1
waves <- nmfs_palette("waves")(3) # annual means; in NMFS branding guide but not in package
seagrass <- nmfs_palette("seagrass")(3)
pal <- c(oceans[3], coral[2], waves[2], coral[3], crustacean[2])

# Formatting
plot_title_font <- list(size = 14)
plot_height <- 350 #in pixels

#calculate pos/neg/neutral ENSO values 
#needs to be greater than 0.5 or less than -0.5 for at least five consecutive seasons
indicator_data <- indicator_data |> mutate(Sign = sign(Value)) |>
  mutate(Sign = ifelse(test = (Value < 0.5 & Value > -0.5), yes = 0, no = Sign)) |> #positive or negative?
  group_by(Sign, Group = with(rle(Sign), rep(seq_along(lengths), lengths))) |> #how many seasons in a row?
  ungroup() |>
  add_count(Group) |>
  mutate(ENSO = "Neutral") |>
  mutate(Units = "Oceanic Niño Index (ONI)")
    
#add ENSO category
indicator_data[which(indicator_data$Sign == -1 & indicator_data$n >= 5),]$ENSO <- "La Niña"
indicator_data[which(indicator_data$Sign == 1 & indicator_data$n >= 5),]$ENSO <- "El Niño"

# Commented out the plotly code because it's not working this year.  So,
# going with base R, because that's what I'm most familiar with
# Fair warning, this coding is tedious.

# x axis ticks and grid
x_grid <- c(seq.POSIXt(min(indicator_data$Date_Time), max(indicator_data$Date_Time), by='10 year'),
            max(indicator_data$Date_Time))
# x axis limit
ts_x_lim <- c(min(indicator_data$Date_Time), max(indicator_data$Date_Time))

# Write a function to do the stem plot (we'll call it 'stem' like Matlabl does)
# Thanks to https://www.r-bloggers.com/2009/11/matlab-style-stem-plot-with-r/
stem <- function(x, y, stemcol) {
  for (l in seq(1, length(x), 1)) {
    lines(c(x[l], x[l]), c(0, y[l]), col = stemcol)
  }
}

#pdf(paste(Dir, '/ONI_ts_', RptYr, '.pdf', sep = ""), width=8, height=5)
par(mar=c(4.1, 4.1, 2.1, 6.1))
plot(c(min(indicator_data$Date_Time), max(indicator_data$Date_Time)), c(0, 0), col = "black", 
     type = "l", lwd = 1,
     xlim = ts_x_lim, ylim = c(-3, 3),
     xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
     xaxs = "i", yaxs = "i", # remove the padding R adds
     bty = "n", # remove box
     xlab = "Year", ylab = "Oceanic Niño Index (ONI)",
     # put a grid behind the plot
     panel.first = c(abline(v = x_grid, col = "#F1F3F3"),
                     abline(h = c(seq(-3, 3, 1)), col = "#F1F3F3")),
     font.lab = 2)
par(new = TRUE)
stem(indicator_data$Date_Time[ElNino_idx], indicator_data$Value[ElNino_idx], pal[2])
stem(indicator_data$Date_Time[LaNina_idx], indicator_data$Value[LaNina_idx], oceans[2])
stem(indicator_data$Date_Time[Neutral_idx], indicator_data$Value[Neutral_idx], ann_grey)
axis(1, at = x_grid, label = c(seq(1950, 2025, 10), " "), col = "transparent", col.ticks = "transparent", xpd=T) 
axis(2, at = c(seq(-3, 3, 1)), las = 1, col = "transparent", col.ticks = "transparent")
legend('topright',
       legend=c('El Niño', 'La Niña', 'Neutral'),
       col=c(pal[2], oceans[2], ann_grey),
       pch=c(15,15,15),
       pt.cex = 2,
       bty='n', 
       horiz = F,
       inset = c(-0.17,-0.03), xpd=T,
       x.intersp=1.3)
# 2. Save the displayed plot to an object
my_plot_object <- recordPlot()

# 3. Clear the current plot window (optional test)
dev.off()


```

```{r}
#| echo: false
#| label: fig-ch3-ENSO
#| fig.cap: 'Oceanic Niño Index from 1950–2025 El Niño periods in red, La Niña periods in blue, and neutral periods in grey'
#| fig.align: 'center'
#| fig-width: 8
#| fig-height: 5
#| out-width: "100%"

my_plot_object
```

**Rationale:** The El Niño – Southern Oscillation (ENSO) cycle is known to have impacts on Pacific fisheries including tuna fisheries. The ONI focuses on ocean temperature, which has the most direct effect on these fisheries.   
\
**Status:** The Oceanic Niño Index (ONI) remained in largely neutral conditions in `r RptYr`. In `r RptYr`, the ONI ranged from `r monthly_min_RptYr` to `r monthly_max_RptYr`.  This is within the range of values observed previously in the time series.  
\
**Description:** The three-month running mean (referred to as a season) of satellite remotely-sensed sea surface temperature (SST) anomalies in the Niño 3.4 region (5°S – 5°N, 120° – 170°W). The Oceanic Niño Index (ONI) is a measure of the El Niño – Southern Oscillation (ENSO) phase. Warm and cool phases, termed El Niño and La Niña respectively, are based in part on an ONI threshold of ± 0.5 °C being met for a minimum of five consecutive overlapping seasons. Additional atmospheric indices are needed to confirm an El Niño or La Niña event, as the ENSO is a coupled ocean-atmosphere phenomenon. The atmospheric half of ENSO is measured using the Southern Oscillation Index.  
\
**Timeframe:** Three-month running average.  
\
**Region/Location:** Niño3.4 region, 5°S – 5°N, 120° – 170°W.  
\
**Measurement Platform:** *In-situ* station, satellite, model.  
\
**Data available at:** <https://www.cpc.ncep.noaa.gov/data/indices/oni.ascii.txt>.   
\
**Sourced from:** @NOAA-CPCUnknown-po. Graphics produced in part using @Stawitz2026-kw.  
\

<!--
## Additional Information

Data were manually copied and pasted into a .csv.  These data have precision to the hundredth of a degree.  They are subsequently rounded in this script to match the data on the more commonly used [CPC site](https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/ONI_v5.php), which are rounded to the tenth of a degree.  

No data are saved for the Council portal because ONI isn't examined as an annual mean.

-->
