### Pacific Decadal Oscillation

```{r}
#| include: false

# This includes all commonly used packages and variables, like RptYr and eliminates the need to change them in every chapter. 
source("_common.R")
```

```{r}
#| include: false
# Set path to variable: Pacific_Decadal_Oscillation
# This is where the data are and where the plots will go
Dir <- here("Pacific_Decadal_Oscillation")
# Check if it exists; if not, create it
if (!dir.exists(Dir)) {
  dir.create(Dir)
}
```

```{r}
#| include: false
### Load data
PDO_full <- read_csv(file = paste(Dir, '/PDO_', RptYr, '.csv', sep = ""))
```

```{r}
#| include: false
### Rename the PDO column to something reasonable
PDO_full <- rename(PDO_full, PDO = `PDO from ERSST V5 https://psl.noaa.gov/pdo/ Using EOF from 1920 to 2014 for N Pacific (see webpage)`)
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# Pull out the values we need for the report 
yr_of_int <- which(year(PDO_full$Date) == RptYr)
prev_yrs <- which(year(PDO_full$Date) < RptYr)
all_yrs <- which(year(PDO_full$Date) <= RptYr)
monthly_min_RptYr <- min(PDO_full$PDO[yr_of_int])
monthly_min_PrevYrs <- min(PDO_full$PDO[prev_yrs])
monthly_max_RptYr <- max(PDO_full$PDO[yr_of_int])
monthly_max_PrevYrs <- max(PDO_full$PDO[prev_yrs])

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The highest monthly value was in the report year.')
}

if (monthly_min_RptYr > monthly_min_PrevYrs) {
  print('The lowest monthly value was in the report year.')
}
```

```{r}
#| include: false
# Write csv for dashboard
# Note that dashboard output has its own folder
# Thanks to Emily Conklin for this chunk of code!

PDO_dashboard <- PDO_full %>%
  mutate(Date_Time = as_datetime(Date), Year = year(Date_Time), 
         Month = month(Date_Time)) %>%
  mutate(Date_Time = Date_Time %>%
           format('%d-%b-%Y %H:%M') %>%
           toupper) %>%
  select(Date_Time, Year, Month, Value = PDO) %>%
  mutate(Anom = NA, ID = "PDO", Value_lm = NA, Anom_lm = NA, Units = "Pacific Decadal Oscillation") |>
  filter(Year <= RptYr) 

#write_csv(PDO_dashboard, file = here("Indicator_Dashboard", "Data", paste('PDO_Dashboard_Data_', RptYr, '.csv', sep = "")))
```

```{r}
#| include: false
# Borrowing code from the dashboard for this chunk
# so that figures look the same across products
indicator_data <- PDO_dashboard |>
  filter(Year <= RptYr) %>% 
  mutate(Date_Time=dmy_hm(Date_Time))

# Create color palette for easy reference 
oceans <- nmfs_palette("oceans")(3) # 1 = report_year, 2 = Negative PDO, 3 = previous years
crustacean <- nmfs_palette("crustacean")(4) # 1 = linear trend
coral <- nmfs_palette("coral")(3) # 2 = Positive PDO,  3 = annual average for Rpt Yr
ann_grey <- "#A5AAAF" # annual means; in NMFS branding guide but not in package
waves <- nmfs_palette("waves")(3) # annual means; in NMFS branding guide but not in package
seagrass <- nmfs_palette("seagrass")(3)
pal <- c(oceans[3], coral[2], waves[2], coral[3], crustacean[2])

# Formatting
plot_title_font <- list(size = 14)
plot_height <- 350 #in pixels

indicator_data <- indicator_data |>
  mutate(Units = "Pacific Decadal Oscillation (PDO)")

# x axis ticks and grid
x_grid <- c(min(indicator_data$Date_Time), 
            seq.POSIXt(dmy_hm(010118600000), max(indicator_data$Date_Time), by='20 year'),
            max(indicator_data$Date_Time))

x_grid <- seq.POSIXt(dmy_hm(010118600000), max(indicator_data$Date_Time), by='20 year')

# x axis limit
ts_x_lim <- c(min(indicator_data$Date_Time), max(indicator_data$Date_Time))

# Identify positive and negative phases
indicator_data$Phase <- ifelse(indicator_data$Value > 0, 'Positive PDO', 'Negative PDO')

# Defining the theme up front so we don't have to repeat it
plotTheme <- theme_minimal() +
  theme(panel.border = element_rect(color='#F1F3F3'),
        panel.grid.major = element_line(color='#F1F3F3'),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        axis.title = element_text(size=14, face="bold"), 
        axis.text = element_text(size=12),
        legend.position = 'inside', 
        legend.position.inside = c(0.9,0.92),
        legend.text = element_text(size=12), 
        legend.key.spacing.x = unit(1.5,'cm'),
        legend.spacing.y = unit(0.1,'cm'))

# Make stem type plot using ggplot
p <- ggplot(indicator_data) + 
  geom_segment(aes(x = Date_Time, y = 0, xend = Date_Time, yend = Value, colour = Phase), linewidth = 0.3) + 
  scale_color_manual(values = c(oceans[2], pal[2]), name=NULL) + 
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = expansion(mult = c(0, 0.02))) +
   scale_y_continuous(labels = seq(-4,4,by=1), breaks = seq(-4,4,by=1), limits = c(-4,4)) +
  xlab('Year') +
  ylab('Pacific Decadal Oscillation (PDO)') +
  plotTheme +
  guides(color = guide_legend(override.aes = list(linewidth = 3), byrow=T))  # Make the lines in the legend thicker

```

```{r}
#| echo: false
#| label: fig-ch4-PDO
#| fig.cap: 'Pacific Decadal Oscillation from 1854–2025 with positive warm periods in red and negative cool periods in blue'
#| fig.align: 'center'
#| fig-width: 8
#| fig-height: 5
#| out-width: "90%"

p
```


**Rationale:** The Pacific Decadal Oscillation (PDO) was initially named by fisheries scientist Steven Hare in 1996 while researching connections between Alaska salmon production cycles and Pacific climate. Like ENSO, the PDO reflects changes between periods of persistently warm or persistently cool ocean temperatures, but over a period of 20 to 30 years (versus six to 18 months for ENSO events). The climatic fingerprints of the PDO are most visible in the Northeastern Pacific, but secondary signatures exist in the tropics. Research has shown that during positive phases of the PDO, the Hawaiʻi-based deep-set longline fishery has a greater likelihood of black-footed albatross sightings and interactions [@Wren2019-ss].  
\
**Status:** The PDO was `r ifelse(mean(indicator_data$Value[which(indicator_data$Year == RptYr)]) < 0, 'negative', 'positive')` in `r RptYr`. The index ranged from `r signif(monthly_min_RptYr, 3)` to `r signif(monthly_max_RptYr, 3)` over the course of the year.  This represents the strongest negative value in the time series, with the next two lowest values occurring in `r indicator_data$Year[order(indicator_data$Value)[2]]` and `indicator_data$Year[order(indicator_data$Value)[3]]`.  
\
**Description:** The PDO is often described as a long-lived El Niño-like pattern of Pacific climate variability. As seen with the better-known ENSO, extremes in the PDO pattern are marked by widespread variations in the Pacific Basin and the North American climate. In parallel with the ENSO phenomenon, the extreme cases of the PDO have been classified as either warm or cool, as defined by ocean temperature anomalies in the northeast and tropical Pacific Ocean. When SST is below average in the [central] North Pacific and warm along the North American coast, and when sea level pressures are below average in the North Pacific, the PDO has a positive value. When the climate patterns are reversed, with warm SST anomalies in the interior and cool SST anomalies along the North American coast, or above average sea level pressures over the North Pacific, the PDO has a negative value. Description inserted from @NOAA-PSLUnknown-jd.  
\
**Timeframe:** Monthly.  
\
**Region/Location:** Pacific Basin north of 20°N.  
\
**Measurement Platform:** *In-situ* station, satellite, model.  
\
**Data available at:** <https://psl.noaa.gov/pdo/>.  
\
**Sourced from:** @NOAA-PSLUnknown-jd, @Mantua1997-na, and @Newman2016-my. Graphics produced in part using @Stawitz2026-kw.  
\
<!--
## Additional Information

The 'PDO ERSST V5' csv time series data were downloaded and manually renamed to PDO_`r RptYr`.csv.  The original file name is pdo.timeseries.ersstv5.csv.

No data are saved for the Council portal because the PDO isn't examined as an annual mean.

-->
