---
title: "North_Pacific_Gyre_Oscillation"
author: "phoebe.woodworth-jefcoats@noaa.gov"
format:
  docx:
    reference-doc: SAFE-Reference-Doc.docx
---

## North Pacific Gyre Oscillation

```{r}
#| include: false
### Load libraries
# Commenting out the plotly, reticulate, python stuff because it is 
# inexplicably not working this year
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
# remotes::install_github("nmfs-fish-tools/nmfspalette")
library(nmfspalette)
# library(plotly)
# library(reticulate)
#devtools::install_github("teunbrand/elementalist")
library(elementalist)
# 
# reticulate::use_miniconda('r-reticulate')
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2025

# Set path to variable: North_Pacific_Gyre_Oscillation
# This is where the data are and where the plots will go
Dir <- here("North_Pacific_Gyre_Oscillation")
# Check if it exists; if not, create it
if (!dir.exists(Dir)) {
  dir.create(Dir)
}
```

```{r}
#| include: false
### Load data
NPGO_full <- read_csv(file = paste(Dir, '/NPGO_', RptYr, '.csv', sep = ""),
                      show_col_types = FALSE)
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# Pull out the values we need for the report 
yr_of_int <- which(NPGO_full$YEAR == RptYr)
prev_yrs <- which(NPGO_full$YEAR < RptYr)
all_yrs <- which(NPGO_full$YEAR <= RptYr)
monthly_min_RptYr <- min(NPGO_full$NPGO[yr_of_int])
monthly_min_PrevYrs <- min(NPGO_full$NPGO[prev_yrs])
monthly_max_RptYr <- max(NPGO_full$NPGO[yr_of_int])
monthly_max_PrevYrs <- max(NPGO_full$NPGO[prev_yrs])

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The highest monthly value was in the report year.')
}

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The lowest monthly value was in the report year.')
}
```

```{r}
#| include: false
# Write csv for dashboard
# Note that dashboard output has its own folder
# Thanks to Emily Conklin for this chunk of code!

# Create a Date field
NPGO_full <- NPGO_full |>
  mutate(Date = paste(NPGO_full$YEAR, NPGO_full$MONTH, '01', sep = "-"))

NPGO_dashboard <- NPGO_full %>%
  mutate(Date_Time = as_datetime(Date), Year = year(Date_Time), 
         Month = month(Date_Time)) %>%
  mutate(Date_Time = Date_Time %>%
           format('%d-%b-%Y %H:%M') %>%
           toupper) %>%
  select(Date_Time, Year, Month, Value = NPGO) %>%
  mutate(Anom = NA, ID = "NPGO", Value_lm = NA, Anom_lm = NA, Units = "North Pacific Gyre Oscillation") |>
  filter(Year <= RptYr) 

write_csv(NPGO_dashboard, file = here("Indicator_Dashboard", "Data", paste('NPGO_Dashboard_Data_', RptYr, '.csv', sep = "")))
```

```{r}
#| include: false
# Borrowing code from the dashboard for this chunk
# so that figures look the same across products
indicator_data <- NPGO_dashboard |>
  filter(Year <= RptYr) %>% 
  mutate(Date_Time=dmy_hm(Date_Time))

# Create color palette for easy reference 
oceans <- nmfs_palette("oceans")(3) # 1 = report_year, 2 = Negative PDO, 3 = previous years
crustacean <- nmfs_palette("crustacean")(4) # 1 = linear trend
coral <- nmfs_palette("coral")(3) # 2 = Positive PDO,  3 = annual average for Rpt Yr
ann_grey <- "#A5AAAF" # annual means; in NMFS branding guide but not in package
waves <- nmfs_palette("waves")(3) # annual means; in NMFS branding guide but not in package
seagrass <- nmfs_palette("seagrass")(3)
pal <- c(oceans[3], coral[2], waves[2], coral[3], crustacean[2])

# # Formatting
# plot_title_font <- list(size = 14)
# plot_height <- 350 #in pixels

indicator_data <- indicator_data |>
  mutate(Units = "North Pacific Gyre Oscillation (NPGO)")

# Identify the current year, to overlay on plot
given_yr <- indicator_data |>
  filter(Year == RptYr)

# x axis ticks and grid
x_grid <- c(seq.POSIXt(min(indicator_data$Date_Time), max(indicator_data$Date_Time), by='10 year'))

# y-axis limits
ts_y_lim <-c(-4,4)

# Plot using ggplot this year (2025)
## Plot using ggplot with legend
# Easier to control the aesthetics and size

# Defining the theme up front so we don't have to repeat it
plotTheme <- theme_minimal() +
  theme(panel.border = element_rect(color='#F1F3F3'),
        panel.grid.major = element_line(color='#F1F3F3'),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        axis.title = element_text(size=14, face="bold"), 
        axis.text = element_text(size=12),
        legend.position = 'bottom', 
        legend.background = element_rect_round(fill = 'gray90', color='transparent', radius = 0.3),
        legend.text = element_text(size=12), legend.key.spacing.x = unit(1.5,'cm'))

# Legend
legendInfo <- as.factor(c(paste0('Mothly index, 1950-', RptYr-1), 
                          paste0('Mothly index, ', RptYr)))

p <- ggplot() +
  geom_hline(yintercept = 0, linewidth=0.25) +
  geom_path(data=indicator_data, aes(Date_Time, Value, color=legendInfo[1])) +
  geom_path(data=given_yr, aes(Date_Time, Value, color=legendInfo[2])) +
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = F) +
  scale_y_continuous(expand = F, limits = ts_y_lim, breaks=(seq(min(ts_y_lim),max(ts_y_lim),by=1))) +
  scale_color_manual(name=NULL, values=pal[c(1,2)]) +
  ylab(indicator_data$Units[1]) +
  xlab('Year') +
  guides(color = guide_legend(ncol = 2)) +
  plotTheme

ggsave(paste0(Dir, '/NPGO_ts_', RptYr, '_ggolot.pdf'), plot=p, width=6, height=5, units='in')

# # The plotly code isn't working this year.  So,
# # going with base R, because that's what I'm most familiar with
# # Fair warning, this coding is tedious.
# 
# # x axis limit
# ts_x_lim <- c(min(indicator_data$Date_Time), max(indicator_data$Date_Time))
# 
# pdf(paste(Dir, '/NPGO_ts_', RptYr, '.pdf', sep = ""), width=9, height=6)
# par(mar=c(6.6, 5.1, 2.1, 2.1))
# plot(indicator_data$Date_Time, indicator_data$Value, type = "l", lwd = 2, 
#      xlim = ts_x_lim, ylim = c(-4, 4), 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[1],
#      xlab = "Year", ylab = "North Pacific Gyre Oscillation (NPGO)",
#      # put a grid behind the plot
#      panel.first = c(abline(v = x_grid, col = "#F1F3F3"),
#                      abline(h = c(seq(-4, 4, 1)), col = "#F1F3F3")),
#      font.lab = 2)
# par(new = TRUE)
# plot(given_yr$Date_Time, given_yr$Value, type = "l", lwd = 2, 
#      xlim = ts_x_lim, ylim = c(-4,4), 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[2],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(c(min(indicator_data$Date_Time), max(indicator_data$Date_Time)), c(0, 0), col = "black", 
#      type = "l", lwd = 1,
#      xlim = ts_x_lim, ylim = c(-4, 4),
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      xlab = " ", ylab = " ")
# axis(1, at = x_grid, label = c(seq(1950, 2020, 10), " "), col = "transparent", col.ticks = "transparent") 
# axis(2, at = c(seq(-4, 4, 1)), las = 1, col = "transparent", col.ticks = "transparent")
# grid.roundrect(x=0.51, y=0.04, width=0.55, height=0.06, 
#                r=unit(0.3, "snpc"), gp=gpar(fill="gray90", col='gray90'), just='centre')
# legend('bottom', 
#        legend=c(paste0('Monthly index, 1950-',RptYr-1), paste0('Monthly index, ',RptYr)), 
#        col=c(pal[1], pal[2]), 
#        lty=1, lwd=1.75, 
#        bg='transparent', 
#        horiz=T, 
#        inset = -0.3, xpd=T, 
#        box.lwd = 0, box.col = 'transparent')
# dev.off()

```

**Rationale:** The North Pacific Gyre Oscillation (NPGO) is related to, but distinct from, the Pacific Decadal Oscillation.  It captures basin-scale processes, making it well suited to capture change across the Hawaiʻi-based longline fishing grounds.    
\
**Status:** The NPGO was negative in `r RptYr`. The index ranged from `r signif(monthly_min_RptYr, 3)` to `r signif(monthly_max_RptYr, 3)` over the course of the year.  This is within the range of values observed previously in the time series.  
\
**Description:** The NPGO is driven by regional and basin-scale changes in wind patterns.  Changed to the wind field driver changes in ocean circulation that affect nutrients and, in turn, phytoplankton.  Because phytoplankton are the foundational food source for marine ecosystems, the NPGO offers insight into potential ecosystem change.  Description paraphrased from Di Lorenzo et al. (2008).  
\
**Timeframe:** Monthly.  
\
**Region/Location:** 180°–110°W; 25°N–62°N.  
\
**Measurement Platform:** Satellite, model.  
\
**Data available at:** <https://www.o3d.org/npgo/data/NPGO.txt>.  
\
Sourced from Di Lorenzo et al. (2008).  
\
Graphics produced in part using Stawitz and Li (2025).  
\
## Additional Information

Data were manually copied and pasted into a .csv.  Each column was formatted to remove scientific notation which was resulting in rounding errors (data type set to 'numeric').

No data are saved for the Council portal because the NPGO isn't examined as an annual mean.
