### North Pacific Gyre Oscillation

```{r}
#| include: false

# This includes all commonly used packages and variables, like RptYr and eliminates the need to change them in every chapter. 
source("_common.R")
```

```{r}
#| include: false
# Set path to variable: North_Pacific_Gyre_Oscillation
# This is where the data are and where the plots will go
Dir <- here("North_Pacific_Gyre_Oscillation")
# Check if it exists; if not, create it
if (!dir.exists(Dir)) {
  dir.create(Dir)
}
```

```{r}
#| include: false
# Install library
library(readr)

# Download NPGO data from website and save for the record.
# Specify the URL
url <- "https://www.o3d.org/npgo/data/NPGO.txt"

# Read the raw lines to find where the data table actually starts
raw_lines <- readLines(url)

# Find lines that don't start with '#' or comment markers (the data lines)
header_lines <- grep("^#", raw_lines)
data_start_line <- max(header_lines)

# Read the data table into a data frame
# The file uses whitespace to separate Year, Month, and NPGO index value
NPGO_full <- read.table(
  url,
  skip = data_start_line,
  col.names = c("YEAR", "MONTH", "NPGO"),
  stringsAsFactors = FALSE
)

# Save a copy if you want. We do save a version for the PIFSC dashboard further down in the code though. 
#write.csv(npgo_data, file = paste(Dir, '/NPGO_', RptYr, '.csv', sep = ""), col.names = F, quote = F)
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# Pull out the values we need for the report 
yr_of_int <- which(NPGO_full$YEAR == RptYr)
prev_yrs <- which(NPGO_full$YEAR < RptYr)
all_yrs <- which(NPGO_full$YEAR <= RptYr)
monthly_min_RptYr <- min(NPGO_full$NPGO[yr_of_int])
monthly_min_PrevYrs <- min(NPGO_full$NPGO[prev_yrs])
monthly_max_RptYr <- max(NPGO_full$NPGO[yr_of_int])
monthly_max_PrevYrs <- max(NPGO_full$NPGO[prev_yrs])

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The highest monthly value was in the report year.')
}

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The lowest monthly value was in the report year.')
}
```

```{r}
#| include: false
# Write csv for dashboard
# Note that dashboard output has its own folder
# Thanks to Emily Conklin for this chunk of code!

# Create a Date field
NPGO_full <- NPGO_full |>
  mutate(Date = paste(NPGO_full$YEAR, NPGO_full$MONTH, '01', sep = "-"))

NPGO_dashboard <- NPGO_full %>%
  mutate(Date_Time = as_datetime(Date), Year = year(Date_Time), 
         Month = month(Date_Time)) %>%
  mutate(Date_Time = Date_Time %>%
           format('%d-%b-%Y %H:%M') %>%
           toupper) %>%
  select(Date_Time, Year, Month, Value = NPGO) %>%
  mutate(Anom = NA, ID = "NPGO", Value_lm = NA, Anom_lm = NA, Units = "North Pacific Gyre Oscillation") |>
  filter(Year <= RptYr) 

#write_csv(NPGO_dashboard, file = here("Indicator_Dashboard", "Data", paste('NPGO_Dashboard_Data_', RptYr, '.csv', sep = "")))
```

```{r}
#| include: false
# Borrowing code from the dashboard for this chunk
# so that figures look the same across products
indicator_data <- NPGO_dashboard |>
  filter(Year <= RptYr) %>% 
  mutate(Date_Time=dmy_hm(Date_Time))

# Create color palette for easy reference 
oceans <- nmfs_palette("oceans")(3) # 1 = report_year, 2 = Negative PDO, 3 = previous years
crustacean <- nmfs_palette("crustacean")(4) # 1 = linear trend
coral <- nmfs_palette("coral")(3) # 2 = Positive PDO,  3 = annual average for Rpt Yr
ann_grey <- "#A5AAAF" # annual means; in NMFS branding guide but not in package
waves <- nmfs_palette("waves")(3) # annual means; in NMFS branding guide but not in package
seagrass <- nmfs_palette("seagrass")(3)
pal <- c(oceans[3], coral[2], waves[2], coral[3], crustacean[2])

# Formatting
indicator_data <- indicator_data |>
  mutate(Units = "North Pacific Gyre Oscillation (NPGO)")

# Identify the current year, to overlay on plot
given_yr <- indicator_data |>
  filter(Year == RptYr)

# x axis ticks and grid
x_grid <- c(seq.POSIXt(min(indicator_data$Date_Time), max(indicator_data$Date_Time), by='10 year'))

# y-axis limits
ts_y_lim <-c(-4,4)

# Plot using ggplot with legend
# Easier to control the aesthetics and size

# Defining the theme up front so we don't have to repeat it
plotTheme <- theme_minimal() +
  theme(panel.border = element_rect(color='#F1F3F3'),
        panel.grid.major = element_line(color='#F1F3F3'),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        axis.title = element_text(size=14, face="bold"), 
        axis.text = element_text(size=12),
        legend.position = 'bottom', 
        legend.background = element_rect_round(fill = 'gray90', color='transparent', radius = 0.3),
        legend.text = element_text(size=12), legend.key.spacing.x = unit(1.5,'cm'))

# Legend
legendInfo <- as.factor(c(paste0('Monthly index, 1950-', RptYr-1), 
                          paste0('Monthly index, ', RptYr)))

p <- ggplot() +
  geom_hline(yintercept = 0, linewidth=0.25) +
  geom_path(data=indicator_data, aes(Date_Time, Value, color=legendInfo[1])) +
  geom_path(data=given_yr, aes(Date_Time, Value, color=legendInfo[2])) +
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = F) +
  scale_y_continuous(expand = F, limits = ts_y_lim, breaks=(seq(min(ts_y_lim),max(ts_y_lim),by=1))) +
  scale_color_manual(name=NULL, values=pal[c(1,2)]) +
  ylab(indicator_data$Units[1]) +
  xlab('Year') +
  guides(color = guide_legend(ncol = 2)) +
  plotTheme

#ggsave(paste0(Dir, '/NPGO_ts_', RptYr, '_ggolot.pdf'), plot=p, width=6, height=5, units='in')
#ggsave(paste0(Dir, '/NPGO_ts_', RptYr, '_ggolot.png'), plot=p, width=8, height=5, units='in')

```

**Rationale:** The North Pacific Gyre Oscillation (NPGO) is related to, but distinct from, the Pacific Decadal Oscillation.  It captures basin-scale processes, making it well suited to capture change across the Hawaiʻi-based longline fishing grounds.    
\
**Status:** The NPGO was negative in `r RptYr`. The index ranged from `r signif(monthly_min_RptYr, 3)` to `r signif(monthly_max_RptYr, 3)` over the course of the year. This is within the range of values observed previously in the time series.  
\
**Description:** The NPGO is driven by regional and basin-scale changes in wind patterns. Changed to the wind field driver changes in ocean circulation that affect nutrients and, in turn, phytoplankton. Because phytoplankton are the foundational food source for marine ecosystems, the NPGO offers insight into potential ecosystem change. Description paraphrased from @Di_Lorenzo2008-sb.  
\
**Timeframe:** Monthly.  
\
**Region/Location:** 180°–110°W; 25°N–62°N.  
\
**Measurement Platform:** Satellite, model.  
\
**Data available at:** <https://www.o3d.org/npgo/data/NPGO.txt>.  
\
**Sourced from:** @Di_Lorenzo2008-sb. Graphics produced in part using @Stawitz2026-kw.  
\
<!--
## Additional Information

Data were manually copied and pasted into a .csv.  Each column was formatted to remove scientific notation which was resulting in rounding errors (data type set to 'numeric').

No data are saved for the Council portal because the NPGO isn't examined as an annual mean. 
-->
