---
title: "Sea_Surface_Temperature"
author: "phoebe.woodworth-jefcoats@noaa.gov"
format:
  docx:
    reference-doc: SAFE-Reference-Doc.docx
---

## Sea Surface Temperature

```{r}
#| include: false
### Load libraries
# Commenting out the plotly, reticulate, python stuff because it is 
# inexplicably not working this year
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
# remotes::install_github("nmfs-fish-tools/nmfspalette")
library(nmfspalette)
# library(plotly)
# library(reticulate)
library(grid)
#devtools::install_github("teunbrand/elementalist")
library(elementalist)
# 
# reticulate::use_miniconda('r-reticulate')
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2025

# Set path to variable: Sea_Surface_Temperature
# This is where the data are and where the plots will go
Dir <- here("Sea_Surface_Temperature")
# Check if it exists; if not, create it
if (!dir.exists(Dir)) {
  dir.create(Dir)
}
```

```{r}
#| include: false
### Load data
# Monthly SST
# Generated in ~5-yr batches due to memory limitations
# There's gotta be a way to append a year that's way better than this.  Implement this next year.
## Johanna made an attempt below in 2025
# read in all the time series file names
SST_files <- list.files(path=Dir, pattern='sst_ts_', full.names = T)
# Read in all the files
SST_full <- lapply(SST_files, read.csv, skip=8, header=F) %>% 
  bind_rows() %>% 
  rename(Date_Time=V1, SST_degC=V2) %>% 
  mutate(Date_Time=dmy_hm(Date_Time)) %>% 
  arrange(Date_Time) # note that this method doesn't read the files in chronological order, so we sort them here to make sure everything lines up properly down the line

```

```{r}
#| include: false
# Remove seasonal means to calculate anomalies
SST_climo <- matrix(NA, nrow = nrow(SST_full), ncol = 1)
for (m in seq(1,12,1)) {
  mo_of_int <- which(month(SST_full$Date_Time) == m)
  SST_climo[mo_of_int,1] <- mean(SST_full$SST_degC[mo_of_int])
}

SST_anom_ts <- SST_full$SST_degC - SST_climo
```

```{r}
#| include: false
### Linear fit 
# Note that this assumes that observations are equally spaced in time, which they're  not
n_obs <- seq(1, nrow(SST_full), 1)
SST_lm <- lm(SST_full$SST_degC ~ n_obs)
SST_anom_lm <- lm(SST_anom_ts ~ n_obs)

# Change over time
delta_SST_lm <- SST_lm$fitted.values[length(n_obs)] - SST_lm$fitted.values[1]

delta_SST_anom_lm <- SST_anom_lm$fitted.values[length(n_obs)] - 
  SST_anom_lm$fitted.values[1]
```

```{r}
#| include: false
### Annual values
SST_full <- SST_full %>% 
    mutate(Year=year(Date_Time), 
           Anom=SST_anom_ts)

ann_SST <- SST_full %>%
  group_by(Year) %>%
  summarise(SST_degC = mean(SST_degC, na.rm = TRUE))
ann_anom <- SST_full %>%
  group_by(Year) %>%
  summarise(Anom = mean(Anom, na.rm = TRUE))
ann_mean_RptYr <- ann_SST$SST_degC[which(ann_SST$Year == RptYr)]
ann_anom_RptYr <- ann_anom$Anom[which(ann_SST$Year == RptYr)]
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# This section includes some error checks to prompt fixing the text
if (any(summary(SST_lm)$coefficients[,4] > 0.05)) {
  print('The linear fit to monthly values is not signficant. Remove text related to the linear trend.')
}

if (any(summary(SST_anom_lm)$coefficients[,4] > 0.05)) {
  print('The linear fit to anomaly values is not signficant. Remove text related to the linear trend.')
}

# Pull out the MONTHLY values we need for the report & plots
yr_of_int <- which(SST_full$Year == RptYr)
prev_yrs <- which(SST_full$Year < RptYr)
all_yrs <- which(SST_full$Year <= RptYr)
monthly_max_RptYr <- max(SST_full$SST_degC[yr_of_int])
monthly_min_RptYr <- min(SST_full$SST_degC[yr_of_int])
monthly_max_PrevYrs <- max(SST_full$SST_degC[prev_yrs])
monthly_min_PrevYrs <- min(SST_full$SST_degC[prev_yrs])

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The greatest monthly value was during the report year.  Revise text to reflect this.')
}

if (monthly_min_RptYr < monthly_min_PrevYrs) {
  print('The lowest monthly value was during the report year.  Revise text to reflect this.')
}
```

```{r}
#| include: false
# Write csv for portal
# Note that output csvs go in their own folder
# Rounding to avoid too many significant figures.
# Thanks to Johanna Wren for the code! https://github.com/pwoodworth-jefcoats/climate-indicators/issues/29
ann_SST <- rename(ann_SST, `Degrees C` = SST_degC)
ann_SST$`Degrees C` <- format(round(ann_SST$`Degrees C`, digits=2), nsmall = 2)
write_csv(ann_SST, file = paste(here(), '/PelagicClimate_', RptYr, '/SST_', RptYr, '.csv', sep = ""))

ann_anom <- rename(ann_anom, `Degrees C` = Anom)
ann_anom$`Degrees C` <- format(round(ann_anom$`Degrees C`, digits=2), nsmall = 2)
write_csv(ann_anom, file = paste(here(), '/PelagicClimate_', RptYr, '/SSTanomaly_', RptYr, '.csv', sep = ""))
```

```{r}
#| include: false
# Write csv for dashboard
# Note that dashboard output has its own folder

# Add columns for month, variable ID, and units
SST_dashboard <- SST_full %>% 
  mutate(Month=month(Date_Time), 
         ID='SST', 
         Value_lm=SST_lm$fitted.values, 
         Anom_lm=SST_anom_lm$fitted.values, 
         Units='Deg C', 
         Value=SST_degC) %>% 
  dplyr::select(Date_Time, Year, Month, Value, Anom, ID, Value_lm, Anom_lm, Units)

write.csv(SST_dashboard, file = here("Indicator_Dashboard", "Data", paste0('SST_Dashboard_Data_', RptYr, '.csv')), row.names =F)
```

```{r}
#| include: false
# Borrowing code from the dashboard for this chunk
# so that figures look the same across products
indicator_data <- SST_dashboard |>
  filter(Year <= RptYr)

# Create color palette for easy reference 
oceans <- nmfs_palette("oceans")(3) # 1 = report_year, 3 = previous years
crustacean <- nmfs_palette("crustacean")(4) # 1 = linear trend
coral <- nmfs_palette("coral")(3) # 3 = annual average for Rpt Yr
ann_grey <- "#A5AAAF" # annual means; in NMFS branding guide as dove gray tint 1
waves <- nmfs_palette("waves")(3) # annual means; in NMFS branding guide but not in package
seagrass <- nmfs_palette("seagrass")(3)
pal <- c(oceans[3], coral[2], waves[2], coral[3], crustacean[2])

# # Formatting
# plot_title_font <- list(size = 14)
# plot_height <- 350 #in pixels

# Calculate annual means 
ann_vals <- indicator_data |>
  group_by(Year) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Identify the current year, to overlay on plot
given_yr <- indicator_data |>
  filter(Year == RptYr)

given_yr_ann <- bind_cols(Value = rep(ann_vals$Value[dim(ann_vals)[1]]),
                          Date_Time = given_yr$Date_Time,
                          Anom = as.numeric(ann_anom$`Degrees C`[which(ann_anom$Year == RptYr)]))

# Commented out the plotly code because it's not working this year.  So,
# going with base R, because that's what I'm most familiar with
# Fair warning, this coding is tedious.
# Starting in 2025 using ggplot code instead becuase easier to control aesthetics and size and it has legend built in

# x axis ticks and grid
# Made a sequence instead so we don't have to manually update it every five years
x_grid <- c(seq.POSIXt(min(indicator_data$Date_Time), max(indicator_data$Date_Time), by='5 year'))

# y-axis limits
ts_y_lim <-c(17, 25)
anom_y_lim <- c(-1.5, 1.5)

# Defining the theme up front so we don't have to repeat it
plotTheme <- theme_minimal() +
  theme(panel.border = element_rect(color='#F1F3F3'),
        panel.grid.major = element_line(color='#F1F3F3'),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        axis.title = element_text(size=14, face="bold"), 
        axis.text = element_text(size=12),
        legend.position = 'bottom', 
        legend.background = element_rect_round(fill = 'gray90', color='transparent', radius = 0.3),
        legend.text = element_text(size=12),legend.key.spacing.y = unit(-0.13,'cm'), legend.key.spacing.x = unit(1.5,'cm'))

# Plot using ggplot
legendInfo <- as.factor(c(paste0('Mothly average, 1985-', RptYr-1), 
                          paste0('Annual average, 1985-', RptYr-1), 
                          paste0('Mothly average, ', RptYr), 
                          paste0('Annual average, ', RptYr),
                          paste0('Long-term Trend')))
levels(legendInfo) <- levels(legendInfo)[c(4,5,1,2,3)] # To arrange in preferred order for the legend, otherwise it's alphabetical

p <- ggplot() +
  geom_path(data=indicator_data, aes(Date_Time, Value, color=legendInfo[2])) +
  geom_path(data=ann_vals, aes(dmy_hm(paste0('01-01-',Year,' 00:00')), Value, color=legendInfo[5])) +
  geom_path(data=indicator_data, aes(Date_Time, Value_lm, color=legendInfo[3])) +
  geom_path(data=given_yr, aes(Date_Time, Value, color=legendInfo[4])) +
  geom_path(data=given_yr_ann, aes(Date_Time, Value, color=legendInfo[1])) +
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = F) +
  scale_y_continuous(expand = F, limits = ts_y_lim, breaks=(seq(min(ts_y_lim),max(ts_y_lim),by=1))) +
  scale_color_manual(name=NULL, values=pal[c(1,3,5,2,4)]) +
  ylab('Sea Surface Temperature (°C)') +
  xlab('Year') +
  guides(color = guide_legend(ncol = 2)) +
  plotTheme

ggsave(paste0(Dir, '/SST_ts_', RptYr, '_ggolot.pdf'), plot=p, width=6, height=5, units='in')
ggsave(paste0(Dir, '/SST_ts_', RptYr, '_ggolot.png'), plot=p, width=6, height=5, units='in')


### Anomaly plot
# Calculate annual means
ann_vals <- indicator_data |>
  group_by(Year) |>
  summarise(Anom = mean(Anom, na.rm = TRUE)) 

p_ann <- ggplot() +
  geom_hline(yintercept = 0, linewidth=0.25) +
  geom_path(data=indicator_data, aes(Date_Time, Anom, color=legendInfo[2])) +
  geom_path(data=ann_vals, aes(dmy_hm(paste0('01-01-',Year,' 00:00')), Anom, color=legendInfo[5])) +
  geom_path(data=indicator_data, aes(Date_Time, Anom_lm, color=legendInfo[3])) +
  geom_path(data=given_yr, aes(Date_Time, Anom, color=legendInfo[4])) +
  geom_path(data=given_yr_ann, aes(Date_Time, Anom, color=legendInfo[1])) +
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = F) +
  scale_y_continuous(expand = F, limits = anom_y_lim, breaks=(seq(min(anom_y_lim),max(anom_y_lim),by=0.5))) +
  scale_color_manual(name=NULL, values=pal[c(1,3,5,2,4)]) +
  ylab('Sea Surface Temperature Anomaly (°C)') +
  xlab('Year') +
  guides(color = guide_legend(ncol = 2)) +
  plotTheme

ggsave(paste0(Dir, '/SST_anom_ts_', RptYr, '_ggolot.pdf'), plot=p_ann, width=6, height=5, units='in')
ggsave(paste0(Dir, '/SST_anom_ts_', RptYr, '_ggolot.png'), plot=p_ann, width=6, height=5, units='in')


# # Plot using base plot. Retired in 2025 
# # x axis limit
# ts_x_lim <- c(min(indicator_data$Date_Time), max(indicator_data$Date_Time))

# pdf(paste(Dir, '/SST_ts_', RptYr, '.pdf', sep = ""), width=9, height=6)
# par(mar=c(8.6, 5.1, 2.1, 2.1))
# plot(indicator_data$Date_Time, indicator_data$Value, type = "l", lwd = 2, 
#      xlim = ts_x_lim, ylim = ts_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[1],
#      xlab = "Year", ylab = "Sea Surface Temperature (°C)",
#      # put a grid behind the plot
#      panel.first = c(abline(v = x_grid, col = "#F1F3F3"),
#                      abline(h = c(seq(17, 25, 1)), col = "#F1F3F3")),
#      font.lab = 2, cex.lab=1.2)
# par(new = TRUE)
# plot(ymd_hm(paste(ann_vals$Year, '0601 00:00', sep = "")), ann_vals$Value, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = ts_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[3],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(indicator_data$Date_Time, indicator_data$Value_lm, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = ts_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[5],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(given_yr$Date_Time, given_yr$Value, type = "l", lwd = 2, 
#      xlim = ts_x_lim, ylim = ts_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[2],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(given_yr_ann$Date_Time, given_yr_ann$Value, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = ts_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[4],
#      xlab = " ", ylab = " ")
# axis(1, at = x_grid, label = c(seq(1985, 2025, 5), " "), col = "transparent", col.ticks = "transparent") 
# axis(2, at = c(seq(17, 25, 1)), las = 1, col = "transparent", col.ticks = "transparent")
# grid.roundrect(x=0.51, y=0.075, width=0.6, height=0.12, 
#                r=unit(0.3, "snpc"), gp=gpar(fill="gray90", col='gray90'), just='centre')
# legend('bottom', 
#        legend=c(paste0('Monthly average, 1985-',RptYr-1), paste0('Annual average, 1985-',RptYr-1), 'Long-term Trend', paste0('Monthly average, ',RptYr), paste0('Annual average, ',RptYr)), 
#        col=c(pal[1], pal[3], pal[5], pal[2], pal[4]), 
#        lty=1, lwd=1.75, 
#        bg='transparent', 
#        horiz=F, ncol=2,
#        inset = -0.435, xpd=T, 
#        box.lwd = 0, box.col = 'transparent')
# dev.off()


# p1 <- plot_ly(indicator_data, x = dmy_hm(indicator_data$Date_Time), y = ~Value,
#               type = "scatter", mode = "lines", line = list(color = pal[1]),
#               name = ~ID[1], height = plot_height) |>
#   add_trace(indicator_data, x = dmy_hm(indicator_data$Date_Time), y = ~Value_lm,
#             type = "scatter", mode = "lines", line = list(color = pal[5]),
#             name = "Long-term Trend") |>
#   add_trace(ann_vals, x = ymd_hm(paste(ann_vals$Year, '0601 00:00', sep = "")), y = ann_vals$Value,
#             type = "scatter", mode = "lines", line = list(color = pal[3]),
#             name = "Annual Mean") |>
#   add_trace(given_yr, x = dmy_hm(given_yr$Date_Time), y = given_yr$Value,
#             type = "scatter", mode = "lines", line = list(color = pal[2]),
#             name = ~ID[1], height = plot_height) |>
#   add_trace(given_yr_ann, x = dmy_hm(given_yr_ann$Date_Time), y = given_yr_ann$Value,
#             type = "scatter", mode = "lines", line = list(color = pal[4]),
#             name = ~ID[1], height = plot_height)
# 
# #apply same layout parameters for all plots
# 
# #custom x axis (min, every decade, report year)
# all_years <- unique(indicator_data$Year)
# first_date <- as.character(parse_date_time(indicator_data$Date_Time[1], orders = "d-b-Y H:M"))
# date_axis <- c(first_date, 
#               all_years[which(all_years %% 10 == 0)], 
#               RptYr)
# 
# p1 <- p1 |> layout(title = list(text = "Indicator Time Series", x = 0.01, font = plot_title_font), #add title
#                    xaxis = list(type = "date", tickformat = "%Y", tickmode = "array", tickvals = date_axis, tickangle = 90),
#                    #xaxis = list(tick0 = min(indicator_data$Date_Time), dtick = "M24"),
#                    yaxis = list(title = indicator_data$Units[1], hoverformat = '.3f', range = list(17, 25), tickvals = list(17, 18, 19, 20, 21, 22, 23, 24, 25)), #add units and round values in hover display; not sure of a better way to set axis limits and ticks...
#                    paper_bgcolor = 'transparent', plot_bgcolor = 'transparent', #transparent background
#                    hovermode = "x unified", #show data for all traces simultaneously
#                    hoverlabel = list(namelength = -1)) #don't cutoff hoverinfo due to length
#   
# # return plot 
# save_image(p1, paste(Dir, '/SST_ts_', RptYr, '.pdf', sep = ""))
# 

# ### Anomaly plot
# # Calculate annual means
# ann_vals <- indicator_data |>
#   group_by(Year) |>
#   summarise(Anom = mean(Anom, na.rm = TRUE)) 

# pdf(paste(Dir, '/SST_anom_ts_', RptYr, '.pdf', sep = ""), width=9, height=6)
# par(mar=c(8.6, 5.1, 2.1, 2.1))
# plot(indicator_data$Date_Time, indicator_data$Anom, type = "l", lwd = 2, 
#      xlim = ts_x_lim, ylim = anom_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[1],
#      xlab = "Year", ylab = "Sea Surface Temperature Anomaly (°C)",
#      # put a grid behind the plot
#      panel.first = c(abline(v = x_grid, col = "#F1F3F3"),
#                      abline(h = c(seq(-1.5, 1.5, 0.5)), col = "#F1F3F3"),
#                      abline(h = 0, col='black')),
#      font.lab = 2, cex.lab=1.2)
# par(new = TRUE)
# plot(ymd_hm(paste(ann_vals$Year, '0601 00:00', sep = "")), ann_vals$Anom, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = anom_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[3],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(indicator_data$Date_Time, indicator_data$Anom_lm, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = anom_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[5],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(given_yr$Date_Time, given_yr$Anom, type = "l", lwd = 2, 
#      xlim = ts_x_lim, ylim = anom_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[2],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(given_yr_ann$Date_Time, given_yr_ann$Anom, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = anom_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[4],
#      xlab = " ", ylab = " ")
# axis(1, at = x_grid, label = c(seq(1985, 2020, 5), 2024, " "), col = "transparent", col.ticks = "transparent") 
# axis(2, at = c(seq(-1.5, 1.5, 0.5)), las = 1, col = "transparent", col.ticks = "transparent")
# grid.roundrect(x=0.51, y=0.075, width=0.6, height=0.12, 
#                r=unit(0.3, "snpc"), gp=gpar(fill="gray90", col='gray90'), just='centre')
# legend('bottom', 
#        legend=c(paste0('Monthly average, 1985-',RptYr-1), paste0('Annual average, 1985-',RptYr-1), 'Long-term Trend', paste0('Monthly average, ',RptYr), paste0('Annual average, ',RptYr)), 
#        col=c(pal[1], pal[3], pal[5], pal[2], pal[4]), 
#        lty=1, lwd=1.75, 
#        bg='transparent', 
#        horiz=F, ncol=2,
#        inset = -0.435, xpd=T, 
#        box.lwd = 0, box.col = 'transparent')
# dev.off()

# p2 <- plot_ly(indicator_data, x = dmy_hm(indicator_data$Date_Time), y = ~Anom, height = plot_height,
#                 type = "scatter", mode = "lines", line = list(color = pal[1]),
#                 name = "Monthly Anomaly") |>
#     add_trace(indicator_data, x = dmy_hm(indicator_data$Date_Time), y = ~Anom_lm,
#               type = "scatter", mode = "lines", line = list(color = pal[5]),
#               name = "Long-term Trend") |>
#     add_trace(ann_vals, x = ymd_hm(paste(ann_vals$Year, '0601 00:00', sep = "")), y = ann_vals$Anom,
#               type = "scatter", mode = "lines", line = list(color = pal[3]),
#               name = "Annual Mean") |>
#   add_trace(given_yr, x = dmy_hm(given_yr$Date_Time), y = given_yr$Anom,
#             type = "scatter", mode = "lines", line = list(color = pal[2]),
#             name = ~ID[1], height = plot_height) |>
#   add_trace(given_yr_ann, x = dmy_hm(given_yr_ann$Date_Time), y = given_yr_ann$Anom,
#             type = "scatter", mode = "lines", line = list(color = pal[4]),
#             name = ~ID[1], height = plot_height) |>
#     layout(xaxis = list(type = "date", tickformat = "%Y", tickmode = "array", tickvals = date_axis, tickangle = 90),
#            yaxis = list(title = indicator_data$Units[1], hoverformat = '.3f', range = list(-1.5, 1.5), tickvals = list(-1.5, -1, -0.5, 0, 0.5, 1.0, 1.5)), #add units and round values in hover display
#            title = list(text = "Anomaly Time Series", x = 0.01, font = plot_title_font), #add title 
#            paper_bgcolor = 'transparent', plot_bgcolor = 'transparent', #transparent background
#            hovermode = "x unified", #show data for all traces simultaneously
#            hoverlabel = list(namelength = -1)) #don't cutoff hoverinfo due to length
# 
# # return plot
# save_image(p2, paste(Dir, '/SST_anom_ts_', RptYr, '.pdf', sep = ""))
```

**Rationale:** Sea surface temperature is one of the most directly observable existing measures for tracking increasing ocean temperatures. SST varies in response to natural climate cycles such as the El Niño – Southern Oscillation (ENSO) and is rising as a result of anthropogenic climate change. Both short-term variability and long-term trends in SST impact the marine ecosystem. Understanding the mechanisms through which organisms are impacted and the time scales of these impacts is an area of active research.  
\
**Status:** Annual mean SST was `r signif(ann_mean_RptYr,3)` ºC in `r RptYr`. Over the period of record, SST across the longline fishing grounds has increased by `r signif(delta_SST_lm, 1)` ºC and the monthly SST anomaly increased by `r signif(delta_SST_anom_lm, 1)` ºC, both at a rate of roughly 0.025 ºC yr^-^^1^. Monthly SST values in `r RptYr` ranged from `r signif(monthly_min_RptYr, 3)`–`r signif(monthly_max_RptYr, 3)` ºC, within the range of temperatures experienced over the past several decades (`r signif(monthly_min_PrevYrs, 3)`–`r signif(monthly_max_PrevYrs, 3)` ºC). Overall, SST was at or above the long-term average across the Hawaiʻi longline region in `r RptYr`.   
\
**Description:** Satellite remotely sensed monthly sea surface temperature (SST) is averaged across the Hawaiʻi-based longline fishing grounds (15° – 45°N, 180° – 120°W). A time series of monthly mean SST averaged over the Hawaiʻi longline region is presented. Additionally, spatial climatologies and anomalies are shown. CoralTemp data are used to calculate this indicator.  
\
**Timeframe:** Monthly.  
\
**Region/Location:** Hawaiʻi longline region: 15° – 45°N, 180° – 120°W.  
\
**Measurement Platform:** Satellite.  
\
**Data available at:** <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/CRW_sst_v3_1_monthly>, <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/CRW_sst_v3_1_1985-2009-clim>, and <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/CRW_sst_v3_1_2024-clim>.   
\
**Sourced from:** NOAA OceanWatch (2025a). Graphics produced in part using Stawitz and Li (2025).  
\  

## Additional Information
SST data are geographically subset and spatially averaged using the pyFerret script `SSTaccess.jnl` which can be found in the [Sea_Surface_Temperature folder](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Sea_Surface_Temperature).  You could do this step in R, but I use [PyFerret](https://ferret.pmel.noaa.gov/Ferret/) because it's freely available software developed by NOAA specifically for working with large gridded datasets.  
\
A plot of the residuals from the linear model showed that they were evenly distributed, although were more positive (~4.0 max) than negative (~-3.0 min). The residuals for the anomaly model were also fairly evenly distributed in terms of values, however they did appear to exhibit some periodicity.  
\
To prepare the spatial data for mapping, you'll need to run `map_data_for_dashboard.R`, which can be found in the [PreProcessing](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Indicator_Dashboard/PreProcessing) folder of the [Indicator_Dashboard](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Indicator_Dashboard) folder in this repository.  

```{r}
#| include: FALSE
# # After running the chunks above and prepping the map data, you can uncomment and run this chunk.
#
# # Load basemap data
# land_proj <- readRDS(here('Indicator_Dashboard', 'Data', 'rnatearth_land.RData'))
# coast_proj <- readRDS(here('Indicator_Dashboard', 'Data', 'rnatearth_coast.RData'))
#
# Load data
maps <- read.csv(paste(Dir, '/SST_map_data_', RptYr, '.csv', sep = ""))

# Filter to the given year and the anomaly
maps_RptYr <- maps |> filter(ID == "SST")
maps_anom <- maps |> filter(ID == "SST_anom")

# Adding a column that assignes line types to contours so we can make a legend
maps_anom$ltyF <- factor(ifelse(maps_anom$layer > 0.5, 'Warmer', ifelse(maps_anom$layer < -0.5, 'Cooler', 'No Change')), levels=c('Cooler', 'Warmer', 'No Change'))

# Map elements and aesthetics
waves <- nmfs_palette("waves")(3)
pal <- rev(waves)
ll_rect_color <- "white" #outline for ll fishing grounds box
fill_scale <- scale_fill_gradientn(name = "SST", colors = pal, limits = c(0, 31))

# Create map
p_map <- ggplot() +
 geom_raster(data = maps_RptYr, mapping = aes(x = x_disp, y = y, fill = layer)) + #data as raster
 annotation_borders('world2', fill='#A5AAAF', colour='gray40', linewidth=0.2) + #add land to the map
 #geom_rect(aes(xmin = 0, xmax = 60, ymin = 15, ymax = 45), color = ll_rect_color, fill = NA, linewidth = 0.5)  + #ll fishing grounds box
 #annotate("text", x = 30, y = 47, label = "Longline fishing grounds", size = 3.2, color = ll_rect_color) +
 fill_scale + #indicator-dependent color and scale from above
 coord_quickmap(expand = F, xlim=c(120,260), ylim=c(0,55)) + #don't expand past x/y limits
 scale_x_continuous(breaks=seq(120,260,30), labels = paste0(c(seq(120,180,30), seq(150,100,-30)), c(rep('°E',2),"",rep('°W',2)))) + # adding degrees to the breaks
 scale_y_continuous(breaks=seq(0,55,15), labels = paste0(seq(0,55,15), c('',rep('°N',3)))) + # adding degrees to the breaks
 xlab("") + ylab("") +
 theme_bw() +
 theme(panel.grid = element_line(color = "black", linewidth = 0.1),
      legend.background = element_rect(fill = 'transparent'), legend.key.height = unit(1, "null"),
      legend.key.width = unit(0.5, 'cm'),  #scales color bar to height of the plot
      legend.margin = margin(0, 0, 0, 0),  #fixing margins on color bar
      legend.text = element_text(size = 10),  #text size for color bar
      plot.title = element_text(hjust = 0.5), 
      plot.margin=unit(c(0,0,0.02,0), "null")) + #centering the plot title. Default is left align
 guides(fill = guide_colorbar(title.position='right', title="Temperature (°C)")) + #rotates the color bar text and centers it
 labs(title = paste(RptYr, 'average (shaded) and anomaly (contours)'))  #add figure title

# Add anomaly
p_map <- p_map +
  geom_contour(data = maps_anom, aes(x = x_disp, y = y, z = layer, linetype=ltyF, linewidth=ltyF),
               breaks = seq(floor(min(maps_anom$layer)), ceiling(max(maps_anom$layer)), 1), color='black') +
  geom_rect(aes(xmin = 180, xmax = 240, ymin = 15, ymax = 45), color = ll_rect_color, fill = NA, linewidth = 0.5)  + #ll fishing grounds box
  annotate("text", x = 210, y = 48, label = "Longline fishing grounds", size = 3.7, color = ll_rect_color) +
  scale_linetype_manual(values = c('dashed','solid','solid'), name='Difference from 1985-2009 average in 1°C intervals') +  #adding legend to bottom of plot. Keep name the same for legends you want to combine (here it's linetype and linewidth)
  scale_linewidth_manual(values=c(0.25,0.25,1), name='Difference from 1985-2009 average in 1°C intervals') +
  guides(fill = guide_colorbar(title.position='right', title="Temperature (°C)",
                               title.theme=element_text(angle=270, hjust=0.5, vjust=0.5)),
         linetype=guide_legend(position = 'bottom', 
                               title.position='top', 
                               title.theme=element_text(hjust=0.5, vjust=1)))  #this bit of code gives different aesthetics to the colorbar on the left and legend on the bottom.

ggsave(filename = paste0(Dir, '/SST_map_', RptYr, '.pdf'), plot=p_map, width = 7, height = 3.8)
ggsave(filename = paste0(Dir, '/SST_map_', RptYr, '.png'), plot=p_map, width = 7, height = 3.8)


# pdf(paste(Dir, '/SST_map_', RptYr, '.pdf', sep = ""), width = 9, height=5)
# print(p)
# dev.off()

#  ### This is Emily's old code. Not sure if better to keep separate or to combine. I only combined to get the bottom legend. 
#  geom_contour(data = maps_anom, mapping = aes(x = x, y = y, z = layer), 
#               breaks = c(seq(floor(min(maps_anom$layer)), -1, 1)),
#               color = "black", linetype = 3) # Dotted negative contours
# p <- p + 
#  geom_contour(data = maps_anom, mapping = aes(x = x, y = y, z = layer), breaks = c(0),
#               color = "black", lwd = 1) # heavy zero line
# p <- p + 
#  geom_contour(data = maps_anom, mapping = aes(x = x, y = y, z = layer), 
#               breaks = c(seq(1, ceiling(max(maps_anom$layer)), 1)),
#               color = "black", lwd = 0.5) # solid positive contours
# 
# p
# # 
```
