### Sea Surface Temperature

```{r}
#| message: false
#| warning: false
#| include: false

# This includes all commonly used packages and variables, like RptYr and eliminates the need to change them in every chapter. 
source("_common.R")
```

```{r}
#| include: false

# Set path to variable: Sea_Surface_Temperature
# This is where the data are and where the plots will go
Dir <- here("Sea_Surface_Temperature")
# Check if it exists; if not, create it
if (!dir.exists(Dir)) {
  dir.create(Dir)
}
```

```{r}
#| include: false
### Load data
# Monthly SST
# Generated in ~5-yr batches due to memory limitations
# There's gotta be a way to append a year that's way better than this.  Implement this next year.
## Johanna made an attempt below in 2025
# read in all the time series file names
SST_files <- list.files(path=Dir, pattern='sst_ts_', full.names = T)
# Read in all the files
SST_full <- lapply(SST_files, read.csv, skip=8, header=F) %>% 
  bind_rows() %>% 
  rename(Date_Time=V1, SST_degC=V2) %>% 
  mutate(Date_Time=dmy_hm(Date_Time)) %>% 
  arrange(Date_Time) # note that this method doesn't read the files in chronological order, so we sort them here to make sure everything lines up properly down the line

```

```{r}
#| include: false
# Remove seasonal means to calculate anomalies
SST_climo <- matrix(NA, nrow = nrow(SST_full), ncol = 1)
for (m in seq(1,12,1)) {
  mo_of_int <- which(month(SST_full$Date_Time) == m)
  SST_climo[mo_of_int,1] <- mean(SST_full$SST_degC[mo_of_int])
}

SST_anom_ts <- SST_full$SST_degC - SST_climo
```

```{r}
#| include: false
### Linear fit 
# Note that this assumes that observations are equally spaced in time, which they're not
n_obs <- seq(1, nrow(SST_full), 1)
SST_lm <- lm(SST_full$SST_degC ~ n_obs)
SST_anom_lm <- lm(SST_anom_ts ~ n_obs)

# Change over time
delta_SST_lm <- SST_lm$fitted.values[length(n_obs)] - SST_lm$fitted.values[1]

delta_SST_anom_lm <- SST_anom_lm$fitted.values[length(n_obs)] - 
  SST_anom_lm$fitted.values[1]
```

```{r}
#| include: false
### Annual values
SST_full <- SST_full %>% 
    mutate(Year=year(Date_Time), 
           Anom=SST_anom_ts)

ann_SST <- SST_full %>%
  group_by(Year) %>%
  summarise(SST_degC = mean(SST_degC, na.rm = TRUE))
ann_anom <- SST_full %>%
  group_by(Year) %>%
  summarise(Anom = mean(Anom, na.rm = TRUE))
ann_mean_RptYr <- ann_SST$SST_degC[which(ann_SST$Year == RptYr)]
ann_anom_RptYr <- ann_anom$Anom[which(ann_SST$Year == RptYr)]
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# This section includes some error checks to prompt fixing the text
if (any(summary(SST_lm)$coefficients[,4] > 0.05)) {
  print('The linear fit to monthly values is not signficant. Remove text related to the linear trend.')
}

if (any(summary(SST_anom_lm)$coefficients[,4] > 0.05)) {
  print('The linear fit to anomaly values is not signficant. Remove text related to the linear trend.')
}

# Pull out the MONTHLY values we need for the report & plots
yr_of_int <- which(SST_full$Year == RptYr)
prev_yrs <- which(SST_full$Year < RptYr)
all_yrs <- which(SST_full$Year <= RptYr)
monthly_max_RptYr <- max(SST_full$SST_degC[yr_of_int])
monthly_min_RptYr <- min(SST_full$SST_degC[yr_of_int])
monthly_max_PrevYrs <- max(SST_full$SST_degC[prev_yrs])
monthly_min_PrevYrs <- min(SST_full$SST_degC[prev_yrs])

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The greatest monthly value was during the report year.  Revise text to reflect this.')
}

if (monthly_min_RptYr < monthly_min_PrevYrs) {
  print('The lowest monthly value was during the report year.  Revise text to reflect this.')
}
```

```{r}
#| include: false
# Write csv for portal
# Note that output csvs go in their own folder
# Rounding to avoid too many significant figures.
# Thanks to Johanna Wren for the code! https://github.com/pwoodworth-jefcoats/climate-indicators/issues/29
ann_SST <- rename(ann_SST, `Degrees C` = SST_degC)
ann_SST$`Degrees C` <- format(round(ann_SST$`Degrees C`, digits=2), nsmall = 2)
#write_csv(ann_SST, file = paste(here(), '/PelagicClimate_', RptYr, '/SST_', RptYr, '.csv', sep = ""))

ann_anom <- rename(ann_anom, `Degrees C` = Anom)
ann_anom$`Degrees C` <- format(round(ann_anom$`Degrees C`, digits=2), nsmall = 2)
#write_csv(ann_anom, file = paste(here(), '/PelagicClimate_', RptYr, '/SSTanomaly_', RptYr, '.csv', sep = ""))
```

```{r}
#| include: false
# Write csv for dashboard
# Note that dashboard output has its own folder

# Add columns for month, variable ID, and units
SST_dashboard <- SST_full %>% 
  mutate(Month=month(Date_Time), 
         ID='SST', 
         Value_lm=SST_lm$fitted.values, 
         Anom_lm=SST_anom_lm$fitted.values, 
         Units='Deg C', 
         Value=SST_degC) %>% 
  dplyr::select(Date_Time, Year, Month, Value, Anom, ID, Value_lm, Anom_lm, Units)

#write.csv(SST_dashboard, file = here("Indicator_Dashboard", "Data", paste0('SST_Dashboard_Data_', RptYr, '.csv')), row.names =F)
```

```{r}
#| include: false
# Borrowing code from the dashboard for this chunk
# so that figures look the same across products
indicator_data <- SST_dashboard |>
  filter(Year <= RptYr)

# Create color palette for easy reference 
oceans <- nmfs_palette("oceans")(3) # 1 = report_year, 3 = previous years
crustacean <- nmfs_palette("crustacean")(4) # 1 = linear trend
coral <- nmfs_palette("coral")(3) # 3 = annual average for Rpt Yr
ann_grey <- "#A5AAAF" # annual means; in NMFS branding guide as dove gray tint 1
waves <- nmfs_palette("waves")(3) # annual means; in NMFS branding guide but not in package
seagrass <- nmfs_palette("seagrass")(3)
pal <- c(oceans[3], coral[2], waves[2], coral[3], crustacean[2])

# Calculate annual means 
ann_vals <- indicator_data |>
  group_by(Year) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Identify the current year, to overlay on plot
given_yr <- indicator_data |>
  filter(Year == RptYr)

given_yr_ann <- bind_cols(Value = rep(ann_vals$Value[dim(ann_vals)[1]]),
                          Date_Time = given_yr$Date_Time,
                          Anom = as.numeric(ann_anom$`Degrees C`[which(ann_anom$Year == RptYr)]))

# x axis ticks and grid
x_grid <- c(seq.POSIXt(min(indicator_data$Date_Time), max(indicator_data$Date_Time), by='5 year'))

# y-axis limits
ts_y_lim <-c(17, 25)
anom_y_lim <- c(-1.5, 1.5)

# Defining the theme up front so we don't have to repeat it
plotTheme <- theme_minimal() +
  theme(panel.border = element_rect(color='#F1F3F3'),
        panel.grid.major = element_line(color='#F1F3F3'),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        axis.title = element_text(size=14, face="bold"), 
        axis.text = element_text(size=12),
        legend.position = 'bottom', 
        legend.background = element_rect_round(fill = 'gray90', color='transparent', radius = 0.3),
        legend.text = element_text(size=12),legend.key.spacing.y = unit(-0.13,'cm'), legend.key.spacing.x = unit(1.5,'cm'))

# Plot using ggplot
legendInfo <- as.factor(c(paste0('Mothly average, 1985-', RptYr-1), 
                          paste0('Annual average, 1985-', RptYr-1), 
                          paste0('Mothly average, ', RptYr), 
                          paste0('Annual average, ', RptYr),
                          paste0('Long-term Trend')))
levels(legendInfo) <- levels(legendInfo)[c(4,5,1,2,3)] # To arrange in preferred order for the legend, otherwise it's alphabetical

p <- ggplot() +
  geom_path(data=indicator_data, aes(Date_Time, Value, color=legendInfo[2])) +
  geom_path(data=ann_vals, aes(dmy_hm(paste0('01-01-',Year,' 00:00')), Value, color=legendInfo[5])) +
  geom_path(data=indicator_data, aes(Date_Time, Value_lm, color=legendInfo[3])) +
  geom_path(data=given_yr, aes(Date_Time, Value, color=legendInfo[4])) +
  geom_path(data=given_yr_ann, aes(Date_Time, Value, color=legendInfo[1])) +
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = F) +
  scale_y_continuous(expand = F, limits = ts_y_lim, breaks=(seq(min(ts_y_lim),max(ts_y_lim),by=1))) +
  scale_color_manual(name=NULL, values=pal[c(1,3,5,2,4)]) +
  ylab('Sea Surface Temperature (°C)') +
  xlab('Year') +
  guides(color = guide_legend(ncol = 2)) +
  plotTheme

#ggsave(paste0(Dir, '/SST_ts_', RptYr, '_ggolot.pdf'), plot=p, width=6, height=5, units='in')
#ggsave(paste0(Dir, '/SST_ts_', RptYr, '_ggolot.png'), plot=p, width=6, height=5, units='in')


### Anomaly plot
# Calculate annual means
ann_vals <- indicator_data |>
  group_by(Year) |>
  summarise(Anom = mean(Anom, na.rm = TRUE)) 

p_ann <- ggplot() +
  geom_hline(yintercept = 0, linewidth=0.25) +
  geom_path(data=indicator_data, aes(Date_Time, Anom, color=legendInfo[2])) +
  geom_path(data=ann_vals, aes(dmy_hm(paste0('01-01-',Year,' 00:00')), Anom, color=legendInfo[5])) +
  geom_path(data=indicator_data, aes(Date_Time, Anom_lm, color=legendInfo[3])) +
  geom_path(data=given_yr, aes(Date_Time, Anom, color=legendInfo[4])) +
  geom_path(data=given_yr_ann, aes(Date_Time, Anom, color=legendInfo[1])) +
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = F) +
  scale_y_continuous(expand = F, limits = anom_y_lim, breaks=(seq(min(anom_y_lim),max(anom_y_lim),by=0.5))) +
  scale_color_manual(name=NULL, values=pal[c(1,3,5,2,4)]) +
  ylab('Sea Surface Temperature Anomaly (°C)') +
  xlab('Year') +
  guides(color = guide_legend(ncol = 2)) +
  plotTheme

#ggsave(paste0(Dir, '/SST_anom_ts_', RptYr, '_ggolot.pdf'), plot=p_ann, width=6, height=5, units='in')
#ggsave(paste0(Dir, '/SST_anom_ts_', RptYr, '_ggolot.png'), plot=p_ann, width=6, height=5, units='in')

```

```{r}
#| warning: false
#| include: false
# # After running the chunks above and prepping the map data, you can uncomment and run this chunk.
#
# Load data
maps <- read.csv(paste(Dir, '/SST_map_data_', RptYr, '.csv', sep = ""))

# Filter to the given year and the anomaly
maps_RptYr <- maps |> filter(ID == "SST")
maps_anom <- maps |> filter(ID == "SST_anom")

# # Adding a column that assignes line types to contours so we can make a legend
# maps_anom$ltyF <- factor(ifelse(maps_anom$layer > 0.5, 'Warmer', ifelse(maps_anom$layer < -0.5, 'Cooler', 'No Change')), levels=c('Cooler', 'Warmer', 'No Change'))

# Map elements and aesthetics
waves <- nmfs_palette("waves")(3)
pal <- rev(waves)
ll_rect_color <- "white" #outline for ll fishing grounds box
fill_scale <- scale_fill_gradient2(name='SST', low = oceans[2], mid = "white", high = coral[2], midpoint = 0, limits=c(-4,4), na.value = 'gray')
#fill_scale <- scale_fill_gradientn(name = "SST", colors = pal, limits = c(0, 31))
# Contour breaks
breakInt <- seq(5,30,by=5)

# Create map (anomaly as base, temp as lines)
p_map <- ggplot() +
  geom_raster(data = maps_anom, mapping = aes(x = x_disp, y = y, fill = layer)) + #data as raster
  geom_textcontour(data = maps_RptYr, aes(x = x_disp, y = y, z = layer), breaks = breakInt, 
                     linetype='dashed', color='black', alpha=0.4, size = 3, ) + # add temperature contours
  annotation_borders('world2', fill='#A5AAAF', colour='gray40', linewidth=0.2) + #add land to the map
  geom_rect(aes(xmin = 180, xmax = 240, ymin = 15, ymax = 45), color = 'black', fill = NA, linewidth = 0.5) + #ll fishing grounds box
  annotate("text", x = 210, y = 48, label = "Longline fishing grounds", size = 3.7, color = 'black') +
  fill_scale + #indicator-dependent color and scale from above
  coord_quickmap(expand = F, xlim=c(120,260), ylim=c(0,55)) + #don't expand past x/y limits
  scale_x_continuous(breaks=seq(120,260,30), labels = paste0(c(seq(120,180,30), seq(150,100,-30)), c(rep('°E',2),"",rep('°W',2)))) + # adding degrees to the breaks
  scale_y_continuous(breaks=seq(0,55,15), labels = paste0(seq(0,55,15), c('',rep('°N',3)))) + # adding degrees to the breaks
  xlab("") + ylab("") +
  theme_bw() +
  theme(panel.grid = element_line(color = "black", linewidth = 0.1),
        legend.background = element_rect(fill = 'transparent'), legend.key.height = unit(1, "null"),
        legend.key.width = unit(0.5, 'cm'),  #scales color bar to height of the plot
        legend.margin = margin(0, 0, 0, 0),  #fixing margins on color bar
        legend.text = element_text(size = 10),  #text size for color bar
        plot.title = element_text(hjust = 0.5), 
        plot.margin=unit(c(0,0,0.02,0), "null")) + #centering the plot title. Default is left align
 guides(fill = guide_colorbar(title.position='right', title="Temperature anomaly (°C)",
                                 title.theme=element_text(angle=270, hjust=0.5, vjust=0.5)),
           linetype=guide_legend(position = 'bottom', 
                                 title.position='top', 
                                 title.theme=element_text(hjust=0.5, vjust=1))) #+  #this bit of code gives different aesthetics to the colorbar on the left and legend on the bottom.
  #labs(title = paste(RptYr, 'anomaly (shaded) and average (contours)'))  #add figure title

#ggsave(filename = paste0(Dir, '/SST_map_', RptYr, '.pdf'), plot=p_map, width = 7, height = 3.8)
#ggsave(filename = paste0(Dir, '/SST_map_', RptYr, '.png'), plot=p_map, width = 7, height = 3.8)

```

```{r}
#| echo: false
#| warning: false
#| label: fig-ch7-SSTmap
#| fig.cap:  !expr paste('Average', RptYr, 'sea surface temperature difference from the 1985–2009 average (shaded) and sea surface temperature (contoured). The black rectangle identifies the area targeted by Hawaiʻi’s longline fisheries. SST is averaged over this area for the time series shown in @fig-ch7-SST and @fig-ch7-SSTanom')
#| fig.align: 'center'
#| fig-width: 7
#| fig-height: 3
#| out-width: "95%"

p_map

```

```{r}
#| echo: false
#| label: fig-ch7-SST
#| fig.cap: 'Time series of monthly average sea surface temperature over the longline fishing grounds outlined in @fig-ch7-SSTmap'
#| fig.align: 'center'
#| fig-width: 6
#| fig-height: 5
#| out-width: "90%"

p

```

```{r}
#| echo: false
#| label: fig-ch7-SSTanom
#| fig.cap: 'Time series of monthly average sea surface temperature anomaly over the longline fishing grounds outlined in @fig-ch7-SSTmap'
#| fig.align: 'center'
#| fig-width: 6
#| fig-height: 5
#| out-width: "90%"

p_ann

```

**Rationale:** Sea surface temperature is one of the most directly observable existing measures for tracking increasing ocean temperatures. SST varies in response to natural climate cycles such as the El Niño – Southern Oscillation (ENSO) and is rising as a result of anthropogenic climate change. Both short-term variability and long-term trends in SST impact the marine ecosystem. Understanding the mechanisms through which organisms are impacted and the time scales of these impacts is an area of active research.  
\
**Status:** Annual mean SST was `r signif(ann_mean_RptYr,3)` ºC in `r RptYr`. Over the period of record, SST across the longline fishing grounds has increased by `r signif(delta_SST_lm, 1)` ºC and the monthly SST anomaly increased by `r signif(delta_SST_anom_lm, 1)` ºC, both at a rate of roughly 0.025 ºC yr^-^^1^. Monthly SST values in `r RptYr` ranged from `r signif(monthly_min_RptYr, 3)`–`r signif(monthly_max_RptYr, 3)` ºC, within the range of temperatures experienced over the past several decades (`r signif(monthly_min_PrevYrs, 3)`–`r signif(monthly_max_PrevYrs, 3)` ºC). Overall, SST was at or above the long-term average across the Hawaiʻi longline region in `r RptYr`.   
\
**Description:** Satellite remotely sensed monthly sea surface temperature (SST) is averaged across the Hawaiʻi-based longline fishing grounds (15° – 45°N, 180° – 120°W). A time series of monthly mean SST averaged over the Hawaiʻi longline region is presented. Additionally, spatial climatologies and anomalies are shown. CoralTemp data are used to calculate this indicator.  
\
**Timeframe:** Monthly.  
\
**Region/Location:** Hawaiʻi longline region: 15° – 45°N, 180° – 120°W.  
\
**Measurement Platform:** Satellite.  
\
**Data available at:** <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/CRW_sst_v3_1_monthly>, <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/CRW_sst_v3_1_1985-2009-clim>, and <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/CRW_sst_v3_1_2024-clim>.   
\
**Sourced from:** @NOAA-OceanWatchUnknown-ie. Graphics produced in part using @Stawitz2026-kw.  
\  
 <!--
## Additional Information
SST data are geographically subset and spatially averaged using the pyFerret script `SSTaccess.jnl` which can be found in the [Sea_Surface_Temperature folder](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Sea_Surface_Temperature).  You could do this step in R, but I use [PyFerret](https://ferret.pmel.noaa.gov/Ferret/) because it's freely available software developed by NOAA specifically for working with large gridded datasets.  
\
A plot of the residuals from the linear model showed that they were evenly distributed, although were more positive (~4.0 max) than negative (~-3.0 min). The residuals for the anomaly model were also fairly evenly distributed in terms of values, however they did appear to exhibit some periodicity.  
\
To prepare the spatial data for mapping, you'll need to run `map_data_for_dashboard.R`, which can be found in the [PreProcessing](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Indicator_Dashboard/PreProcessing) folder of the [Indicator_Dashboard](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Indicator_Dashboard) folder in this repository.  
-->
