### Temperature at 200-300m Depth

```{r}
#| message: false
#| warning: false
#| include: false

# This includes all commonly used packages and variables, like RptYr and eliminates the need to change them in every chapter. 
source("_common.R")
```

```{r}
#| include: false

# Set path to variable: Temperature_at_Depth
# This is where the data are and where the plots will go
Dir <- here("Temperature_at_Depth")
# Check if it exists; if not, create it
if (!dir.exists(Dir)) {
  dir.create(Dir)
}
```

```{r}
#| include: false
### Load data
# Monthly temperature
TatD_full <- read.csv(file = paste0(Dir, '/T_at_200300_ts.dat'), skip = 9, header = FALSE) %>% 
  rename(Date_Time=V1, TatD_degC=V2) %>% 
  mutate(Date_Time=dmy_hm(Date_Time))
```

```{r}
#| include: false
# Remove seasonal means to calculate anomalies
TatD_climo <- matrix(NA, nrow = nrow(TatD_full), ncol = 1)
for (m in seq(1,12,1)) {
  mo_of_int <- which(month(TatD_full$Date_Time) == m)
  TatD_climo[mo_of_int,1] <- mean(TatD_full$TatD_degC[mo_of_int])
}

TatD_anom_ts <- TatD_full$TatD_degC - TatD_climo
```

```{r}
#| include: false
### Linear fit 
# Note that this assumes that observations are equally spaced in time, which they're  not
n_obs <- seq(1, length(TatD_full$TatD_degC), 1)
TatD_lm <- lm(TatD_full$TatD_degC ~ n_obs)
TatD_anom_lm <- lm(TatD_anom_ts ~ n_obs)

# Change over time
delta_TatD_lm <- TatD_lm$fitted.values[length(n_obs)] - TatD_lm$fitted.values[1]

delta_TatD_anom_lm <- TatD_anom_lm$fitted.values[length(n_obs)] - TatD_anom_lm$fitted.values[1]
```

```{r}
#| include: false
### Annual values
TatD_full <- TatD_full %>% 
    mutate(Year=year(Date_Time), 
           Anom=TatD_anom_ts)

ann_TatD <- TatD_full %>%
  group_by(Year) %>%
  summarise(TatD_degC = mean(TatD_degC, na.rm = TRUE))
ann_anom <- TatD_full %>%
  group_by(Year) %>%
  summarise(Anom = mean(Anom, na.rm = TRUE))
ann_mean_RptYr <- ann_TatD$TatD_degC[which(ann_TatD$Year == RptYr)]
ann_anom_RptYr <- ann_anom$Anom[which(ann_TatD$Year == RptYr)]
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# This section includes some error checks to prompt fixing the text
if (any(summary(TatD_lm)$coefficients[,4] > 0.05)) {
  print('The linear fit to monthly values is not signficant. Remove text related to the linear trend.')
}

if (any(summary(TatD_anom_lm)$coefficients[,4] > 0.05)) {
  print('The linear fit to anomaly values is not signficant. Remove text related to the linear trend.')
}

# Pull out the MONTHLY values we need for the report & plots
yr_of_int <- which(TatD_full$Year == RptYr)
prev_yrs <- which(TatD_full$Year < RptYr)
all_yrs <- which(TatD_full$Year <= RptYr)
monthly_max_RptYr <- max(TatD_full$TatD_degC[yr_of_int])
monthly_min_RptYr <- min(TatD_full$TatD_degC[yr_of_int])
monthly_max_PrevYrs <- max(TatD_full$TatD_degC[prev_yrs])
monthly_min_PrevYrs <- min(TatD_full$TatD_degC[prev_yrs])

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The greatest monthly value was during the report year.  Revise text to reflect this.')
}

if (monthly_min_RptYr < monthly_min_PrevYrs) {
  print('The lowest monthly value was during the report year.  Revise text to reflect this.')
}
```

```{r}
#| include: false
# Write csv for portal
# Note that output csvs go in their own folder
# Rounding to avoid too many significant figures.
# Thanks to Johanna Wren for the code! https://github.com/pwoodworth-jefcoats/climate-indicators/issues/29
ann_TatD <- rename(ann_TatD, `Degrees C` = TatD_degC)
ann_TatD$`Degrees C` <- format(round(ann_TatD$`Degrees C`, digits=2), nsmall = 2)
#write_csv(ann_TatD, file = paste(here(), '/PelagicClimate_', RptYr, '/Tat200to300m_', RptYr, '.csv', sep = ""))

# Anomaly time series not included given that it tracks actual time series so closely.
```

```{r}
#| include: false
# Write csv for dashboard
# Note that dashboard output has its own folder

# Add columns for month, variable ID, and units
TatD_dashboard <- TatD_full %>% 
  mutate(Month=month(Date_Time), 
         ID='TatD', 
         Value_lm=TatD_lm$fitted.values, 
         Anom_lm=NA, # No anomaly for this variable  #TatD_anom_lm$fitted.values, 
         Units='Deg C', 
         Anom=NA, # No anomaly for this variable
         Value=TatD_degC) %>% 
  dplyr::select(Date_Time, Year, Month, Value, Anom, ID, Value_lm, Anom_lm, Units)

#write.csv(TatD_dashboard, file = here("Indicator_Dashboard", "Data", paste('TatD_Dashboard_Data_', RptYr, '.csv', sep = "")), row.names = F)
```

```{r}
#| include: false
# Borrowing code from the dashboard for this chunk
# so that figures look the same across products
indicator_data <- TatD_dashboard |>
  filter(Year <= RptYr)

# Create color palette for easy reference 
oceans <- nmfs_palette("oceans")(3) # 1 = report_year, 3 = previous years
crustacean <- nmfs_palette("crustacean")(4) # 1 = linear trend
coral <- nmfs_palette("coral")(3) # 3 = annual average for Rpt Yr
ann_grey <- "#A5AAAF" # annual means; in NMFS branding guide as dove gray tint 1
waves <- nmfs_palette("waves")(3) # annual means; in NMFS branding guide but not in package
seagrass <- nmfs_palette("seagrass")(3)
pal <- c(oceans[3], coral[2], waves[2], coral[3], crustacean[2])

# # Formatting
# plot_title_font <- list(size = 14)
# plot_height <- 350 #in pixels

# Calculate annual means 
ann_vals <- indicator_data |>
  group_by(Year) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Identify the current year, to overlay on plot
given_yr <- indicator_data |>
  filter(Year == RptYr)

given_yr_ann <- bind_cols(Value = rep(ann_vals$Value[dim(ann_vals)[1]]),
                          Date_Time = given_yr$Date_Time,
                          Anom = ann_anom$`Anom`[which(ann_anom$Year == RptYr)])

# x axis ticks and grid
# Made a sequence instead so we don't have to manually update it every five years
x_grid <- c(seq.POSIXt(min(indicator_data$Date_Time), max(indicator_data$Date_Time), by='5 year'))
# x axis limit
ts_x_lim <- c(min(indicator_data$Date_Time), max(indicator_data$Date_Time))

# y-axis limits
ts_y_lim <-c(10.8, 11.7)
anom_y_lim <- c(-1.5, 1.5)

# Defining the theme up front so we don't have to repeat it
plotTheme <- theme_minimal() +
  theme(panel.border = element_rect(color='#F1F3F3'),
        panel.grid.major = element_line(color='#F1F3F3'),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        axis.title = element_text(size=14, face="bold"), 
        axis.text = element_text(size=12),
        legend.position = 'bottom', 
        legend.background = element_rect_round(fill = 'gray90', color='transparent', radius = 0.3),
        legend.text = element_text(size=11), legend.key.spacing.y = unit(-0.13,'cm'), legend.key.spacing.x = unit(1.5,'cm'))

# Plot using ggplot
legendInfo <- as.factor(c(paste0('Mothly average, 1980-', RptYr-2), 
                          paste0('Annual average, 1980-', RptYr-2), 
                          paste0('Mothly average, ', RptYr-1), 
                          paste0('Annual average, ', RptYr-1),
                          paste0('Long-term Trend')))
levels(legendInfo) <- levels(legendInfo)[c(4,5,1,2,3)] # To arrange in preferred order for the legend, otherwise it's alphabetical

p <- ggplot() +
  geom_path(data=indicator_data, aes(Date_Time, Value, color=legendInfo[2])) +
  geom_path(data=ann_vals, aes(dmy_hm(paste0('01-01-',Year,' 00:00')), Value, color=legendInfo[5])) +
  geom_path(data=indicator_data, aes(Date_Time, Value_lm, color=legendInfo[3])) +
  geom_path(data=given_yr, aes(Date_Time, Value, color=legendInfo[4])) +
  geom_path(data=given_yr_ann, aes(Date_Time, Value, color=legendInfo[1])) +
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = F) +
  scale_y_continuous(expand = F, limits = ts_y_lim, breaks=(seq(min(ts_y_lim),max(ts_y_lim),by=0.1))) +
  scale_color_manual(name=NULL, values=pal[c(1,3,5,2,4)]) +
  ylab('200-300 m Temperature (°C)') +
  xlab('Year') +
  guides(color = guide_legend(ncol = 2)) +
  plotTheme

#ggsave(paste0(Dir, '/TempAtDepth_ts_', RptYr, '_ggolot.pdf'), plot=p, width=6, height=5, units='in')
#ggsave(paste0(Dir, '/TempAtDepth_ts_', RptYr, '_ggolot.png'), plot=p, width=6, height=5, units='in')
```

```{r}
#| include: FALSE
# After running the chunks above and prepping the map data, you can uncomment and run this chunk.
# 
# Load data
maps <- read.csv(paste(Dir, '/TempAtDepth_map_data_', RptYr, '.csv', sep = ""))
 
# Filter to the given year and the anomaly
maps_RptYr <- maps |> filter(ID == "TatD")
maps_anom <- maps |> filter(ID == "TatD_anom")

# # Adding a column that assignes line types to contours so we can make a legend
# maps_anom$ltyF <- factor(ifelse(maps_anom$layer > 0.5, 'Warmer', ifelse(maps_anom$layer < -0.5, 'Cooler', 'No Change')), levels=c('Cooler', 'Warmer', 'No Change'))
 
# Map elements and aesthetics
waves <- nmfs_palette("waves")(3)
pal <- rev(waves)
ll_rect_color <- "black" #outline for ll fishing grounds box
fill_scale <- scale_fill_gradient2(name='SST', low = oceans[2], mid = "white", high = coral[2], midpoint = 0, na.value = 'gray')
#fill_scale <- scale_fill_gradientn(name = "TatD", colors = pal, limits = c(0, 31))
# Contour breaks
breakInt <- seq(0,20, by=4)

# Create map
p_map <- ggplot() +
  geom_raster(data = maps_anom, mapping = aes(x = x_disp, y = y, fill = layer)) + #data as raster
  geom_textcontour(data = maps_RptYr, aes(x = x_disp, y = y, z = layer), breaks = breakInt, 
                     linetype='dashed', color='black', alpha=0.4, size = 3, ) + # add temperature contours
  annotation_borders('world2', fill='#A5AAAF', colour='gray40', linewidth=0.2) + #add land to the map
  geom_rect(aes(xmin = 180, xmax = 240, ymin = 15, ymax = 45), color = ll_rect_color, fill = NA, linewidth = 0.5) + #ll fishing grounds box
  annotate("text", x = 210, y = 48, label = "Longline fishing grounds", size = 3.7, color = 'black') +
  fill_scale + #indicator-dependent color and scale from above
  coord_quickmap(expand = F, xlim=c(120,260), ylim=c(0,55)) + #don't expand past x/y limits
  scale_x_continuous(breaks=seq(120,260,30), labels = paste0(c(seq(120,180,30), seq(150,100,-30)), c(rep('°E',2),"",rep('°W',2)))) + # adding degrees to the breaks
  scale_y_continuous(breaks=seq(0,55,15), labels = paste0(seq(0,55,15), c('',rep('°N',3)))) + # adding degrees to the breaks
  xlab("") + ylab("") +
  theme_bw() +
  theme(panel.grid = element_line(color = "black", linewidth = 0.1),
        legend.background = element_rect(fill = 'transparent'), legend.key.height = unit(1, "null"),
        legend.key.width = unit(0.5, 'cm'),  #scales color bar to height of the plot
        legend.margin = margin(0, 0, 0, 0),  #fixing margins on color bar
        legend.text = element_text(size = 10),  #text size for color bar
        plot.title = element_text(hjust = 0.5), 
        plot.margin=unit(c(0,0,0.02,0), "null")) + #centering the plot title. Default is left align
 guides(fill = guide_colorbar(title.position='right', title="Temperature anomaly (°C)",
                                 title.theme=element_text(angle=270, hjust=0.5, vjust=0.5)),
           linetype=guide_legend(position = 'bottom', 
                                 title.position='top', 
                                 title.theme=element_text(hjust=0.5, vjust=1))) #+  #this bit of code gives different aesthetics to the colorbar on the left and legend on the bottom.
  #labs(title = paste(RptYr, 'anomaly (shaded) and average (contours)'))  #add figure title

#ggsave(filename = paste0(Dir, '/TempAtDepth_map_', RptYr, '_test.pdf'), plot=p_map, width = 7, height = 3.8)
#ggsave(filename = paste0(Dir, '/TempAtDepth_map_', RptYr, '_test.png'), plot=p_map, width = 7, height = 3.8)

```

```{r}
#| echo: false
#| warning: false
#| label: fig-ch8-TatDmap
#| fig.cap:  !expr paste('The difference in temperatues at 200-300 m depth in ', RptYr, 'from the 1980-2009 average (shaded) with the temperatue at 200-300 m depth as contours. The black rectangle identifies the area targeted by Hawaiʻi’s longline fisheries. Temperature is averaged over this area for the time series shown in @fig-ch8-TatD')
#| fig.align: 'center'
#| fig-width: 7
#| fig-height: 3
#| out-width: "95%"

p_map

```

```{r}
#| echo: false
#| label: fig-ch8-TatD
#| fig.cap: 'Time series of monthly 200–300 m temperatures over the longline fishing grounds outlined in @fig-ch8-TatDmap'
#| fig.align: 'center'
#| fig-width: 6
#| fig-height: 5
#| out-width: "90%"

p

```

**Rationale:** The temperature at 200–300 m reflects the temperature in the mid-range of depths targeted by the deep-set bigeye tuna fishery. Bigeye have preferred thermal habitat, generally staying within temperatures ranging from 8–14 °C while they are at depth [@Howell2010-rq]. Changes in ocean temperature at depth will impact tuna, and in turn, potentially impact their catchability. Understanding the drivers of sub-surface temperature trends and their ecosystem impacts is an area of active research.  
\
**Status:** In `r RptYr`, 200–300 m temperatures ranged from `r signif(monthly_min_RptYr, 4)`–`r signif(monthly_max_RptYr, 4)` °C with an average value of `r signif(ann_mean_RptYr,4)` °C. These temperatures are within the range of temperatures experienced over the past several decades (`r signif(monthly_min_PrevYrs, 4)`–`r signif(monthly_max_PrevYrs, 4)` °C) and are within the bounds of bigeye tuna’s preferred deep daytime thermal habitat (8–14 °C). Over the period of record (1980–`r RptYr`), 200–300 m temperatures have declined by `r abs(signif(delta_TatD_lm, 1))` °C. The spatial pattern of temperature anomalies was mixed with temperatures at depth just north of the main Hawaiian Islands roughly 0.5–1.5 °C below average, and temperatures north and west of the Hawaiian Islands 0.5-2 °C above average.   
\
**Description:** Ocean temperature at 200–300 m depth is averaged across the Hawaiʻi-based longline fishing grounds (15° – 45°N, 180° – 120°W). Global Ocean Data Assimilation System (GODAS) data are used. GODAS incorporates global ocean data from moorings, expendable bathythermographs (XBTs), and Argo floats.   
\
**Timeframe:** Annual, monthly.  
\
**Region/Location:** Hawaii longline region: 15° – 45°N, 180° – 120°W.  
\
**Measurement Platform:** _In-situ_ sensors, model.  
\
**Sourced from:** @NOAAUnknown-ei and @APDRCUnknown-sl.  Graphics produced in part using @Stawitz2026-kw. NCEP Global Ocean Data Assimilation System (GODAS) data provided by the NOAA PSL, Boulder, Colorado, USA, from their website at https://psl.noaa.gov

<!--
## Additional Information
Temperature-at-depth data are geographically subset and spatially averaged using the pyFerret script `TatDepth_access.jnl` which can be found in the [Temperature_at_Depth](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Temperature_at_Depth) folder.  You could do this step in R, but I use [PyFerret](https://ferret.pmel.noaa.gov/Ferret/) because it's freely available software developed by NOAA specifically for working with large gridded datasets.  
\
A plot of the residuals from the linear model showed that they were evenly distributed. The residuals for the anomaly model were also fairly evenly distributed in terms of values.  Both sets of residual exhibited periodicity that closely track the time series pattern.  
\
To prepare the spatial data for mapping, you'll need to run `map_data_for_dashboard.R`, which can be found in the [PreProcessing](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Indicator_Dashboard/PreProcessing) folder of the [Indicator_Dashboard](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Indicator_Dashboard) folder in this repository.
-->
