### Ocean Color

```{r}
#| message: false
#| warning: false
#| include: false

# This includes all commonly used packages and variables, like RptYr and eliminates the need to change them in every chapter. 
source("_common.R")
```

```{r}
#| include: false

# Set path to variable: Ocean_Color
# This is where the data are and where the plots will go
Dir <- here("Ocean_Color")
# Check if it exists; if not, create it
if (!dir.exists(Dir)) {
  dir.create(Dir)
}
```

```{r}
#| include: false
### Load data
# Monthly ocean color
# read in all the time series file names
Chl_files <- list.files(path=Dir, pattern='chl_ts_', full.names = T)
# Read in all the files
Chl_full <- lapply(Chl_files, read.csv, skip=8, header=F) %>% 
  bind_rows() %>% 
  rename(Date_Time=V1, Chl_mgm3=V2) %>% 
  mutate(Date_Time=dmy_hm(Date_Time)) %>% 
  arrange(Date_Time) # note that this method doesn't read the files in chronological order, so we sort them here to make sure everything lines up properly down the line
```

```{r}
#| include: false
# Remove seasonal means to calculate anomalies
Chl_climo <- matrix(NA, nrow = nrow(Chl_full), ncol = 1)
for (m in seq(1,12,1)) {
  mo_of_int <- which(month(Chl_full$Date_Time) == m)
  Chl_climo[mo_of_int,1] <- mean(Chl_full$Chl_mgm3[mo_of_int])
}

Chl_anom_ts <- Chl_full$Chl_mgm3 - Chl_climo
```

```{r}
#| include: false
### Linear fit 
# Note that this assumes that observations are equally spaced in time, which they're  not
n_obs <- seq(1, length(Chl_full$Chl_mgm3), 1)
Chl_lm <- lm(Chl_full$Chl_mgm3 ~ n_obs)
Chl_anom_lm <- lm(Chl_anom_ts ~ n_obs)

# Change over time
delta_Chl_lm <- Chl_lm$fitted.values[length(n_obs)] - Chl_lm$fitted.values[1]

delta_Chl_anom_lm <- Chl_anom_lm$fitted.values[length(n_obs)] - Chl_anom_lm$fitted.values[1]
```

```{r}
#| include: false
### Annual values
Chl_full <- Chl_full %>% 
    mutate(Year = year(Date_Time), 
           Anom = Chl_anom_ts) # Add in anomaly to make things easier down the road

ann_Chl <- Chl_full %>%
  group_by(Year) %>%
  summarise(Chl_mgm3 = mean(Chl_mgm3, na.rm = TRUE))
ann_anom <- Chl_full %>%
  group_by(Year) %>%
  summarise(Anom = mean(Anom, na.rm = TRUE))
ann_mean_RptYr <- ann_Chl$Chl_mgm3[which(ann_Chl$Year == RptYr)]
ann_anom_RptYr <- ann_anom$Anom[which(ann_Chl$Year == RptYr)]
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# This section includes some error checks to prompt fixing the text
if (any(summary(Chl_lm)$coefficients[,4] > 0.05)) {
  print('The linear fit to monthly values is not signficant. Remove text related to the linear trend.')
}

if (any(summary(Chl_anom_lm)$coefficients[,4] > 0.05)) {
  print('The linear fit to anomaly values is not signficant. Remove text related to the linear trend.')
}

# Pull out the MONTHLY values we need for the report & plots
yr_of_int <- which(Chl_full$Year == RptYr)
prev_yrs <- which(Chl_full$Year < RptYr)
all_yrs <- which(Chl_full$Year <= RptYr)
monthly_max_RptYr <- max(Chl_full$Chl_mgm3[yr_of_int])
monthly_min_RptYr <- min(Chl_full$Chl_mgm3[yr_of_int])
monthly_max_PrevYrs <- max(Chl_full$Chl_mgm3[prev_yrs])
monthly_min_PrevYrs <- min(Chl_full$Chl_mgm3[prev_yrs])

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The greatest monthly value was during the report year.  Revise text to reflect this.')
}

if (monthly_min_RptYr < monthly_min_PrevYrs) {
  print('The lowest monthly value was during the report year.  Revise text to reflect this.')
}
```

```{r}
#| include: false
# Write csv for portal
# Note that output csvs go in their own folder
# Rounding to avoid too many significant figures.
# Thanks to Johanna Wren for the code! https://github.com/pwoodworth-jefcoats/climate-indicators/issues/29
ann_Chl <- rename(ann_Chl, `mg/m^3` = Chl_mgm3)
ann_Chl$`mg/m^3` <- format(round(ann_Chl$`mg/m^3`, digits=2), nsmall = 2)
#write_csv(ann_Chl, file = paste(here(), '/PelagicClimate_', RptYr, '/CHL_', RptYr, '.csv', sep = ""))

ann_anom <- rename(ann_anom, `mg/m^3` = Anom)
ann_anom$`mg/m^3` <- format(round(ann_anom$`mg/m^3`, digits=2), nsmall = 2)
#write_csv(ann_anom, file = paste(here(), '/PelagicClimate_', RptYr, '/CHLanomaly_', RptYr, '.csv', sep = ""))
```

```{r}
#| include: false
# Write csv for dashboard
# Note that dashboard output has its own folder

# Add columns for month, variable ID, and units
Chl_dashboard <- Chl_full %>% 
  mutate(Month=month(Date_Time), 
         ID='Chl', 
         Value_lm=Chl_lm$fitted.values, 
         Anom_lm=Chl_anom_lm$fitted.values, 
         Units='mg chl per m3', 
         Value=Chl_mgm3) %>% 
  dplyr::select(Date_Time, Year, Month, Value, Anom, ID, Value_lm, Anom_lm, Units)

#write.csv(Chl_dashboard, file = here("Indicator_Dashboard", "Data", paste('Chl_Dashboard_Data_', RptYr, '.csv', sep = "")), row.names = F)
```

```{r}
#| include: false
# Borrowing code from the dashboard for this chunk
# so that figures look the same across products
indicator_data <- Chl_dashboard |>
  filter(Year <= RptYr)

# Create color palette for easy reference 
oceans <- nmfs_palette("oceans")(3) # 1 = report_year, 3 = previous years
crustacean <- nmfs_palette("crustacean")(4) # 1 = linear trend
coral <- nmfs_palette("coral")(3) # 3 = annual average for Rpt Yr
ann_grey <- "#A5AAAF" # annual means; in NMFS branding guide as dove gray tint 1
waves <- nmfs_palette("waves")(3) # annual means; in NMFS branding guide but not in package
seagrass <- nmfs_palette("seagrass")(3)
pal <- c(oceans[3], coral[2], waves[2], coral[3], crustacean[2])

# Calculate annual means 
ann_vals <- indicator_data |>
  group_by(Year) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Identify the current year, to overlay on plot
given_yr <- indicator_data |>
  filter(Year == RptYr)

given_yr_ann <- bind_cols(Value = rep(ann_vals$Value[dim(ann_vals)[1]]),
                          Date_Time = given_yr$Date_Time,
                          Anom = ann_anom$`mg/m^3`[which(ann_anom$Year == RptYr)])

# x axis ticks and grid
x_grid <- c(seq.POSIXt(dmy_hm(010120000000), max(indicator_data$Date_Time), by='5 year'))
# x-axis limits
ts_x_lim <- c(min(indicator_data$Date_Time), max(indicator_data$Date_Time))

# y-axis limits
ts_y_lim <-c(0.07, 0.19)
anom_y_lim <- c(-0.035, 0.035)

# Defining the theme up front so we don't have to repeat it
plotTheme <- theme_minimal() +
  theme(panel.border = element_rect(color='#F1F3F3'),
        panel.grid.major = element_line(color='#F1F3F3'),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        axis.title = element_text(size=14, face="bold"), 
        axis.text = element_text(size=12),
        legend.position = 'bottom', 
        legend.background = element_rect_round(fill = 'gray90', color='transparent', radius = 0.3),
        legend.text = element_text(size=11), legend.key.spacing.y = unit(-0.13,'cm'), legend.key.spacing.x = unit(1.5,'cm'))

# Plot using ggplot
legendInfo <- as.factor(c(paste0('Mothly average, 1998-', RptYr-1), 
                          paste0('Annual average, 1998-', RptYr-1), 
                          paste0('Mothly average, ', RptYr), 
                          paste0('Annual average, ', RptYr),
                          paste0('Long-term Trend')))
levels(legendInfo) <- levels(legendInfo)[c(4,5,1,2,3)] # To arrange in preferred order for the legend, otherwise it's alphabetical

p <- ggplot() +
  geom_path(data=indicator_data, aes(Date_Time, Value, color=legendInfo[2])) +
  geom_path(data=ann_vals, aes(dmy_hm(paste0('01-01-',Year,' 00:00')), Value, color=legendInfo[5])) +
  geom_path(data=indicator_data, aes(Date_Time, Value_lm, color=legendInfo[3])) +
  geom_path(data=given_yr, aes(Date_Time, Value, color=legendInfo[4])) +
  geom_path(data=given_yr_ann, aes(Date_Time, Value, color=legendInfo[1])) +
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = F) +
  scale_y_continuous(expand = F, limits = ts_y_lim, breaks=c(0.07, seq(0.08, 0.8, 0.02), 0.19)) +
  scale_color_manual(name=NULL, values=pal[c(1,3,5,2,4)]) +
  ylab(expression(bold("Chlorophyll-a Concentration (mg " * m^-3* ")"))) +
  xlab('Year') +
  guides(color = guide_legend(ncol = 2)) +
  plotTheme

#ggsave(paste0(Dir, '/Chl_ts_', RptYr, '_ggolot.pdf'), plot=p, width=6, height=5, units='in')
#ggsave(paste0(Dir, '/Chl_ts_', RptYr, '_ggolot.png'), plot=p, width=6, height=5, units='in')

### Anomaly plot
# Calculate annual means
ann_vals <- indicator_data |>
  group_by(Year) |>
  summarise(Anom = mean(Anom, na.rm = TRUE)) 

p_ann <- ggplot() +
  geom_hline(yintercept = 0, linewidth=0.25) +
  geom_path(data=indicator_data, aes(Date_Time, Anom, color=legendInfo[2])) +
  geom_path(data=ann_vals, aes(dmy_hm(paste0('01-01-',Year,' 00:00')), Anom, color=legendInfo[5])) +
  geom_path(data=indicator_data, aes(Date_Time, Anom_lm, color=legendInfo[3])) +
  geom_path(data=given_yr, aes(Date_Time, Anom, color=legendInfo[4])) +
  geom_path(data=given_yr_ann, aes(Date_Time, as.numeric(Anom), color=legendInfo[1])) +
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = F) +
  scale_y_continuous(expand = F, limits = anom_y_lim, 
                     breaks=c(seq(-0.035, 0.005, 0.01), 0, seq(0.005, 0.035, 0.01)), 
                     labels=c(seq(-0.035, 0.005, 0.01), 0, seq(0.005, 0.035, 0.01))) +
  scale_color_manual(name=NULL, values=pal[c(1,3,5,2,4)]) +
  ylab(expression(bold("Chlorophyll-a Concentration (mg "*m^-3*")"))) +
  xlab('Year') +
  guides(color = guide_legend(ncol = 2)) +
  plotTheme

#ggsave(paste0(Dir, '/Chl_anom_ts_', RptYr, '_ggolot.pdf'), plot=p_ann, width=7, height=5, units='in')
#ggsave(paste0(Dir, '/Chl_anom_ts_', RptYr, '_ggolot.png'), plot=p_ann, width=7, height=5, units='in')

```

```{r}
#| include: FALSE
# # After running the chunks above and prepping the map data, you can uncomment and run this chunk.
# 
# Load data
maps <- read.csv(paste(Dir, '/Chl_map_data_', RptYr, '.csv', sep = ""))

# Filter to the given year and the anomaly
maps_RptYr <- maps |> filter(ID == "Chl")
maps_anom <- maps |> filter(ID == "Chl_anom")
 
# Map elements and aesthetics
seagrass <- nmfs_palette("seagrass")(3)
pal <- seagrass
ll_rect_color <- "black" #outline for ll fishing grounds box
fill_scale <- scale_fill_gradientn(
        colors = c("#2166AC", "#4393C3", "#F7F7F7", "#D6604D", "#B2182B"), #c('#00437D', oceans[2], 'white', coral[2], '#A31502'),
        # Linearly space the 5 colors across the [-0.2, 0.2] range
        values = rescale(c(-0.6, -0.2, 0, 0.2, 0.6)),
        limits = c(-0.6, 0.6), # CRITICAL: Sets the color scale endpoints at -0.2 and +0.2
        oob = scales::squish,  # Values < -0.2 and > 0.2 squish directly into darker red
        breaks = c(-0.6, -0.4, -0.2, 0, 0.2, 0.4, 0.6),
        labels = c("≤ -0.6", "-0.4", "-0.2", "0", "0.2", "0.4", "≥ 0.6"),
        name = expression(paste("Chl Anomaly (mg m"^"-3", ")"))
    )
#fill_scale <- scale_fill_gradientn(name = "Chl", trans = "log10", colors = pal, limits = c(0.01, 10), oob = squish) #log scale
breakInt <- seq(0,1,by=0.1)

# Create map
p_map <- ggplot() +
  geom_raster(data = maps_anom, mapping = aes(x = x_disp, y = y, fill = layer)) + #data as raster
  geom_contour(data=maps_RptYr, aes(x = x_disp, y = y, z = layer, linetype = factor(after_stat(level))), breaks = c(0.1, 0.2, 0.3), color = "black", alpha=0.4, linewidth = 0.5, show.legend = F) +
  scale_linetype_manual(values = c("0.1" = "dashed", "0.2" = "solid", "0.3" = "longdash"), guide='none') + # Map contour values to specific line styles
  annotation_borders('world2', fill='#A5AAAF', colour='gray40', linewidth=0.2) + #add land to the map
  geom_rect(aes(xmin = 180, xmax = 240, ymin = 15, ymax = 45), color = ll_rect_color, fill = NA, linewidth = 0.5) + #ll fishing grounds box
  annotate("text", x = 210, y = 48, label = "Longline fishing grounds", size = 3.7, color = 'black') +
  fill_scale + #indicator-dependent color and scale from above
  coord_quickmap(expand = F, xlim=c(120,260), ylim=c(0,55)) + #don't expand past x/y limits
  scale_x_continuous(breaks=seq(120,260,30), labels = paste0(c(seq(120,180,30), seq(150,100,-30)), c(rep('°E',2),"",rep('°W',2)))) + # adding degrees to the breaks
  scale_y_continuous(breaks=seq(0,55,15), labels = paste0(seq(0,55,15), c('',rep('°N',3)))) + # adding degrees to the breaks
  xlab("") + ylab("") +
  theme_bw() +
  theme(panel.grid = element_line(color = "black", linewidth = 0.1),
        legend.background = element_rect(fill = 'transparent'), legend.key.height = unit(1, "null"),
        legend.key.width = unit(0.5, 'cm'),  #scales color bar to height of the plot
        legend.margin = margin(0, 0, 0, 0),  #fixing margins on color bar
        legend.text = element_text(size = 10),  #text size for color bar
        plot.title = element_text(hjust = 0.5), 
        plot.margin=unit(c(0,0,0.02,0), "null")) + #centering the plot title. Default is left align
 guides(fill = guide_colorbar(title.position='right', title="Temperature anomaly (°C)",
                                 title.theme=element_text(angle=270, hjust=0.5, vjust=0.5)),
           linetype=guide_legend(position = 'bottom', 
                                 title.position='top', 
                                 title.theme=element_text(hjust=0.5, vjust=1))) #+  #this bit of code gives different aesthetics to the colorbar on the left and legend on the bottom.
  #labs(title = paste(RptYr, 'anomaly (shaded) and average (contours)'))  #add figure title
  
# ggsave(filename = paste0(Dir, '/Chl_map_', RptYr, '.pdf'), plot=p_map, width = 7, height = 3.8)
# ggsave(filename = paste0(Dir, '/Chl_map_', RptYr, '.png'), plot=p_map, width = 7, height = 3.8)

```

**Rationale:** Phytoplankton are the foundational food source for the fishery. Changes in phytoplankton abundance have been linked to both natural climate variability and anthropogenic climate change. These changes have the potential to impact fish abundance, size, and catch.  
\
**Status:** The mean monthly chlorophyll concentration was `r signif(ann_mean_RptYr, 2)` mg chl m^-^^3^ in `r RptYr`. Monthly mean chlorophyll concentrations ranged from `r signif(monthly_min_RptYr, 2)`–`r signif(monthly_max_RptYr, 2)` mg chl m^-^^3^, which was within the range of values observed during the previous years of the time series (`r signif(monthly_min_PrevYrs, 2)`–`r signif(monthly_max_PrevYrs, 2)` mg chl m^-^^3^).  Both monthly average chlorophyll concentration and chlorophyll anomalies have declined by about `r abs(signif(delta_Chl_anom_lm, 1))` mg chl m^-^^3^ over the period of record.  Chlorophyll concentrations were fairly average across the Hawaiʻi-based longline fishing grounds in `r RptYr`.  
\
**Description:** Satellite remotely sensed ocean color is used to determine chlorophyll concentrations in the pelagic surface ocean. A time series of mean monthly chlorophyll-a concentrations averaged over the Hawaiʻi longline region is presented. Additionally, spatial climatologies and anomalies are shown. European Space Agency (ESA) Climate Change Initiative (CCI) data are used for this indicator [Sathyendranath2018-to].  
\
**Timeframe:** Monthly  
\
**Region/Location:** Hawaiʻi-based longline region: 5° – 45°N, 180° – 120°W  
\
**Measurement Platform:** Satellite  
\
**Data available at:** <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/esa-cci-chla-monthly-v6-0>, <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/esa-cci-chla-1998-2009-clim-v6-0>, and <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/esa-cci-chla-2025-clim_v6-0.html>.  
\
**Sourced from:** @NOAA-OceanWatchUnknown-lk and @Sathyendranath2018-to. Graphics produced in part using Stawitz and Li (2025).  
\

<!--
## Additional Information
Ocean color data are geographically subset and spatially averaged using the pyFerret script `CHLaccess.jnl` which can be found in the Ocean_Color folder.  You could do this step in R, but I use [PyFerret](https://ferret.pmel.noaa.gov/Ferret/) because it's freely available software developed by NOAA specifically for working with large gridded datasets.  
\
A plot of the residuals from the linear model for the observations and anomalies showed that they were evenly distributed, although the anomaly residuals were more positive (~0.03 max) than negative (~-0.02 min). These anomaly residual exhibited periodicity that closely track the time series pattern.  
\
To prepare the spatial data for mapping, you'll need to run `map_data_for_dashboard.R`, which can be found in the [PreProcessing](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Indicator_Dashboard/PreProcessing) folder of the [Indicator_Dashboard](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Indicator_Dashboard) folder in this repository.
-->

