---
title: "Median_Phytoplankton_Size"
author: "phoebe.woodworth-jefcoats@noaa.gov"
format:
  docx:
    reference-doc: SAFE-Reference-Doc.docx
---

## Estimated Median Phytoplankton Size

```{r}
#| include: false
### Load libraries
# Commenting out the plotly, reticulate, python stuff because it is 
# inexplicably not working this year
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
# remotes::install_github("nmfs-fish-tools/nmfspalette")
library(nmfspalette)
# library(plotly)
# library(reticulate)
library(grid)
#devtools::install_github("teunbrand/elementalist")
library(elementalist)
# 
# reticulate::use_miniconda('r-reticulate')
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2025

# Set path to variable: Median_Phytoplankton_Size
# This is where the data are and where the plots will go
Dir <- here("Median_Phytoplankton_Size")
# Check if it exists; if not, create it
if (!dir.exists(Dir)) {
  dir.create(Dir)
}
```

```{r}
#| include: false
### Load data
# Monthly estimated median phytoplankton size
# Generated in ~5-yr batches due to memory limitations
# There's gotta be a way to append a year that's way better than this.  Implement this next year.
## Johanna made an attempt below in 2025
# read in all the time series file names
med_phyto_files <- list.files(path=Dir, pattern='medphyto_ts_', full.names = T)
# Read in all the files
med_phyto_full <- lapply(med_phyto_files, read.csv, skip=8, header=F) %>% 
  bind_rows() %>% 
  rename(Date_Time=V1, ESD_um=V2) %>% 
  mutate(Date_Time=dmy_hm(Date_Time)) %>% 
  arrange(Date_Time) # note that this method doesn't read the files in chronological order, so we sort them here to make sure everything lines up properly down the line
```

```{r}
#| include: false
# Remove seasonal means to calculate anomalies
medphyto_climo <- matrix(NA, nrow = nrow(med_phyto_full), ncol = 1)
for (m in seq(1,12,1)) {
  mo_of_int <- which(month(med_phyto_full$Date_Time) == m)
  medphyto_climo[mo_of_int,1] <- mean(med_phyto_full$ESD_um[mo_of_int])
}

med_phyto_anom_ts <- med_phyto_full$ESD_um - medphyto_climo
```

```{r}
#| include: false
### Linear fit 
# Note that this assumes that observations are equally spaced in time, which they're  not
n_obs <- seq(1, length(med_phyto_full$ESD_um), 1)
medphyto_lm <- lm(med_phyto_full$ESD_um ~ n_obs)
medphyto_anom_lm <- lm(med_phyto_anom_ts ~ n_obs)

# Change over time
delta_size_lm <- medphyto_lm$fitted.values[length(n_obs)] - medphyto_lm$fitted.values[1]
delta_size_pct = (medphyto_lm$fitted.values[length(n_obs)] - medphyto_lm$fitted.values[1]) / 
  medphyto_lm$fitted.values[1] * 100

delta_size_anom_lm <- medphyto_anom_lm$fitted.values[length(n_obs)] - 
  medphyto_anom_lm$fitted.values[1]
```

```{r}
#| include: false
### Annual values
med_phyto_full <- med_phyto_full %>% 
    mutate(Year = year(Date_Time), 
           Anom = med_phyto_anom_ts)  # Add in anomaly to make things easier down the road

ann_size <- med_phyto_full %>%
  group_by(Year) %>%
  summarise(ESD_um = mean(ESD_um, na.rm = TRUE))
ann_anom <- med_phyto_full %>%
  group_by(Year) %>%
  summarise(Anom = mean(Anom, na.rm = TRUE))
ann_mean_RptYr <- ann_size$ESD_um[which(ann_size$Year == RptYr)]
ann_anom_RptYr <- ann_anom$Anom[which(ann_size$Year == RptYr)]
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# This section includes some error checks to prompt fixing the text
if (any(summary(medphyto_lm)$coefficients[,4] > 0.05)) {
  print('The linear fit to monthly values is not signficant. Remove text related to the linear trend.')
}

if (any(summary(medphyto_anom_lm)$coefficients[,4] > 0.05)) {
  print('The linear fit to anomaly values is not signficant. Remove text related to the linear trend.')
}

# Pull out the MONTHLY values we need for the report & plots
yr_of_int <- which(med_phyto_full$Year == RptYr)
prev_yrs <- which(med_phyto_full$Year < RptYr)
all_yrs <- which(med_phyto_full$Year <= RptYr)
monthly_max_RptYr <- max(med_phyto_full$ESD_um[yr_of_int])
monthly_min_RptYr <- min(med_phyto_full$ESD_um[yr_of_int])
monthly_max_PrevYrs <- max(med_phyto_full$ESD_um[prev_yrs])
monthly_min_PrevYrs <- min(med_phyto_full$ESD_um[prev_yrs])

if (monthly_max_RptYr > monthly_max_PrevYrs) {
  print('The greatest monthly value was during the report year.  Revise text to reflect this.')
}

if (monthly_min_RptYr < monthly_min_PrevYrs) {
  print('The lowest monthly value was during the report year.  Revise text to reflect this.')
}
```

```{r}
#| include: false
# Write csv for portal
# Note that output csvs go in their own folder
# Rounding to avoid too many significant figures.
# Thanks to Johanna Wren for the code! https://github.com/pwoodworth-jefcoats/climate-indicators/issues/29
ann_size <- rename(ann_size, `ESD in um` = ESD_um)
ann_size$`ESD in um` <- format(round(ann_size$`ESD in um`, digits=2), nsmall = 2)
write_csv(ann_size, file = paste(here(), '/PelagicClimate_', RptYr, '/MedianPhytoSize_', RptYr, '.csv', sep = ""))

ann_anom <- rename(ann_anom, `ESD in um` = Anom)
ann_anom$`ESD in um` <- format(round(ann_anom$`ESD in um`, digits=2), nsmall = 2)
write_csv(ann_anom, file = paste(here(), '/PelagicClimate_', RptYr, '/MedianPhytoSizeAnomaly_', RptYr, '.csv', sep = ""))
```

```{r}
#| include: false
# Write csv for dashboard
# Note that dashboard output has its own folder

# Add columns for month, variable ID, and units
MD50_dashboard <- med_phyto_full %>% 
  mutate(Month=month(Date_Time), 
         ID='MD50', 
         Value_lm=medphyto_lm$fitted.values, 
         Anom_lm=medphyto_anom_lm$fitted.values, 
         Units='ESD in um', 
         Value=ESD_um) %>% 
  dplyr::select(Date_Time, Year, Month, Value, Anom, ID, Value_lm, Anom_lm, Units)

write.csv(MD50_dashboard, file = here("Indicator_Dashboard", "Data", paste('MD50_Dashboard_Data_', RptYr, '.csv', sep = "")), row.names = F)
```

```{r}
#| include: false
# Borrowing code from the dashboard for this chunk
# so that figures look the same across products
indicator_data <- MD50_dashboard |>
  filter(Year <= RptYr)

# Create color palette for easy reference 
oceans <- nmfs_palette("oceans")(3) # 1 = report_year, 3 = previous years
crustacean <- nmfs_palette("crustacean")(4) # 1 = linear trend
coral <- nmfs_palette("coral")(3) # 3 = annual average for Rpt Yr
ann_grey <- "#A5AAAF" # annual means; in NMFS branding guide as dove gray tint 1
waves <- nmfs_palette("waves")(3) # annual means; in NMFS branding guide but not in package
seagrass <- nmfs_palette("seagrass")(3)
pal <- c(oceans[3], coral[2], waves[2], coral[3], crustacean[2])

# # Formatting
# plot_title_font <- list(size = 14)
# plot_height <- 350 #in pixels

# Calculate annual means 
ann_vals <- indicator_data |>
  group_by(Year) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Identify the current year, to overlay on plot
given_yr <- indicator_data |>
  filter(Year == RptYr)

given_yr_ann <- bind_cols(Value = rep(ann_vals$Value[dim(ann_vals)[1]]),
                          Date_Time = given_yr$Date_Time,
                          Anom = as.numeric(ann_anom$`ESD in um`[which(ann_anom$Year == RptYr)]))

# Commented out the plotly code because it's not working this year.  So,
# going with base R, because that's what I'm most familiar with
# Fair warning, this coding is tedious.

# x axis ticks and grid
# Made a sequence instead so we don't have to manually update it every five years
x_grid <- c(min(indicator_data$Date_Time),
            seq.POSIXt(dmy_hm(010120000000), max(indicator_data$Date_Time), by='5 year'))
# x axis limit
ts_x_lim <- c(min(indicator_data$Date_Time), max(indicator_data$Date_Time))

# y-axis limits
ts_y_lim <-c(1.1, 1.9)
anom_y_lim <- c(-0.15, 0.15)

# Defining the theme up front so we don't have to repeat it
plotTheme <- theme_minimal() +
  theme(panel.border = element_rect(color='#F1F3F3'),
        panel.grid.major = element_line(color='#F1F3F3'),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        axis.title = element_text(size=14, face="bold"), 
        axis.text = element_text(size=12),
        legend.position = 'bottom', 
        legend.background = element_rect_round(fill = 'gray90', color='transparent', radius = 0.3),
        legend.text = element_text(size=12),legend.key.spacing.y = unit(-0.13,'cm'), legend.key.spacing.x = unit(1.5,'cm'))

# Plot using ggplot
legendInfo <- as.factor(c(paste0('Mothly average, 1985-', RptYr-1), 
                          paste0('Annual average, 1985-', RptYr-1), 
                          paste0('Mothly average, ', RptYr), 
                          paste0('Annual average, ', RptYr),
                          paste0('Long-term Trend')))
levels(legendInfo) <- levels(legendInfo)[c(4,5,1,2,3)] # To arrange in preferred order for the legend, otherwise it's alphabetical

p <- ggplot() +
  geom_path(data=indicator_data, aes(Date_Time, Value, color=legendInfo[2])) +
  geom_path(data=ann_vals, aes(dmy_hm(paste0('01-01-',Year,' 00:00')), Value, color=legendInfo[5])) +
  geom_path(data=indicator_data, aes(Date_Time, Value_lm, color=legendInfo[3])) +
  geom_path(data=given_yr, aes(Date_Time, Value, color=legendInfo[4])) +
  geom_path(data=given_yr_ann, aes(Date_Time, Value, color=legendInfo[1])) +
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = F) +
  scale_y_continuous(expand = F, limits = ts_y_lim, breaks=(seq(min(ts_y_lim),max(ts_y_lim),by=0.2))) +
  scale_color_manual(name=NULL, values=pal[c(1,3,5,2,4)]) +
  ylab('Median Phytoplankton Size (µm ESD)') +
  xlab('Year') +
  guides(color = guide_legend(ncol = 2)) +
  plotTheme

ggsave(paste0(Dir, '/Median_Phyto_Size_ts_', RptYr, '_ggolot.pdf'), plot=p, width=6, height=5, units='in')
ggsave(paste0(Dir, '/Median_Phyto_Size_ts_', RptYr, '_ggolot.png'), plot=p, width=6, height=5, units='in')

### Anomaly plot
# Calculate annual means
ann_vals <- indicator_data |>
  group_by(Year) |>
  summarise(Anom = mean(Anom, na.rm = TRUE)) 

p_ann <- ggplot() +
  geom_hline(yintercept = 0, linewidth=0.25,color='darkgray') +
  geom_path(data=indicator_data, aes(Date_Time, Anom, color=legendInfo[2])) +
  geom_path(data=ann_vals, aes(dmy_hm(paste0('01-01-',Year,' 00:00')), Anom, color=legendInfo[5])) +
  geom_path(data=indicator_data, aes(Date_Time, Anom_lm, color=legendInfo[3])) +
  geom_path(data=given_yr, aes(Date_Time, Anom, color=legendInfo[4])) +
  geom_path(data=given_yr_ann, aes(Date_Time, Anom, color=legendInfo[1])) +
  scale_x_datetime(breaks = x_grid, date_labels = '%Y', expand = F) +
  scale_y_continuous(expand = F, limits = anom_y_lim, breaks=c(-0.15,-0.1,-0.05,0,0.05,0.1,0.15), labels = c(-0.15,"-0.10",-0.05,"0",0.05,"0.10",0.15)) +
  scale_color_manual(name=NULL, values=pal[c(1,3,5,2,4)]) +
  ylab("Median Phytoplankton Size \nAnomaly (µm ESD)") +
  xlab('Year') +
  guides(color = guide_legend(ncol = 2)) +
  plotTheme

ggsave(paste0(Dir, '/Median_Phyto_Size_anom_ts_', RptYr, '_ggolot.pdf'), plot=p_ann, width=7, height=5, units='in')
ggsave(paste0(Dir, '/Median_Phyto_Size_anom_ts_', RptYr, '_ggolot.png'), plot=p_ann, width=7, height=5, units='in')


# pdf(paste(Dir, '/Median_Phyto_Size_ts_', RptYr, '.pdf', sep = ""), width=9, height=6)
# par(mar=c(8.6, 5.1, 2.1, 2.1))
# plot(indicator_data$Date_Time, indicator_data$Value, type = "l", lwd = 2, 
#      xlim = ts_x_lim, ylim = ts_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[1],
#      xlab = "Year", ylab = expression(bold("Median Phytoplankton Size (" * mu * "m ESD)")),
#      # put a grid behind the plot
#      panel.first = c(abline(v = x_grid, col = "#F1F3F3"),
#                      abline(h = c(seq(1.1, 1.9, 0.1)), col = "#F1F3F3")),
#      font.lab = 2, cex.lab=1.2)
# par(new = TRUE)
# plot(ymd_hm(paste(ann_vals$Year, '0601 00:00', sep = "")), ann_vals$Value, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = ts_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[3],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(indicator_data$Date_Time, indicator_data$Value_lm, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = ts_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[5],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(given_yr$Date_Time, given_yr$Value, type = "l", lwd = 2, 
#      xlim = ts_x_lim, ylim = ts_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[2],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(given_yr_ann$Date_Time, given_yr_ann$Value, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = ts_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[4],
#      xlab = " ", ylab = " ")
# axis(1, at = x_grid, label = c(1998, seq(2000, 2020, 5), 2025, " "), col = "transparent", col.ticks = "transparent") 
# axis(2, at = c(seq(1.1, 1.9, 0.1)), las = 1, col = "transparent", col.ticks = "transparent")
# grid.roundrect(x=0.51, y=0.075, width=0.6, height=0.12, 
#                r=unit(0.3, "snpc"), gp=gpar(fill="gray90", col='gray90'), just='centre')
# legend('bottom', 
#        legend=c(paste0('Monthly average, 1998-',RptYr-1), paste0('Annual average, 1998-',RptYr-1), 'Long-term Trend', paste0('Monthly average, ',RptYr), paste0('Annual average, ',RptYr)), 
#        col=c(pal[1], pal[3], pal[5], pal[2], pal[4]), 
#        lty=1, lwd=1.75, 
#        bg='transparent', 
#        horiz=F, ncol=2,
#        inset = -0.435, xpd=T, 
#        box.lwd = 0, box.col = 'transparent')
# dev.off()

# p1 <- plot_ly(indicator_data, x = dmy_hm(indicator_data$Date_Time), y = ~Value,
#               type = "scatter", mode = "lines", line = list(color = pal[1]),
#               name = ~ID[1], height = plot_height) |>
#   add_trace(indicator_data, x = dmy_hm(indicator_data$Date_Time), y = ~Value_lm,
#             type = "scatter", mode = "lines", line = list(color = pal[5]),
#             name = "Long-term Trend") |>
#   add_trace(ann_vals, x = ymd_hm(paste(ann_vals$Year, '0601 00:00', sep = "")), y = ann_vals$Value,
#             type = "scatter", mode = "lines", line = list(color = pal[3]),
#             name = "Annual Mean") |>
#   add_trace(given_yr, x = dmy_hm(given_yr$Date_Time), y = given_yr$Value,
#             type = "scatter", mode = "lines", line = list(color = pal[2]),
#             name = ~ID[1], height = plot_height) |>
#   add_trace(given_yr_ann, x = dmy_hm(given_yr_ann$Date_Time), y = given_yr_ann$Value,
#             type = "scatter", mode = "lines", line = list(color = pal[4]),
#             name = ~ID[1], height = plot_height)
# 
# #apply same layout parameters for all plots
# 
# #custom x axis (min, every decade, report year)
# all_years <- unique(indicator_data$Year)
# first_date <- as.character(parse_date_time(indicator_data$Date_Time[1], orders = "d-b-Y H:M"))
# date_axis <- c(first_date, 
#               all_years[which(all_years %% 10 == 0)], 
#               RptYr)
# 
# p1 <- p1 |> layout(title = list(text = "Indicator Time Series", x = 0.01, font = plot_title_font), #add title
#                    xaxis = list(type = "date", tickformat = "%Y", tickmode = "array", tickvals = date_axis, tickangle = 90),
#                    #xaxis = list(tick0 = min(indicator_data$Date_Time), dtick = "M24"),
#                    yaxis = list(title = indicator_data$Units[1], hoverformat = '.3f', range = list(1.1, 1.9), tickvals = list(1.1, 1.2, 1.3, 1.4, 1.5, 1.6, 1.7, 1.8, 1.9)), #add units and round values in hover display; not sure of a better way to set axis limits and ticks...
#                    paper_bgcolor = 'transparent', plot_bgcolor = 'transparent', #transparent background
#                    hovermode = "x unified", #show data for all traces simultaneously
#                    hoverlabel = list(namelength = -1)) #don't cutoff hoverinfo due to length
#   
# # return plot 
# save_image(p1, paste(Dir, '/Median_Phyto_Size_ts_', RptYr, '.pdf', sep = ""))


# ### Anomaly plot
# # Calculate annual means
# ann_vals <- indicator_data |>
#   group_by(Year) |>
#   summarise(Anom = mean(Anom, na.rm = TRUE)) 
# 
# pdf(paste(Dir, '/Median_Phyto_Size_anom_ts_', RptYr, '.pdf', sep = ""), width=9, height=6)
# par(mar=c(8.6, 5.1, 2.1, 2.1))
# plot(indicator_data$Date_Time, indicator_data$Anom, type = "l", lwd = 2, 
#      xlim = ts_x_lim, ylim = anom_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[1],
#      xlab = "Year", ylab = expression(bold("Median Phytoplankton Size Anomaly (" * mu * "m ESD)")),
#      # put a grid behind the plot
#      panel.first = c(abline(v = x_grid, col = "#F1F3F3"),
#                      abline(h = c(seq(-0.15, 0.15, 0.05)), col = "#F1F3F3"),
#                      abline(h = 0, col='black')),
#      font.lab = 2, cex.lab=1.2)
# par(new = TRUE)
# plot(ymd_hm(paste(ann_vals$Year, '0601 00:00', sep = "")), ann_vals$Anom, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = anom_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[3],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(indicator_data$Date_Time, indicator_data$Anom_lm, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = anom_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[5],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(given_yr$Date_Time, given_yr$Anom, type = "l", lwd = 2, 
#      xlim = ts_x_lim, ylim = anom_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[2],
#      xlab = " ", ylab = " ")
# par(new = TRUE)
# plot(given_yr_ann$Date_Time, given_yr_ann$Anom, type = "l", lwd = 2,
#      xlim = ts_x_lim, ylim = anom_y_lim, 
#      xaxt = "n", yaxt = "n", # no axes ticks, so we can customize them
#      xaxs = "i", yaxs = "i", # remove the padding R adds
#      bty = "n", # remove box
#      col = pal[4],
#      xlab = " ", ylab = " ")
# axis(1, at = x_grid, label = c(1998, seq(2000, 2020, 5), 2024, " "), col = "transparent", col.ticks = "transparent") 
# axis(2, at = c(seq(-0.15, -0.05, 0.05),0,seq(0.05,0.15,0.05)), labels=as.character(c(seq(-0.15, -0.05, 0.05),0,seq(0.05,0.15,0.05))), 
#      las = 1, col = "transparent", col.ticks = "transparent")
# grid.roundrect(x=0.51, y=0.075, width=0.6, height=0.12, 
#                r=unit(0.3, "snpc"), gp=gpar(fill="gray90", col='gray90'), just='centre')
# legend('bottom', 
#        legend=c(paste0('Monthly average, 1998-',RptYr-1), paste0('Annual average, 1998-',RptYr-1), 'Long-term Trend', paste0('Monthly average, ',RptYr), paste0('Annual average, ',RptYr)), 
#        col=c(pal[1], pal[3], pal[5], pal[2], pal[4]), 
#        lty=1, lwd=1.75, 
#        bg='transparent', 
#        horiz=F, ncol=2,
#        inset = -0.435, xpd=T, 
#        box.lwd = 0, box.col = 'transparent')
# dev.off()
#   
# p2 <- plot_ly(indicator_data, x = dmy_hm(indicator_data$Date_Time), y = ~Anom, height = plot_height,
#                 type = "scatter", mode = "lines", line = list(color = pal[1]),
#                 name = "Monthly Anomaly") |>
#     add_trace(indicator_data, x = dmy_hm(indicator_data$Date_Time), y = ~Anom_lm,
#               type = "scatter", mode = "lines", line = list(color = pal[5]),
#               name = "Long-term Trend") |>
#     add_trace(ann_vals, x = ymd_hm(paste(ann_vals$Year, '0601 00:00', sep = "")), y = ann_vals$Anom,
#               type = "scatter", mode = "lines", line = list(color = pal[3]),
#               name = "Annual Mean") |>
#   add_trace(given_yr, x = dmy_hm(given_yr$Date_Time), y = given_yr$Anom,
#             type = "scatter", mode = "lines", line = list(color = pal[2]),
#             name = ~ID[1], height = plot_height) |>
#   add_trace(given_yr_ann, x = dmy_hm(given_yr_ann$Date_Time), y = given_yr_ann$Anom,
#             type = "scatter", mode = "lines", line = list(color = pal[4]),
#             name = ~ID[1], height = plot_height) |>
#     layout(xaxis = list(type = "date", tickformat = "%Y", tickmode = "array", tickvals = date_axis, tickangle = 90),
#            yaxis = list(title = indicator_data$Units[1], hoverformat = '.3f', range = list(-0.15, 0.15), tickvals = list(-0.15, -0.1, -0.05, 0, 0.05, 0.1, 0.15)), #add units and round values in hover display
#            title = list(text = "Anomaly Time Series", x = 0.01, font = plot_title_font), #add title 
#            paper_bgcolor = 'transparent', plot_bgcolor = 'transparent', #transparent background
#            hovermode = "x unified", #show data for all traces simultaneously
#            hoverlabel = list(namelength = -1)) #don't cutoff hoverinfo due to length
#   
# # return plot
# save_image(p2, paste(Dir, '/Median_Phyto_Size_anom_ts_', RptYr, '.pdf', sep = ""))
```

**Rationale:** Phytoplankton are the base of the food web and their abundance influences the food available to all higher trophic levels from zooplankton through tuna and billfish. Some studies project that climate change will result in both fewer and smaller phytoplankton. This would reduce the food available to all members of the food web. Understanding trends in phytoplankton abundance and size structure, how they are influenced by oceanographic conditions, and how they influence fish abundance and size structure are areas of active research.  
\
**Status:** The mean monthly phytoplankton cell size was `r signif(ann_mean_RptYr,3)` $\mu$m Equivalent Spherical Diameter (ESD) in `r RptYr`. Monthly mean cell size ranged from `r signif(monthly_min_RptYr,3)`–`r signif(monthly_max_RptYr,3)` $\mu$m ESD during the year, within the range of values observed over the period of record (`r signif(monthly_min_PrevYrs,3)`–`r signif(monthly_max_PrevYrs,3)` $\mu$m ESD). Over the period of record, there has been weakly significant decline in monthly median phytoplankton size. Over the time series, median phytoplankton size has declined by `r abs(signif(delta_size_lm,2))` $\mu$m ESD, or by `r abs(signif(delta_size_pct,2))`%. The monthly anomaly has declined as well, by `r abs(signif(delta_size_anom_lm,2))` $\mu$m ESD. Average estimated median phytoplankton size was below average across much of the fishing grounds, particularly north of about 30°N.    
\
**Description:** Median phytoplankton cell size can be estimated from satellite remotely sensed SST and ocean color (Barnes et al. 2011). A time series of monthly median phytoplankton cell size averaged over the Hawaiʻi longline region is presented, as well as a time series of anomalies. NOAA CoralTemp (see SST indicator) and ESA CCI data (see ocean color indicator) are used to calculate median phytoplankton cell size.  
\
**Timeframe:** Monthly  
\
**Region:** Hawaiʻi-based longline region: 15° – 45°N, 180° – 120°W    
\
**Measurement Platform:** Satellite  
\
**Data available at:** <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/md50_exp_2025>, <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/md50_exp-1998-2009-clim>, and <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/md50_exp-2025-clim>.  
\
**Sourced from:** Barnes et al. (2011) and NOAA OceanWatch (2026c). Graphics produced in part using Stawitz and Li (2025).  
\
## Additional Information
Median phytoplankton size data are geographically subset and spatially averaged using the pyFerret script `MedPhytoSize_access.jnl` which can be found in the [Median_Phytoplankton_Size folder](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Median_Phytoplankton_Size).  You could do this step in R, but I use [PyFerret](https://ferret.pmel.noaa.gov/Ferret/) because it's freely available software developed by NOAA specifically for working with large gridded datasets.  
\
A plot of the residuals from the linear model showed that they were evenly distributed, although were more negative (~-0.39 min) than positive (~0.35 max). The residuals for the anomaly model were also fairly evenly distributed between -0.1 and 0.1.  
\
To prepare the spatial data for mapping, you'll need to run `map_data_for_dashboard.R`, which can be found in the [PreProcessing](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Indicator_Dashboard/PreProcessing) folder of the [Indicator_Dashboard](https://github.com/pwoodworth-jefcoats/climate-indicators/tree/main/Indicator_Dashboard) folder in this repository.

```{r}
#| include: FALSE
# # After running the chunks above and prepping the map data, you can uncomment and run this chunk.
# 
# # Load basemap data
# land_proj <- readRDS(here('Indicator_Dashboard', 'Data', 'rnatearth_land.RData'))
# coast_proj <- readRDS(here('Indicator_Dashboard', 'Data', 'rnatearth_coast.RData'))
# 
# Load data
maps <- read.csv(paste(Dir, '/Median_Phyto_map_data_', RptYr, '.csv', sep = ""))

# Filter to the given year and the anomaly
maps_RptYr <- maps |> filter(ID == "MD50")
maps_anom <- maps |> filter(ID == "MD50_anom")

# Adding a column that assignes line types to contours so we can make a legend
maps_anom$ltyF <- factor(ifelse(maps_anom$layer > 0.1, 'Larger', ifelse(maps_anom$layer < -0.1, 'Smaller', 'No Change')), levels=c('Smaller', 'Larger', 'No Change'))

# Map elements and aesthetics
seagrass <- nmfs_palette("seagrass")(3)
pal <- seagrass
ll_rect_color <- "white" #outline for ll fishing grounds box
fill_scale <- scale_fill_gradientn(name = "MD50", colors = pal, limits = c(0, 8.25))
 
# Create map
p_map <- ggplot() +
 geom_raster(data = maps_RptYr, mapping = aes(x = x_disp, y = y, fill = layer)) + #data as raster
 borders('world2', fill='#A5AAAF', colour='gray40', linewidth=0.2) + #add land to the map
 # geom_rect(aes(xmin = 0, xmax = 60, ymin = 15, ymax = 45), color = ll_rect_color, fill = NA, linewidth = 0.5)  + #ll fishing grounds box
 # annotate("text", x = 30, y = 47, label = "Longline fishing grounds", size = 3.2, color = ll_rect_color) +
 fill_scale + #indicator-dependent color and scale from above
coord_quickmap(expand = F, xlim=c(120,260), ylim=c(0,55)) + #don't expand past x/y limits
 scale_x_continuous(breaks=seq(120,260,30), labels = paste0(c(seq(120,180,30), seq(150,100,-30)), c(rep('°E',2),"",rep('°W',2)))) + # adding degrees to the breaks
 scale_y_continuous(breaks=seq(0,55,15), labels = paste0(seq(0,55,15), c('',rep('°N',3)))) + # adding degrees to the breaks
 xlab("") + ylab("") +
 theme_bw() +
 theme(panel.grid = element_line(color = "black", linewidth = 0.1),
      legend.background = element_rect(fill = 'transparent'), legend.key.height = unit(1, "null"),
      legend.key.width = unit(0.5, 'cm'),  #scales color bar to height of the plot
      legend.margin = margin(0, 0, 0, 0),  #fixing margins on color bar
      legend.text = element_text(size = 10),  #text size for color bar
      plot.title = element_text(hjust = 0.5)) + #centering the plot title. Default is left align
 guides(fill = guide_colorbar(title.position='right', title="Median Phytoplankton Size (ESD in µm)")) + #rotates the color bar text and centers it
 labs(title = paste(RptYr, 'average (shaded) and anomaly (contours)'))  #add figure title

# Add anomaly
p_map <- p_map +
  geom_contour(data = maps_anom, aes(x = x_disp, y = y, z = layer, linetype=ltyF, linewidth=ltyF),
               breaks = seq(floor(min(maps_anom$layer)), ceiling(max(maps_anom$layer)), 0.1), color='black') +
  geom_rect(aes(xmin = 180, xmax = 240, ymin = 15, ymax = 45), color = ll_rect_color, fill = NA, linewidth = 0.5)  + #ll fishing grounds box
  annotate("text", x = 210, y = 48, label = "Longline fishing grounds", size = 3.7, color = ll_rect_color) +
  scale_linetype_manual(values = c('dashed','solid','solid'), name='Difference from 1998-2009 average in 0.1 µm intervals') +  #adding legend to bottom of plot. Keep name the same for legends you want to combine (here it's linetype and linewidth)
  scale_linewidth_manual(values=c(0.25,0.25,1), name='Difference from 1998-2009 average in 0.1 µm intervals') +
  guides(fill = guide_colorbar(title.position='right', title="Median Phytoplankton Size (ESD in µm)",
                               title.theme=element_text(angle=270, hjust=0.5, vjust=0.5)),
         linetype=guide_legend(position = 'bottom', title.position='top', title.theme=element_text(hjust=0.5, vjust=1)))  #this bit of code gives different aesthetics to the colorbar on the left and legend on the bottom.

ggsave(filename = paste0(Dir, '/Median_Phyto_Size_map_', RptYr, '.pdf'), plot=p_map, width = 7, height = 3.8)
ggsave(filename = paste0(Dir, '/Median_Phyto_Size_map_', RptYr, '.png'), plot=p_map, width = 7, height = 3.8)

# pdf(paste(Dir, '/Median_Phyto_Size_map_', RptYr, '.pdf', sep = ""))
# print(p)
# dev.off()

# # Add anomaly
# p <- p +
#   geom_contour(data = maps_anom, mapping = aes(x = x, y = y, z = layer),
#                breaks = c(seq(floor(min(maps_anom$layer)), -0.1, 0.1)),
#                color = "black", linetype = 3) # Dotted negative contours
# p <- p +
#   geom_contour(data = maps_anom, mapping = aes(x = x, y = y, z = layer), breaks = c(0),
#                color = "black", lwd = 1) # heavy zero line
# p <- p +
#   geom_contour(data = maps_anom, mapping = aes(x = x, y = y, z = layer),
#                breaks = c(seq(0.1, ceiling(max(maps_anom$layer)), 0.1)),
#                color = "black", lwd = 0.5) # solid positive contours
# 
# p
# 
```
