---
title: "Bigeye_Forecast"
author: "phoebe.woodworth-jefcoats@noaa.gov"
format:
  docx:
    reference-doc: SAFE-Reference-Doc.docx
---

## Bigeye Forecast

```{r}
#| include: false
### Load libraries
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
# remotes::install_github("nmfs-fish-tools/nmfspalette")
library(nmfspalette)
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2023

# Set path to variable: Bigeye_Forecast
# This is where the data are and where the plots will go
Dir <- here("Bigeye_Forecast")
```

```{r}
#| include: false
### Load data
# Annual estimated median phytoplankton size over forecast area
med_phyto_fct_9802 <- read_delim(file = paste(Dir,  '/MedPhyto_Bigeye_Forecast_ts_9802.dat', sep = ""), skip = 5, col_names = FALSE)
med_phyto_fct_0307 <- read_delim(file = paste(Dir,  '/MedPhyto_Bigeye_Forecast_ts_0307.dat', sep = ""), skip = 5, col_names = FALSE)
med_phyto_fct_0812 <- read_delim(file = paste(Dir,  '/MedPhyto_Bigeye_Forecast_ts_0812.dat', sep = ""), skip = 5, col_names = FALSE)
med_phyto_fct_1317 <- read_delim(file = paste(Dir,  '/MedPhyto_Bigeye_Forecast_ts_1317.dat', sep = ""), skip = 5, col_names = FALSE)
med_phyto_fct_1823 <- read_delim(file = paste(Dir,  '/MedPhyto_Bigeye_Forecast_ts_1823.dat', sep = ""), skip = 5, col_names = FALSE)

# Concatonate
med_phyto_fct_full <- rbind(med_phyto_fct_9802, 
                        med_phyto_fct_0307, 
                        med_phyto_fct_0812, 
                        med_phyto_fct_1317, 
                        med_phyto_fct_1823)

# CPUE and WPUE data
Fishery_data <- read_csv(file = paste(Dir, '/CPUEWPUE_', RptYr, '.csv', sep = ""))
```

```{r}
#| include: false
### Annual values of median phytoplankton size
yrs <- year(dmy_hm(med_phyto_fct_full$X1))
med_phyto_fct_full <- bind_cols(med_phyto_fct_full, yrs)
med_phyto_fct_full <- rename(med_phyto_fct_full, Date_Time = X1)
med_phyto_fct_full <- rename(med_phyto_fct_full, ESD_um = X2)
med_phyto_fct_full <- rename(med_phyto_fct_full, Year = ...3)
```

```{r}
#| include: false
### Linear fit 
phyto_years <- which(med_phyto_fct_full$Year >= 1998 & med_phyto_fct_full$Year <= RptYr - 4)
bigeye_years <- which(Fishery_data$LANDYR >= 2002 & Fishery_data$LANDYR <= RptYr)

# Build data frames
CPUE_model_data <- bind_cols(med_phyto_fct_full$ESD_um[phyto_years], Fishery_data$CPUEall[bigeye_years])
CPUE_model_data <- rename(CPUE_model_data, ESD_um = ...1)
CPUE_model_data <- rename(CPUE_model_data, CPUE = ...2)

WPUE_model_data <- bind_cols(med_phyto_fct_full$ESD_um[phyto_years], Fishery_data$WPUE[bigeye_years])
WPUE_model_data <- rename(WPUE_model_data, ESD_um = ...1)
WPUE_model_data <- rename(WPUE_model_data, WPUE = ...2)

CPUE_forecast_model <- lm(CPUE ~ ESD_um, data = CPUE_model_data)
WPUE_forecast_model <- lm(WPUE ~ ESD_um, data = WPUE_model_data)
```

```{r}
#| echo: false
# Note that the above needs to be 'echo' and not 'include' so that the error checks print.

# This section includes some error checks to prompt fixing the text
if (any(summary(CPUE_forecast_model)$coefficients[,4] > 0.05)) {
  print('Double-check the CPUE forecast.')
}

if (any(summary(WPUE_forecast_model)$coefficients[,4] > 0.05)) {
  print('Double-check the WPUE forecast.')
}
```

```{r}
# Forecasts & confidence intervals
# note to self, compare the confidence intervals to Matlab approach
Years <- seq(2002, RptYr + 4, 1)
CPUE_forecast <- predict(CPUE_forecast_model, data.frame(ESD_um = med_phyto_fct_full$ESD_um), interval = "confidence", level = 0.95)
WPUE_forecast <- predict(WPUE_forecast_model, data.frame(ESD_um = med_phyto_fct_full$ESD_um), interval = "confidence", level = 0.95)

Bigeye_Forecast <- as_tibble(cbind(Years, CPUE_forecast[,1], WPUE_forecast[,1]))
Bigeye_Forecast <- rename(Bigeye_Forecast, CPUEforecast = V2)
Bigeye_Forecast <- rename(Bigeye_Forecast, WPUEforecast = V3)
```

```{r}
#| include: false
### Create plot for report
# Note that there's some final work that happens in Illustrator, namely, making the shading work

# Create color palette for easy reference
waves <- nmfs_palette("waves")(4) # 3 = CPUE
urchin <- nmfs_palette("urchin")(5) # 4 = WPUE

pdf(paste(Dir, '/Bigeye_Forecast_CPUE_', RptYr, '.pdf', sep = ""),
    width = 5.5, height = 3.5)
plot(Bigeye_Forecast$Years, Bigeye_Forecast$CPUEforecast, type = "l", lty = 3, col = waves[3],
     xlim = c(2002, RptYr + 4), ylim = c(2, 5), xlab = "", ylab = "",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(Fishery_data$LANDYR[bigeye_years], CPUE_model_data$CPUE, type = "l", lwd = 2, col = waves[3],
     xlim = c(2002, RptYr + 4), ylim = c(2, 5), xlab = "", ylab = "",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(Bigeye_Forecast$Years, CPUE_forecast[,2], type = "l", lwd = 0.5, col = waves[3],
     xlim = c(2002, RptYr + 4), ylim = c(2, 5), xlab = "", ylab = "",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(Bigeye_Forecast$Years, CPUE_forecast[,3], type = "l", lwd = 0.5, col = waves[3],
     xlim = c(2002, RptYr + 4), ylim = c(2, 5), xlab = "", ylab = "",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
axis((1), at = seq(2002, RptYr + 4, 1))
axis((2), at = seq(2, 5, 0.5))
dev.off()

pdf(paste(Dir, '/Bigeye_Forecast_WPUE_', RptYr, '.pdf', sep = ""),
    width = 5.5, height = 3.5)
plot(Bigeye_Forecast$Years, Bigeye_Forecast$WPUEforecast, type = "l", lty = 3, col = urchin[4],
     xlim = c(2002, RptYr + 4), ylim = c(75, 175), xlab = "", ylab = "",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(Fishery_data$LANDYR[bigeye_years], WPUE_model_data$WPUE, type = "l", lwd = 2, col = urchin[4],
     xlim = c(2002, RptYr + 4), ylim = c(75, 175), xlab = "", ylab = "",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(Bigeye_Forecast$Years, WPUE_forecast[,2], type = "l", lwd = 0.5, col = urchin[4],
     xlim = c(2002, RptYr + 4), ylim = c(75, 175), xlab = "", ylab = "",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(Bigeye_Forecast$Years, WPUE_forecast[,3], type = "l", lwd = 0.5, col = urchin[4],
     xlim = c(2002, RptYr + 4), ylim = c(75, 175), xlab = "", ylab = "",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
axis((1), at = seq(2002, RptYr + 4, 1))
axis((2), at = seq(75, 175, 25))
dev.off()
```

```{r}
#| include: false
### Create plot for report
# Create a vector that repeats the annual mean the same number of times as there are obs for the latest year
# (although there are other ways to plot this piece)
ann_mean_x12 <- rep(ann_mean_RptYr,length(yr_of_int))
ann_anom_x12 <- rep(ann_anom_RptYr,length(yr_of_int))

# Create axes limit values, to make things simpler
sz_xlim <- c(min(dmy_hm(med_phyto_full$Date_Time)), max(dmy_hm(med_phyto_full$Date_Time)))
ts_ylim <- c(1.1, 1.9)
anom_ylim <- c(-0.15, 0.15)

# Create color palette for easy reference 
oceans <- nmfspalette::nmfs_palette("oceans")(3) # 1 = RptYr, 3 = previous years
crustacean <- nmfspalette::nmfs_palette("crustacean")(4) # 1 = linear trend
coral <- nmfspalette::nmfs_palette("coral")(3) # 3 = annual average for Rpt Yr
ann_grey <- "#D0D0D0" # annual means; in NMFS branding guide but not in package
pal <- c(oceans[3], oceans[1], ann_grey, coral[3], crustacean[1])

# Plot
# pdf(paste(Dir, '/Median_Phyto_size_ts_', RptYr, '.pdf', sep = ""), width = 7, height = 4.5)
plot(ymd_hm(paste(ann_size$Year, '0601 00:00', sep = "")), ann_size$ESD_um, type = "l", lwd = 2, col = ann_grey,
     xlim = sz_xlim, ylim = ts_ylim, xlab = "Year", ylab = "Median Phytoplankton Size (ESD in um)",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(dmy_hm(med_phyto_full$Date_Time[all_yrs]), medphyto_lm$fitted.values, type = "l", lwd = 2, col = crustacean[1], 
     xlim = sz_xlim, ylim = ts_ylim, xlab = " ", ylab = " ",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(dmy_hm(med_phyto_full$Date_Time[all_yrs]), med_phyto_full$ESD_um[all_yrs], type = "l", lwd = 2, col = oceans[3], 
     xlim = sz_xlim, ylim = ts_ylim, xlab = " ", ylab = " ",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i") 
par(new = TRUE)
plot(dmy_hm(med_phyto_full$Date_Time[yr_of_int]), med_phyto_full$ESD_um[yr_of_int], type = "l", lwd = 2, col = oceans[1], 
     xlim = sz_xlim, ylim = ts_ylim, xlab = " ", ylab = " ", 
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(dmy_hm(med_phyto_full$Date_Time[yr_of_int]), ann_mean_x12, type = "l", lwd = 2, col = coral[3], 
     xlim = sz_xlim, ylim = ts_ylim, xlab = " ", ylab = " ", 
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
axis((1), at = ymd_hm(paste(ann_size$Year, '0101 00:00', sep = "")), tck = 0.025, labels = year(make_date(seq(1998, 2022, 1))))
axis((2), at = seq(1.1, 1.9, 0.1), tck = 0.025, las = 1)
axis((3), at = ymd_hm(paste(ann_size$Year, '0101 00:00', sep = "")), tck = 0.025, labels = FALSE)
axis((4), at = seq(1.1, 1.9, 0.1), tck = 0.025, labels = FALSE)
legend(x = "bottomright", 
       legend = c(paste("Monthly average, 1998 - ", RptYr - 1, sep = ""), 
                  paste("Monthly average, ", RptYr, sep = ""),
                  paste("Annual average, 1998 - ", RptYr - 1, sep = ""), 
                  paste("Annual average, ", RptYr, sep = ""),
                  "Long-term trend"), 
       lty = 1, lwd = 2,
       col = pal, 
       bty = "n")
# dev.off()

# pdf(paste(Dir, '/Median_Phyto_size_anom_ts_', RptYr, '.pdf', sep = ""), width = 7, height = 4.5)
plot(ymd_hm(paste(ann_anom$Year, '0601 00:00', sep = "")), ann_anom$Anom, type = "l", lwd = 2, col = ann_grey,
     xlim = sz_xlim, ylim = anom_ylim, xlab = "Year", ylab = "Median Phytoplankton Size Anomaly (ESD in um)",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(dmy_hm(med_phyto_full$Date_Time[all_yrs]), medphyto_anom_lm$fitted.values, type = "l", lwd = 2, col = crustacean[1], 
     xlim = sz_xlim, ylim = anom_ylim, xlab = " ", ylab = " ",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(dmy_hm(med_phyto_full$Date_Time[all_yrs]), med_phyto_full$Anom[all_yrs], type = "l", lwd = 2, col = oceans[3], 
     xlim = sz_xlim, ylim = anom_ylim, xlab = " ", ylab = " ",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i") 
par(new = TRUE)
plot(dmy_hm(med_phyto_full$Date_Time[yr_of_int]), med_phyto_full$Anom[yr_of_int], type = "l", lwd = 2, col = oceans[1], 
     xlim = sz_xlim, ylim = anom_ylim, xlab = " ", ylab = " ", 
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(dmy_hm(med_phyto_full$Date_Time[yr_of_int]), ann_anom_x12, type = "l", lwd = 2, col = coral[3], 
     xlim = sz_xlim, ylim = anom_ylim, xlab = " ", ylab = " ", 
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
axis((1), at = ymd_hm(paste(ann_size$Year, '0101 00:00', sep = "")), tck = 0.025, labels = year(make_date(seq(1998, 2022, 1))))
axis((2), at = seq(-0.15, 0.15, 0.05), tck = 0.025, las = 1)
axis((3), at = ymd_hm(paste(ann_size$Year, '0101 00:00', sep = "")), tck = 0.025, labels = FALSE)
axis((4), at = seq(-0.15, 0.15, 0.05), tck = 0.025, labels = FALSE)
legend(x = "bottomright", 
       legend = c(paste("Monthly average, 1998 - ", RptYr - 1, sep = ""), 
                  paste("Monthly average, ", RptYr, sep = ""),
                  paste("Annual average, 1998 - ", RptYr - 1, sep = ""), 
                  paste("Annual average, ", RptYr, sep = ""),
                  "Long-term trend"), 
       lty = 1, lwd = 2,
       col = pal, 
       bty = "n")
# dev.off()
# _axt = "n" removes tick labels, _axs = "i" removes whitespace beyond axes maxima
# bty = "n" removes the box around the legend
```

```{r}
#| include: false
# Write csv for portal
# Note that output csvs go in their own folder
Bigeye_Forecast <- rename(Bigeye_Forecast, `Fish per 1000 hooks (bigeye)` = CPUEforecast)
Bigeye_Forecast <- rename(Bigeye_Forecast, `kg per 1000 hooks (bigeye)` = WPUEforecast)
write_csv(Bigeye_Forecast, file = paste(here(), '/PelagicClimate_', RptYr, '/BigeyeForecast_', RptYr, '.csv', sep = ""))
```

Rationale: Phytoplankton are the base of the food web and their abundance influences the food available to all higher trophic levels from zooplankton through tuna and billfish. Some studies project that climate change will result in both fewer and smaller phytoplankton. This would reduce the food available to all members of the food web. Understanding trends in phytoplankton abundance and size structure, how they are influenced by oceanographic conditions, and how they influence fish abundance and size structure are areas of active research.  
\
Status: The mean monthly phytoplankton cell size was `r signif(ann_mean_RptYr,3)` $\mu$m Equivalent Spherical Diameter (ESD) in `r RptYr`. Monthly mean cell size ranged from `r signif(monthly_min_RptYr,3)`–`r signif(monthly_max_RptYr,3)` $\mu$m ESD during the year, within the range of values observed over the period of record (`r signif(monthly_min_PrevYrs,3)`–`r signif(monthly_max_PrevYrs,3)` $\mu$m ESD). Over the period of record, there has been weakly significant decline in monthly median phytoplankton size. Over the time series, median phytoplankton size has declined by `r signif(delta_size_lm,2)` $\mu$m ESD, or by `r signif(delta_size_pct,2)`%. The monthly anomaly has declined as well, by `r signif(delta_size_anom_lm,2)` $\mu$m ESD. Average estimated median phytoplankton size was below average across much of the fishing grounds.  
\
Description: Median phytoplankton cell size can be estimated from satellite remotely sensed SST and ocean color (Barnes et al. 2011). A time series of monthly median phytoplankton cell size averaged over the Hawaiʻi longline region is presented, as well as a time series of anomalies. NOAA CoralTemp (see SST indicator) and ESA CCI data (see ocean color indicator) are used to calculate median phytoplankton cell size.  
\
Timeframe: Annual  
\
Region: Hawaiʻi-based longline fishing grounds (0 – 40N, 180 – 150W and 15 – 36N, 150 – 125W)
\
Measurement Platform: Satellite  
\
Data available at: <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/md50_exp>  
\
Sourced from: HDAR, Langley et al. (2006), Woodworth-Jefcoats and Wren (2020), Barnes et al. (2011), and NOAA OceanWatch (2024c). Graphics produced in part using Stawitz (2023).  
\
## Additional Information  
Median phytoplankton size data are geographically subset and spatially averaged using the pyFerret script `BigeyeForecast_access.jnl` which can be found in the Bigeye_Forecast folder.  You could do this step in R, but I use [PyFerret](https://ferret.pmel.noaa.gov/Ferret/) because it's freely available software developed by NOAA specifically for working with large gridded datasets.  Time series of bigeye tuna CPUE and WPUE and created using `SAFE_LogDeal_DataDownload.Rmd` in the Fisheries_Indicators folder.  Note that actually running this script requires access to a controlled database.
\
A plot of the residuals from the linear model showed that they were evenly distributed, although were more negative (~-0.39 min) than positive (~0.35 max). The residuals for the anomaly model were also fairly evenly distributed.