---
title: "CO2_description"
author: "Phoebe.Woodworth-Jefcoats@noaa.gov"
date: "2023-10-16"
output: md_document
preserve_yaml: FALSE
---

```{r}
#| include: false
### Load libraries
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
library(markdown)
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2023

# Identify the variable we're focused on
var <- 'CO2'
```

```{r, include = FALSE, global = TRUE}
# Access indicator data
indicator_data <- read_csv(here('Indicator_Dashboard','Data', paste('Dashboard_Data_', RptYr, '.csv', sep = "")))

# And just use our variable of interest
indicator_data <- indicator_data |> filter(ID == var)
```

```{r, include = FALSE}
# Values that are included in the text

# Annual mean for report year
ann_mean_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Increase over time from linear model
if (!is.na(indicator_data$Value_lm[1])) {
  delta_CO2_lm <- indicator_data$Value_lm[length(indicator_data$Value_lm)] - 
    indicator_data$Value_lm[1]
}

# Increase in anomaly over time from linear model
if (!is.na(indicator_data$Anom_lm[1])) {
  delta_CO2_anom_lm <- indicator_data$Anom_lm[length(indicator_data$Anom_lm)] - 
    indicator_data$Anom_lm[1]
}

# Monthly min & max for report year
monthly_min_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))

# Monthly min & max for previous years
monthly_min_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))
```

## Atmospheric Carbon Dioxide
__About the Indicator__  
The primary driver of anthropogenic (human-caused) climate change is the increasing concentration of atmospheric carbon dioxide, CO~2~, due to the burning of fossil fuels. Therefore, atmospheric CO~2~ serves as a measure of what human activity has already done to affect the climate system through greenhouse gas emissions. It provides quantitative information in a simplified, standardized format that decision makers can easily understand.  The concentration of atmospheric CO~2~, and, in turn, its warming influence, is increasing more quickly over time. 

Increasing CO~2~ concentrations cause the atmosphere to warm. Much of this heat is then absorbed by the ocean, raising ocean temperatures.  The ocean also becomes more acidic as it absorbs CO~2~. 

__Status of the Indicator__  
Atmospheric CO~2~ is increasing exponentially. This means that atmospheric CO~2~ is increasing more quickly over time. In `r RptYr`, the annual mean concentration of CO~2~ was `r signif(ann_mean_RptYr,5)` ppm.  This is the highest annual value recorded. This year also saw the highest monthly value, which was `r monthly_max_RptYr` ppm. In 1959, the first year full of the time series, the atmospheric concentration of CO~2~ was 316 ppm. The annual mean passed 350 ppm in 1988, and 400 ppm in 2015.  

__Indicator Data__  
Description: Monthly mean atmospheric carbon dioxide (CO~2~) at Mauna Loa Observatory, Hawaiʻi in parts per million (ppm) from March 1958 to present. The observed increase in monthly average carbon dioxide concentration is primarily due to CO~2~ emissions from fossil fuel burning. Carbon dioxide remains in the atmosphere for a very long time, and emissions from any location mix throughout the atmosphere in approximately one year. The annual variations at Mauna Loa, Hawaiʻi are due to the seasonal imbalance between the photosynthesis and respiration of terrestrial plants. During the summer growing season, photosynthesis exceeds respiration, and CO~2~ is removed from the atmosphere. In the winter (outside the growing season), respiration exceeds photosynthesis, and CO~2~ is returned to the atmosphere. The seasonal cycle is strongest in the northern hemisphere because of its larger land mass. Note that due to the eruption of the Mauna Loa Volcano, measurements from Mauna Loa Observatory were suspended as of 29 November 2022. Observations from December 2022 to 4 July 2023 are from a site at the Maunakea Observatories, approximately 21 miles north of the Mauna Loa Observatory.  Mauna Loa observations resumed in July 2023.  

Timeframe: Annual, monthly.  

Region/Location: Mauna Loa, Hawaiʻi, but representative of global atmospheric carbon dioxide concentration.  

Measurement Platform: *In-situ* station.  

Data available at: <https://gml.noaa.gov/ccgg/trends/data.html>.   

Sourced from: Keeling et al. (1976), Thoning et al. (1989), and NOAA (2024a).  Graphics produced in part using Stawitz (2023).

__References__  
Keeling et al. (1976): C.D. Keeling, R.B. Bacastow, A.E. Bainbridge, C.A. Ekdahl, P.R. Guenther, and L.S. Waterman, (1976), Atmospheric carbon dioxide variations at Mauna Loa Observatory, Hawaii, *Tellus*, vol. 28, 538-551  

NOAA (2024a): Dr. Pieter Tans, NOAA/GML (gml.noaa.gov/ccgg/trends/) and Dr. Ralph Keeling, Scripps Institution of Oceanography (scrippsco2.ucsd.edu/), Trends in Atmospheric Carbon Dioxide <https://gml.noaa.gov/ccgg/trends/data.html>, Accessed: 13 March 2024.  

Thoning et al. (1989): K.W. Thoning, P.P. Tans, and W.D. Komhyr, (1989), [Atmospheric carbon dioxide at Mauna Loa Observatory 2. Analysis of the NOAA GMCC data, 1974-1985](https://gml.noaa.gov/publications/showpub.php?pubid=2828), J. Geophys. Research, vol. 94, 8549-8565  

Stawitz C (2023): nmfspalette: A Color Palette for NOAA Fisheries. R package version 3.0.0.000. <https://nmfs-fish-tools.github.io/nmfspalette/>