---
title: "ENSO_description"
author: "Phoebe.Woodworth-Jefcoats@noaa.gov"
date: "2023-12-06"
output: md_document
preserve_yaml: FALSE
---

```{r}
#| include: false
### Load libraries
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
library(markdown)
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2023

# Identify the variable we're focused on
var <- 'ENSO'
```

```{r, include = FALSE, global = TRUE}
# Access indicator data
indicator_data <- read_csv(here('Indicator_Dashboard','Data', paste('Dashboard_Data_', RptYr, '.csv', sep = "")))

# And just use our variable of interest
indicator_data <- indicator_data |> filter(ID == var)
```

```{r, include = FALSE}
# Values that are included in the text

# Annual mean for report year
ann_mean_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Increase over time from linear model
if (!is.na(indicator_data$Value_lm[1])) {
  delta_SST_lm <- indicator_data$Value_lm[length(indicator_data$Value_lm)] - 
    indicator_data$Value_lm[1]
}

# Increase in anomaly over time from linear model
if (!is.na(indicator_data$Anom_lm[1])) {
  delta_SST_anom_lm <- indicator_data$Anom_lm[length(indicator_data$Anom_lm)] - 
    indicator_data$Anom_lm[1]
}

# Monthly min & max for report year
monthly_min_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))

# Monthly min & max for previous years
monthly_min_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))
```

## El Niño – Southern Oscillation
__About the Indicator__  
The El Niño – Southern Oscillation (ENSO) is one of the most prominent modes of natural climate variability in the North Pacific.  ENSO cycles, which typically last 6 - 18 months, are known to impact Pacific fisheries, including tuna fisheries, because they cause interannual changes in sea surface temperature and phytoplankton abundance.

The Oceanic Niño Index (ONI) is a measure of ENSO phase that focuses on ocean temperature, which has a direct effect on these fisheries.  

Understanding the effects of natural climate variability like ENSO on the ocean, marine ecosystems, and fisheries is an active area of research.

__Status of the Indicator__  
The Oceanic Niño Index (ONI) indicated a transition from La Niña to El Niño conditions in `r RptYr`. In `r RptYr`, the ONI ranged from `r monthly_min_RptYr` to `r monthly_max_RptYr`.  This is within the range of values observed previously in the time series.

__Indicator Data__  
Description: The three-month running mean (referred to as a season) of satellite remotely-sensed sea surface temperature (SST) anomalies in the Niño 3.4 region (5°S – 5°N, 120° – 170°W). The Oceanic Niño Index (ONI) is a measure of the El Niño – Southern Oscillation (ENSO) phase. Warm and cool phases, termed El Niño and La Niña respectively, are based in part on an ONI threshold of ± 0.5 °C being met for a minimum of five consecutive overlapping seasons. Additional atmospheric indices are needed to confirm an El Niño or La Niña event, as the ENSO is a coupled ocean-atmosphere phenomenon. The atmospheric half of ENSO is measured using the Southern Oscillation Index.

Timeframe: Three-month running average.

Region/Location: Niño 3.4 region, 5°S – 5°N, 120° – 170°W.

Measurement Platform: *In-situ* station, satellite, model.

Data available at:  <https://www.cpc.ncep.noaa.gov/data/indices/oni.ascii.txt>. 

Sourced from NOAA CPC (2024). Graphics produced in part using Stawitz (2023).

__References__  
NOAA CPC (2024): <https://www.cpc.ncep.noaa.gov/data/indices/oni.ascii.txt>, Accessed: 19 March 2024. 

nmfspalette: Stawitz C (2023): nmfspalette: A Color Palette for NOAA Fisheries. R package version 3.0.0.000. <https://nmfs-fish-tools.github.io/nmfspalette/>