---
title: "MD50_description"
author: "Phoebe.Woodworth-Jefcoats@noaa.gov"
date: "2023-12-06"
output: md_document
preserve_yaml: FALSE
---

```{r}
#| include: false
### Load libraries
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
library(markdown)
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2025

# Identify the variable we're focused on
var <- 'MD50'
```

```{r, include = FALSE, global = TRUE}
# Access indicator data
indicator_data <- read_csv(here('Indicator_Dashboard','Data', paste('Dashboard_Data_', RptYr, '.csv', sep = "")))

# And just use our variable of interest
indicator_data <- indicator_data |> filter(ID == var)
```

```{r, include = FALSE}
# Values that are included in the text

# Annual mean for report year
ann_mean_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Increase over time from linear model
if (!is.na(indicator_data$Value_lm[1])) {
  delta_size_lm <- indicator_data$Value_lm[length(indicator_data$Value_lm)] - 
    indicator_data$Value_lm[1]
  delta_size_pct = (indicator_data$Value_lm[length(indicator_data$Value_lm)] - 
                      indicator_data$Value_lm[1]) / indicator_data$Value_lm[1] * 100
}

# Increase in anomaly over time from linear model
if (!is.na(indicator_data$Anom_lm[1])) {
  delta_size_anom_lm <- indicator_data$Anom_lm[length(indicator_data$Anom_lm)] - 
    indicator_data$Anom_lm[1]
}

# Monthly min & max for report year
monthly_min_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))

# Monthly min & max for previous years
monthly_min_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))
```

## Median Phytoplankton Size
__About the Indicator__  
Phytoplankton are the base of the food web and their abundance influences the 
food available to all higher trophic levels from zooplankton through tuna and 
billfish. Sea surface temperature and ocean color data can be combined to estimate 
median phytoplankton size. Changes to median phytoplankton can propagate through 
the food web and influence fish size structure, weight-per-unit-effort, and the 
bigeye tuna recruitment index. Furthermore, median phytoplankton size can be 
combined with the recruitment index to forecast bigeye tuna catch rates up to 
four years in advance.

Some studies project that climate change will result in both fewer and smaller 
phytoplankton. This would reduce the food available to all members of the food web. 

Understanding variability and trends in phytoplankton abundance and size structure, 
how they are influenced by oceanographic conditions, and how they influence fish 
abundance and size structure are areas of active research.

__Status of the Indicator__  
The mean monthly phytoplankton cell size for `r RptYr` was `r signif(ann_mean_RptYr,3)` $\mu$m 
Equivalent Spherical Diameter (ESD), ranging from `r signif(monthly_min_RptYr,3)`–`r signif(monthly_max_RptYr,3)` $\mu$m ESD
during the year. These values remain within the historical records
(`r signif(monthly_min_PrevYrs,3)`–`r signif(monthly_max_PrevYrs,3)` $\mu$m ESD). 
Over the full time series, the median phytoplankton size experienced a weakly significant decline of `r abs(signif(delta_size_lm,2))` $\mu$m ESD (`r abs(signif(delta_size_pct,2))`%), accompanied by a `r abs(signif(delta_size_anom_lm,2))` $\mu$m ESD decrease in the monthly mean anomaly. Average estimated median phytoplankton size was below average across much of the fishing grounds, particularly north of 30°N.

__Indicator Data__  
Description: Median phytoplankton cell size can be estimated from satellite 
remotely sensed SST and ocean color (Barnes et al. 2011). A time series of monthly 
median phytoplankton cell size averaged over the Hawaiʻi longline region is presented, 
as well as a time series of anomalies. NOAA CoralTemp (see SST indicator) and 
ESA CCI data (see ocean color indicator) are used to calculate median phytoplankton cell size.

Timeframe: Monthly

Region: Hawaiʻi longline region: 15°–45°N, 180°–120°W

Measurement Platform: Satellite

Data available at: <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/md50_exp_2025.html, 
<https://oceanwatch.pifsc.noaa.gov/erddap/griddap/md50_exp-1998-2009-clim.html>, and 
<https://oceanwatch.pifsc.noaa.gov/erddap/griddap/md50_exp-2025-clim.html>.  

Sourced from: Barnes et al. (2011) and NOAA OceanWatch (2026c). Graphics produced in part using Stawitz and Li (2025). 

__References__  
Barnes et al. (2011): Barnes, C., Irigoien, X., De Oliveira, J.A.A., Maxwell, D., and S. Jennings, 2011. Predicting marine phytoplankton community size structure form empirical relationships with remotely sensed variables. Journal of Plankton Research, 33(1), pp. 13-24. doi: 10.1093/plankt/fbq088  

NOAA OceanWatch (2026c): <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/md50_exp_2025> Accessed: 6 April 2026. 

Stawitz C, Li B (2025). nmfspalette: A Color Palette for NOAA Fisheries. R package version 3.0.0.000. <https://nmfs-fish-tools.github.io/nmfspalette/>