---
title: NPGO_description"
author: "Phoebe.Woodworth-Jefcoats@noaa.gov"
date: "2023-12-06"
output: md_document
preserve_yaml: FALSE
---

```{r}
#| include: false
### Load libraries
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
library(markdown)
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2024

# Identify the variable we're focused on
var <- 'NPGO'
```

```{r, include = FALSE, global = TRUE}
# Access indicator data
indicator_data <- read_csv(here('Indicator_Dashboard','Data', paste('Dashboard_Data_', RptYr, '.csv', sep = "")))

# And just use our variable of interest
indicator_data <- indicator_data |> filter(ID == var)
```

```{r, include = FALSE}
# Values that are included in the text

# Annual mean for report year
ann_mean_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Increase over time from linear model
if (!is.na(indicator_data$Value_lm[1])) {
  delta_npgo_lm <- indicator_data$Value_lm[length(indicator_data$Value_lm)] - 
    indicator_data$Value_lm[1]
}

# Increase in anomaly over time from linear model
if (!is.na(indicator_data$Anom_lm[1])) {
  delta_npgo_anom_lm <- indicator_data$Anom_lm[length(indicator_data$Anom_lm)] - 
    indicator_data$Anom_lm[1]
}

# Monthly min & max for report year
monthly_min_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))

# Monthly min & max for previous years
monthly_min_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))
```

## North Pacific Gyre Oscillation
__About the Indicator__  
The North Pacific Gyre Oscillation (NPGO) is related to, but distinct from, the 
Pacific Decadal Oscillation.  It capture basin-scale processes, making it well 
suited to capture change across the Hawaiʻi-based longline fishing grounds

Understanding the effects of natural climate variability like the NPGO on the 
ocean, marine ecosystems, and fisheries is an active area of research.

__Status of the Indicator__  
The NPGO was negative in `r RptYr`. The index ranged from `r signif(monthly_min_RptYr, 3)` 
to `r signif(monthly_max_RptYr, 3)` over the course of the year.  This is within 
the range of values observed previously in the time series.

__Indicator Data__  
Description: The NPGO is driven by regional and basin-scale changes in wind
patterns. Changes to the wind field drive changes in ocean circulation that 
affect nutrients and, in turn, phytoplankton. Because phytoplankton are the 
foundational food source for marine ecosystems, the NPGO offers insight into 
potential ecosystem change. Description paraphrased from Di Lorenzo et al. (2008).

Timeframe: Monthly.

Region/Location: 180°–110°W; 25°N–62°N.

Measurement Platform: satellite, model.

Data available at: <https://www.o3d.org/npgo/data/NPGO.txt>.  

Sourced from Di Lorenzo et al. (2008).  Graphics produced in part using Stawitz and Li (2025).

__References__  
Di Lorenzo et al. (2008): Di Lorenzo, E., Schneider, N., Cobb, K. M., Franks, P. J. S., Chhak, K., Miller, A. J., McWilliams, J. C., Bograd, S. J., Arango, H., Curchitser, E., Powell, T. M., and Rivière, P., 2008: North Pacific Gyre Oscillation links ocean climate and ecosystem change. Geophysical Research Letters, 35(8), L08607, doi: 10.1029/2007GL032838.  

Stawitz C, Li B (2025). nmfspalette: A Color Palette for NOAA Fisheries. R package version 3.0.0.000. <https://nmfs-fish-tools.github.io/nmfspalette/>