---
title: "SST_description"
author: "Phoebe.Woodworth-Jefcoats@noaa.gov"
date: "2023-10-16"
output: md_document
preserve_yaml: FALSE
---

```{r}
#| include: false
### Load libraries
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
library(markdown)
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2023

# Identify the variable we're focused on
var <- 'SST'
```

```{r, include = FALSE, global = TRUE}
# Access indicator data
indicator_data <- read_csv(here('Indicator_Dashboard','Data', paste('Dashboard_Data_', RptYr, '.csv', sep = "")))

# And just use our variable of interest
indicator_data <- indicator_data |> filter(ID == var)
```

```{r, include = FALSE}
# Values that are included in the text

# Annual mean for report year
ann_mean_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Increase over time from linear model
if (!is.na(indicator_data$Value_lm[1])) {
  delta_SST_lm <- indicator_data$Value_lm[length(indicator_data$Value_lm)] - 
    indicator_data$Value_lm[1]
}

# Increase in anomaly over time from linear model
if (!is.na(indicator_data$Anom_lm[1])) {
  delta_SST_anom_lm <- indicator_data$Anom_lm[length(indicator_data$Anom_lm)] - 
    indicator_data$Anom_lm[1]
}

# Monthly min & max for report year
monthly_min_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))

# Monthly min & max for previous years
monthly_min_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))
```

## Sea Surface Temperature
__About the Indicator__  
Sea surface temperature (SST) is one of the most directly observable existing measures for
tracking increasing ocean temperatures. SST varies in response to natural climate cycles such as the El
Niño – Southern Oscillation (ENSO). Natural variability in SST impacts the marine ecosystem and pelagic fisheries. For example, higher SSTs can lead to the subtropical front being farther north and vice versa, which in turn affects the distance fishers may need to travel to reach longline fishing grounds. Changes in SST can also influence the number, location, strength, and seasonal timing of tropical cyclones.

SST is rising as a result of anthropogenic (human-caused) climate change.  Increasing carbon dioxide
concentrations cause the atmosphere to warm. Much of this heat is then absorbed by the ocean, raising sea
surface temperature. This means SST is a measure of how greenhouse gas emissions are already impacting the
ocean. Over time, rising sea surface temperatures will warm deeper ocean waters. Changes in ocean
temperature will affect tuna, and in turn, potentially their catchability. For example, fish may move to
deeper waters or their habitat could be compressed geographically or vertically. Rising ocean temperatures
are projected to lead to lower phytoplankton abundance, which is observed through ocean color and
estimated via chlorophyll-a (chl-a). Combined, rising ocean temperatures and lower phytoplankton abundance
may lead to smaller median phytoplankton sizes. Smaller phytoplankton may alter fish size structure,
weight-per-unit-effort, and the bigeye tuna recruitment index. 

Both short-term variability and long-term trends in SST impact the marine ecosystem. Understanding
the mechanisms through which organisms are impacted and the time scales of these impacts is an area of
active research.

__Status of the Indicator__  
Annual mean SST was `r signif(ann_mean_RptYr,3)` ºC in `r RptYr`. Over the period of record, SST across the longline fishing grounds has increased by `r signif(delta_SST_lm, 1)` ºC and the monthly SST anomaly increased by `r signif(delta_SST_anom_lm, 1)` ºC, both at a rate of roughly 0.03 ºC yr^-^^1^. Monthly SST values in `r RptYr` ranged from `r signif(monthly_min_RptYr, 3)`–`r signif(monthly_max_RptYr, 3)` ºC, within the range of temperatures experienced over the past several decades (`r signif(monthly_min_PrevYrs, 3)`–`r signif(monthly_max_PrevYrs, 3)` ºC). Overall, SST was above the long-term average across most of the Hawaiʻi longline region in `r RptYr`. The exception to this was a patch of slightly cooler waters in the southeastern corner of the fishing grounds where very little fishing takes place.

__Indicator Data__  
Description: Satellite remotely sensed monthly sea surface temperature (SST) is averaged across the
Hawaiʻi-based longline fishing grounds (15° – 45°N, 180° – 120°W). A time series of monthly mean SST
averaged over the Hawaiʻi longline region is presented. Additionally, spatial climatologies and anomalies
are shown. CoralTemp data are used to calculate this indicator. 

Timeframe: Monthly.

Region/Location: Hawaiʻi longline region: 15° – 45°N, 180° – 120°W.

Measurement Platform: Satellite.

Data available at: <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/CRW_sst_v3_1_monthly>, <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/CRW_sst_v3_1_1985-2009-clim>, and <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/CRW_sst_v3_1_2023-clim>.   

Sourced from: NOAA OceanWatch (2024a). Graphics produced in part using Stawitz (2023).

__References__  
NOAA OceanWatch (2024a): <https://oceanwatch.pifsc.noaa.gov/erddap/griddap/CRW_sst_v3_1_monthly> Accessed: 3 April 2024.

Stawitz C (2023): nmfspalette: A Color Palette for NOAA Fisheries. R package version 3.0.0.000. <https://nmfs-fish-tools.github.io/nmfspalette/>