---
title: "TatD_description"
author: "Phoebe.Woodworth-Jefcoats@noaa.gov"
date: "2023-12-06"
output: md_document
preserve_yaml: FALSE
---

```{r}
#| include: false
### Load libraries
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
library(markdown)
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2023

# Identify the variable we're focused on
var <- 'TatD'
```

```{r, include = FALSE, global = TRUE}
# Access indicator data
indicator_data <- read_csv(here('Indicator_Dashboard','Data', paste('Dashboard_Data_', RptYr, '.csv', sep = "")))

# And just use our variable of interest
indicator_data <- indicator_data |> filter(ID == var)
```

```{r, include = FALSE}
# Values that are included in the text

# Annual mean for report year
ann_mean_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = mean(Value, na.rm = TRUE))

# Increase over time from linear model
if (!is.na(indicator_data$Value_lm[1])) {
  delta_TatD_lm <- indicator_data$Value_lm[length(indicator_data$Value_lm)] - 
    indicator_data$Value_lm[1]
}

# Increase in anomaly over time from linear model
if (!is.na(indicator_data$Anom_lm[1])) {
  delta_TatD_anom_lm <- indicator_data$Anom_lm[length(indicator_data$Anom_lm)] - 
    indicator_data$Anom_lm[1]
}

# Monthly min & max for report year
monthly_min_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_RptYr <- indicator_data |> filter(Year == RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))

# Monthly min & max for previous years
monthly_min_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = min(Value, na.rm = TRUE))
monthly_max_PrevYrs <- indicator_data |> filter(Year < RptYr) |>
  summarise(Value = max(Value, na.rm = TRUE))
```

## Temperature at Depth
__About the Indicator__  
The temperature at 200–300 m reflects the temperature in the mid-range of depths targeted by the deep-set bigeye tuna fishery. Bigeye have preferred thermal habitat, generally staying within temperatures ranging from 8–14 °C while they are at depth (Howell et al. 2010). Changes in ocean temperature at depth will impact tuna, and in turn, potentially impact their catchability. For example, fish may move to deeper waters or their habitat could be compressed geographically or vertically. Understanding the drivers of sub-surface temperature trends and their ecosystem impacts is an area of active research.

Climate change will eventually affect temperatures at depth.  Over time, rising sea surface temperatures will warm deeper waters.  

It's possible that natural climate variability influences temperatures at 200–300 m below the surface where the bigeye fishery sets their hooks.  However, this relationship has yet to be established.

__Status of the Indicator__  
In `r RptYr`, 200–300 m temperatures ranged from `r signif(monthly_min_RptYr, 4)`–`r signif(monthly_max_RptYr, 4)` °C with an average value of `r signif(ann_mean_RptYr,4)` °C. These temperatures are within the range of temperatures experienced over the past several decades (`r signif(monthly_min_PrevYrs, 4)`–`r signif(monthly_max_PrevYrs, 4)` °C) and are within the bounds of bigeye tuna’s preferred deep daytime thermal habitat (8–14 °C). Over the period of record (1980–`r RptYr`), 200–300 m temperatures have declined by `r abs(signif(delta_TatD_lm, 1))` °C. The spatial pattern of temperature anomalies was mixed with temperatures at depth around the main Hawaiian Islands roughly 0.5–1.5 °C below average, and temperatures north of about 30°N 0–0.5 °C above average.

__Indicator Data__  
Description: Ocean temperature at 200–300 m depth is averaged across the Hawaiʻi-based longline fishing grounds (15° – 45°N, 180° – 120°W). Global Ocean Data Assimilation System (GODAS) data are used. GODAS incorporates global ocean data from moorings, expendable bathythermographs (XBTs), and Argo floats. 

Timeframe: Annual, monthly.

Region/Location: Hawaii longline region: 15° – 45°N, 180° – 120°W.

Measurement Platform: In-situ sensors, model.

Sourced from: NOAA (2024d) and APDRC (2024).  Graphics produced in part using Stawitz (2023).

__References__  
APDRC (2024): <http://apdrc.soest.hawaii.edu:80/dods/public_data/Reanalysis_Data/GODAS/monthly/potmp> Accessed 4 April 2024.   

Howell et al. (2010): Howell, E.A., Hawn, D.R., and J.J. Polovina, 2010. Spatiotemporal variability in bigeye tuna (Thunnus obesus) dive behavior in the central North Pacific Ocean. Progress in Oceanography, 86, pp. 81-93. doi: 10.1016/j.pocean.2010.04.013.  

NOAA (2024d): NOAA Office of Oceanic and Atmospheric Research’s Earth System Research Laboratories’ Physical Sciences Laboratory, <https://psl.noaa.gov/data/gridded/data.godas.html> Accessed 4 April 2024.  

Stawitz C (2023): nmfspalette: A Color Palette for NOAA Fisheries. R package version 3.0.0.000. <https://nmfs-fish-tools.github.io/nmfspalette/>